---
output: github_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

# ggpaintr

`ggpaintr` is an R package for the maintained `ggpaintr` workflow. It turns a
single ggplot-like formula string into a small Shiny app with generated
controls, rendered plots, generated code, upload support, copy customization,
custom Shiny integration hooks, custom placeholder registries, and export
helpers.

## Installation

```r
# install.packages("pak")
pak::pkg_install("willju-wangqian/ggpaintr")
```

## Core concepts

`ggpaintr` works by treating a ggplot-like formula string as a template with
placeholders that become Shiny inputs.

- `var` selects a data column in the generated UI
- `text` collects free text
- `num` collects numeric input
- `expr` collects raw R code for places like faceting or labels
- `upload` lets the app use uploaded data
- `ggpaintr_normalize_column_names()` cleans local column names before `var`
  selection when the source data is not already syntactic
- you can register your own placeholder types per app with
  `ggpaintr_placeholder()`

`upload` currently supports `.csv` and `.rds`.

```r
"ggplot(data = iris, aes(x = var, y = var)) +
  geom_point(size = num, color = text) +
  labs(title = text) +
  facet_wrap(expr)"
```

## Quick start

The low-level workflow is: parse a formula, build a runtime from user inputs,
then inspect the generated code and plot.

```{r quick-start}
library(ggpaintr)
library(ggplot2)

obj <- paintr_formula(
  "ggplot(data = mtcars, aes(x = var, y = var)) +
    geom_point(size = num) +
    labs(title = text)"
)

runtime <- paintr_build_runtime(
  obj,
  list(
    "ggplot+3+2" = "mpg",
    "ggplot+3+3" = "disp",
    "geom_point+2" = 2,
    "labs+2" = "Mtcars scatter",
    "geom_point+checkbox" = TRUE,
    "labs+checkbox" = TRUE
  )
)

runtime$code_text
inherits(runtime$plot, "ggplot")
```

## Feature tour

### Build a plot from a formula

You can use the low-level runtime helpers directly when you want to inspect or
test the generated plot code outside Shiny.

```{r formula-runtime}
obj <- paintr_formula(
  "ggplot(data = iris, aes(x = var, y = var)) +
    geom_point(color = text) +
    labs(title = text)"
)

runtime <- paintr_build_runtime(
  obj,
  list(
    "ggplot+3+2" = "Sepal.Length",
    "ggplot+3+3" = "Petal.Length",
    "geom_point+2" = "steelblue",
    "labs+2" = "Iris scatter",
    "geom_point+checkbox" = TRUE,
    "labs+checkbox" = TRUE
  )
)

runtime$code_text
```

### Launch an interactive app

Use `ggpaintr_app()` for the default wrapper when you want the package to build
the Shiny app shell for you.

```r
library(ggpaintr)

ggpaintr_app("
ggplot(data = iris, aes(x = var, y = var)) +
  geom_point(aes(color = var), size = num) +
  labs(title = text) +
  facet_wrap(expr)
")
```

### Work with uploaded data

Use `upload` when the app should let users supply a dataset at runtime.

```r
ggpaintr_app("
ggplot(data = upload, aes(x = var, y = var)) +
  geom_point(size = num) +
  labs(title = text)
")
```

The upload control currently supports `.csv` and `.rds`.

Uploads are normalized through the same column-name helper automatically after
read-in. `.rds` uploads must already be tabular or be coercible with
`as.data.frame()`.

### Prepare local data with non-syntactic column names

If your local data uses spaces or punctuation in column names, normalize it
once before you build the app.

```r
messy_sales <- data.frame(
  left = 1:4,
  right = c(2, 4, 6, 8),
  check.names = FALSE
)
names(messy_sales) <- c("first column", "second-column")

sales <- ggpaintr_normalize_column_names(messy_sales)
names(sales)

ggpaintr_app("
ggplot(data = sales, aes(x = var, y = var)) +
  geom_point()
")
```

### Put transforms around `var` in the formula

`var` is a column picker. When you want derived mappings, wrap the placeholder
in the formula itself and let the selected column slot into that expression.

```r
ggpaintr_app("
ggplot(data = mtcars, aes(x = var + 1, y = log(var))) +
  geom_point()
")
```

### Customize control text with `copy_rules`

Use `copy_rules` to change user-facing labels without changing the runtime
behavior of the generated app.

```r
custom_copy <- list(
  shell = list(
    draw_button = list(label = "Render plot"),
    export_button = list(label = "Save app")
  ),
  params = list(
    x = list(var = list(label = "X axis variable")),
    y = list(var = list(label = "Y axis variable")),
    title = list(text = list(label = "Plot title"))
  )
)

ggpaintr_app(
  "ggplot(data = iris, aes(x = var, y = var)) +
     geom_point() +
     labs(title = text)",
  copy_rules = custom_copy
)
```

### Register a custom placeholder type

Use `ggpaintr_placeholder()` and `ggpaintr_effective_placeholders()` when you
want to add a new control type without editing package internals.

```{r custom-placeholder}
sales <- data.frame(
  day = as.Date("2024-01-01") + 0:4,
  value = c(10, 13, 12, 16, 18)
)

date_placeholder <- ggpaintr_placeholder(
  keyword = "date",
  build_ui = function(id, copy, meta, context) {
    shiny::dateInput(id, copy$label)
  },
  resolve_expr = function(value, meta, context) {
    if (is.null(value) || identical(as.character(value), "")) {
      return(ggpaintr_missing_expr())
    }

    rlang::expr(as.Date(!!as.character(value)))
  },
  copy_defaults = list(label = "Choose a date for {param}")
)

placeholders <- ggpaintr_effective_placeholders(
  list(date = date_placeholder)
)

obj <- paintr_formula(
  "ggplot(data = sales, aes(x = day, y = value)) +
    geom_line() +
    geom_vline(xintercept = date)",
  placeholders = placeholders
)

names(obj$placeholders)
```

```r
ggpaintr_app(
  "ggplot(data = sales, aes(x = day, y = value)) +
    geom_line() +
    geom_vline(xintercept = date)",
  placeholders = placeholders
)
```

### Export a standalone app

Use `generate_shiny()` when you want an explicit, editable Shiny app file that
you can continue customizing outside the package wrapper.

```{r export, eval=FALSE}
obj <- paintr_formula(
  "ggplot(data = upload, aes(x = var, y = var)) +
    geom_point() +
    labs(title = text)"
)

app_file <- tempfile(fileext = ".R")
generate_shiny(obj, list(), app_file)

# The generated file keeps an explicit ui, an explicit server, and a visible
# copy_rules hook so you can keep editing it as a normal Shiny app. If you pass
# custom placeholders, the exported file also rebuilds the placeholder registry.
```

### Embed `ggpaintr` in your own Shiny app

The phase-1 extensibility layer lets you keep your own layout while reusing the
`ggpaintr` state and bind helpers.

```{r integrate, eval=FALSE}
library(shiny)

ids <- ggpaintr_ids(
  control_panel = "builder_controls",
  draw_button = "render_plot",
  export_button = "export_app",
  plot_output = "main_plot",
  error_output = "main_error",
  code_output = "main_code"
)

ui <- fluidPage(
  sidebarLayout(
    sidebarPanel(ggpaintr_controls_ui(ids = ids)),
    mainPanel(ggpaintr_outputs_ui(ids = ids))
  )
)

server <- function(input, output, session) {
  paintr_state <- ggpaintr_server_state(
    "ggplot(data = iris, aes(x = var, y = var)) +
      geom_point() +
      labs(title = text)",
    ids = ids
  )

  ggpaintr_bind_control_panel(input, output, paintr_state, ids = ids)
  ggpaintr_bind_draw(input, paintr_state, ids = ids)
  ggpaintr_bind_export(output, paintr_state, ids = ids)
  ggpaintr_bind_plot(output, paintr_state, ids = ids)
  ggpaintr_bind_error(output, paintr_state, ids = ids)
  ggpaintr_bind_code(output, paintr_state, ids = ids)
}

shinyApp(ui, server)
```

### Customize the plot in your own `renderPlot()`

If you want to keep the `ggpaintr` runtime but take over plot rendering, use
`ggpaintr_plot_value()` inside your own `renderPlot()`.

```{r customize-plot, eval=FALSE}
server <- function(input, output, session) {
  paintr_state <- ggpaintr_server_state(
    "ggplot(data = iris, aes(x = var, y = var)) + geom_point()"
  )

  ggpaintr_bind_control_panel(input, output, paintr_state)
  ggpaintr_bind_draw(input, paintr_state)
  ggpaintr_bind_export(output, paintr_state)
  ggpaintr_bind_error(output, paintr_state)
  ggpaintr_bind_code(output, paintr_state)

  output$outputPlot <- renderPlot({
    plot_obj <- ggpaintr_plot_value(paintr_state$runtime())

    if (is.null(plot_obj)) {
      plot.new()
      return(invisible(NULL))
    }

    plot_obj + ggplot2::theme_minimal()
  })
}
```

## Behavior and boundaries

- Structural formula errors fail early during parsing.
- `var` with no data source fails while preparing the UI.
- `var` is a column picker. Formula-level transforms such as `var + 1` or
  `log(var)` are supported inside the formula text, not as direct input values.
- For local data with non-syntactic names, call
  `ggpaintr_normalize_column_names()` first; uploads apply the same
  normalization automatically.
- Missing local data objects are deferred to draw-time inline errors.
- Advanced integrations can customize the plot through `ggpaintr_plot_value()`.
- Custom placeholders are registered per app through
  `ggpaintr_effective_placeholders()`.
- `upload` currently supports `.csv` and `.rds`.

## Where to go next

See `vignette("ggpaintr-workflow")` for the main workflow and
`vignette("ggpaintr-extensibility")` for custom Shiny integration recipes.
See `vignette("ggpaintr-placeholder-registry")` for the full placeholder
registry API, hook contract, and export guidance.
