---
output: github_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)

if (!requireNamespace("ggpaintr", quietly = TRUE)) {
  devtools::install(quiet = TRUE)
}
library(ggpaintr)
```

# ggpaintr

`ggpaintr` turns one ggplot-like formula string into a small Shiny app with
generated controls, rendered plots, generated code, upload support, copy
customization, custom Shiny integration hooks, and custom placeholder
registries.

## Installation

```r
# install.packages("pak")
pak::pkg_install("willju-wangqian/ggpaintr")
```

## Core concepts

`ggpaintr` treats a ggplot-like formula string as a template with placeholders
that become Shiny inputs.

- `var` selects a data column in the generated UI
- `text` collects free text
- `num` collects numeric input
- `expr` collects raw R code for places like faceting or labels
- `upload` lets the app use uploaded data
- `ptr_normalize_column_names()` cleans local column names before `var`
  selection when the source data is not already syntactic
- you can register your own placeholder types per app with
  `ptr_define_placeholder()`

`upload` currently supports `.csv` and `.rds`.

```r
"ggplot(data = iris, aes(x = var, y = var)) +
  geom_point(size = num, color = text) +
  labs(title = text) +
  facet_wrap(expr)"
```

## Why this design

`ggpaintr` keeps the author-facing input as one formula string on purpose.

- formula strings stay close to the way many R users already sketch ggplot code
- the same parsed object drives generated controls, runtime completion,
  and generated code
- one placeholder registry powers parse, UI, runtime, and copy rules for both
  built-in and custom placeholders
- the package gives you both a default wrapper and a supported Shiny embedding
  layer, so you can start simple and grow into a more custom app

## Quick start

Most users should start with `ptr_app()`.

```r
library(ggpaintr)

ptr_app("
ggplot(data = iris, aes(x = var, y = var)) +
  geom_point(aes(color = var), size = num) +
  labs(title = text) +
  facet_wrap(expr)
")
```

## Supported public API

The maintained public path is intentionally narrow.

- Start with `ptr_app()` for the default app wrapper, or `ptr_app_bslib()`
  for a `bslib`-themed variant that doubles as a worked example of building
  custom shells from the public API.
- Use `ptr_server()`, `ptr_server_state()`, `ptr_build_ids()`, and the
  `ptr_register_*()` helpers when you need to embed `ggpaintr` in a larger
  Shiny app.
- Use `ptr_define_placeholder()` and `ptr_merge_placeholders()` for
  custom placeholder types.
- Use `ptr_runtime_input_spec()`, `ptr_parse_formula()`,
  `ptr_exec()`, and `ptr_assemble_plot()` for advanced runtime or
  testing workflows.

Other helpers in `R/` are internal implementation support rather than part of
the maintained community-facing API.

## Feature tour

### Work with uploaded data

Use `upload` when the app should let users supply a dataset at runtime.

```r
ptr_app("
ggplot(data = upload, aes(x = var, y = var)) +
  geom_point(size = num) +
  labs(title = text)
")
```

The upload control currently supports `.csv` and `.rds`.

Uploads are normalized through the same column-name helper automatically after
read-in. `.rds` uploads must already be tabular or be coercible with
`as.data.frame()`.

### Prepare local data with non-syntactic column names

If your local data uses spaces or punctuation in column names, normalize it
once before you build the app.

```r
messy_sales <- data.frame(
  left = 1:4,
  right = c(2, 4, 6, 8),
  check.names = FALSE
)
names(messy_sales) <- c("first column", "second-column")

sales <- ptr_normalize_column_names(messy_sales)
names(sales)

ptr_app("
ggplot(data = sales, aes(x = var, y = var)) +
  geom_point()
")
```

### Put transforms around `var` in the formula

`var` is a column picker. When you want derived mappings, wrap the placeholder
in the formula itself and let the selected column slot into that expression.

```r
ptr_app("
ggplot(data = mtcars, aes(x = var + 1, y = log(var))) +
  geom_point()
")
```

### Customize control text with `ui_text`

Use `ui_text` to change user-facing labels without changing the runtime
behavior of the generated app.

```r
custom_copy <- list(
  shell = list(
    draw_button = list(label = "Render plot")
  ),
  params = list(
    x = list(var = list(label = "X axis variable")),
    y = list(var = list(label = "Y axis variable")),
    title = list(text = list(label = "Plot title"))
  )
)

ptr_app(
  "ggplot(data = iris, aes(x = var, y = var)) +
     geom_point() +
     labs(title = text)",
  ui_text = custom_copy
)
```

### Embed `ggpaintr` in your own Shiny app

The supported integration layer lets you keep your own layout while reusing the
`ggpaintr` state and bind helpers.

```{r integrate, eval=FALSE}
library(shiny)

ids <- ptr_build_ids(
  control_panel = "builder_controls",
  draw_button = "render_plot",
  plot_output = "main_plot",
  error_output = "main_error",
  code_output = "main_code"
)

ui <- fluidPage(
  sidebarLayout(
    sidebarPanel(ptr_input_ui(ids = ids)),
    mainPanel(ptr_output_ui(ids = ids))
  )
)

server <- function(input, output, session) {
  ptr_state <- ptr_server_state(
    "ggplot(data = iris, aes(x = var, y = var)) +
      geom_point() +
      labs(title = text)",
    ids = ids
  )

  ptr_setup_controls(input, output, ptr_state, ids = ids)
  ptr_register_draw(input, ptr_state, ids = ids)
  ptr_register_plot(output, ptr_state, ids = ids)
  ptr_register_error(output, ptr_state, ids = ids)
  ptr_register_code(output, ptr_state, ids = ids)
}

shinyApp(ui, server)
```

### Customize the plot in your own `renderPlot()`

If you want to keep the `ggpaintr` runtime but take over plot rendering, use
`ptr_extract_plot()` inside your own `renderPlot()`. Wrap any extra ggplot
components in `ptr_gg_extra()` so the default code binder keeps the
`outputCode` pane in sync with what is actually drawn.

```{r customize-plot, eval=FALSE}
server <- function(input, output, session) {
  ptr_state <- ptr_server_state(
    "ggplot(data = iris, aes(x = var, y = var)) + geom_point()"
  )

  ptr_setup_controls(input, output, ptr_state)
  ptr_register_draw(input, ptr_state)
  ptr_register_error(output, ptr_state)
  ptr_register_code(output, ptr_state)

  output$outputPlot <- renderPlot({
    plot_obj <- ptr_extract_plot(ptr_state$runtime())

    if (is.null(plot_obj)) {
      plot.new()
      return(invisible(NULL))
    }

    plot_obj + ptr_gg_extra(ptr_state, ggplot2::theme_minimal(base_size = 16))
  })
}
```

### Register a custom placeholder type

Use `ptr_define_placeholder()` and `ptr_merge_placeholders()` when you
want to add a new control type without editing package internals.

```{r custom-placeholder}
sales <- data.frame(
  day = as.Date("2024-01-01") + 0:4,
  value = c(10, 13, 12, 16, 18)
)

date_placeholder <- ptr_define_placeholder(
  keyword = "date",
  build_ui = function(id, copy, meta, context) {
    shiny::dateInput(id, copy$label)
  },
  resolve_expr = function(value, meta, context) {
    if (is.null(value) || identical(as.character(value), "")) {
      return(ptr_missing_expr())
    }

    rlang::expr(as.Date(!!as.character(value)))
  },
  copy_defaults = list(label = "Choose a date for {param}")
)

placeholders <- ptr_merge_placeholders(
  list(date = date_placeholder)
)

obj <- ptr_parse_formula(
  "ggplot(data = sales, aes(x = day, y = value)) +
    geom_line() +
    geom_vline(xintercept = date)",
  placeholders = placeholders
)

names(obj$placeholders)
```

```r
ptr_app(
  "ggplot(data = sales, aes(x = day, y = value)) +
    geom_line() +
    geom_vline(xintercept = date)",
  placeholders = placeholders
)
```

### Advanced developer workflow

Use the low-level runtime helpers directly when you want to inspect generated
code, write tests, or build developer tooling around parsed formulas.

`ptr_runtime_input_spec()` is the supported way to discover the runtime
input ids needed by `ptr_exec()`.

```{r advanced-runtime}
obj <- ptr_parse_formula(
  "ggplot(data = mtcars, aes(x = var, y = var)) +
    geom_point(size = num) +
    labs(title = text)"
)

spec <- ptr_runtime_input_spec(obj)
spec

inputs <- setNames(vector("list", nrow(spec)), spec$input_id)
checkbox_rows <- spec$role == "layer_checkbox"
inputs[checkbox_rows] <- rep(list(TRUE), sum(checkbox_rows))
inputs[["ggplot+3+2"]] <- "mpg"
inputs[["ggplot+3+3"]] <- "disp"
inputs[["geom_point+2"]] <- 2
inputs[["labs+2"]] <- "Mtcars scatter"

runtime <- ptr_exec(obj, inputs)

runtime$code_text
inherits(runtime$plot, "ggplot")
```

For upload-backed formulas, the spec also includes the derived dataset-name
input that accompanies each upload control.

```{r advanced-upload-spec}
upload_obj <- ptr_parse_formula(
  "ggplot(data = upload, aes(x = var, y = var)) + geom_point()"
)

ptr_runtime_input_spec(upload_obj)
```

## Current stability guarantees

- built-in placeholders are `var`, `text`, `num`, `expr`, and `upload`
- custom placeholders share the same parse, UI, runtime, and copy-rule path
  as built-ins
- `upload` currently supports `.csv` and `.rds`
- the supported app-facing surface is the current `ptr_*` wrapper and
  Shiny-integration layer
- only the five top-level ids exposed by `ptr_build_ids()` are configurable
- runtime failures are labeled as `Input error` or `Plot error` and stay on
  the shared inline error path

## Not guaranteed / implementation details

- raw placeholder ids such as `"ggplot+3+2"` are not a stable hand-authored API;
  discover them with `ptr_runtime_input_spec()`
- deeper traversal details such as `index_path` encoding and internal companion
  id conventions remain package internals
- unsupported upload formats are outside the current boundary
- the formula-string model is the current author-facing interface; future
  hardening should compile it into a richer internal runtime contract rather
  than replace that authoring model outright

## Current behavior boundary

- Structural formula errors fail early during parsing.
- `var` with no data source fails while preparing the UI.
- `var` is a column picker. Formula-level transforms such as `var + 1` or
  `log(var)` are supported inside the formula text, not as direct input values.
- For local data with non-syntactic names, call
  `ptr_normalize_column_names()` first; uploads apply the same
  normalization automatically.
- Missing local data objects are deferred to draw-time inline errors.
- Advanced integrations can customize the plot through `ptr_extract_plot()`.

## Where to go next

See `vignette("ggpaintr-workflow")` for the main workflow and
`vignette("ggpaintr-extensibility")` for supported Shiny integration recipes.
See `vignette("ggpaintr-placeholder-registry")` for the placeholder registry
API and hook contract.
