---
output: github_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

# ggpaintr

`ggpaintr` is an R package for the maintained `ggpaintr` workflow. It turns a
single ggplot-like formula string into a small Shiny app with generated
controls, rendered plots, generated code, upload support, and inline runtime
feedback.

## Installation

```r
# install.packages("pak")
pak::pkg_install("willju-wangqian/ggpaintr")
```

## The `ggpaintr` model

Supported placeholders inside a formula string are:

- `var`
- `text`
- `num`
- `expr`
- `upload`

`upload` currently supports `.csv` and `.rds`.

## Quick start

```{r quick-start}
library(ggpaintr)
library(ggplot2)

obj <- paintr_formula(
  "ggplot(data = mtcars, aes(x = var, y = var)) +
    geom_point(size = num) +
    labs(title = text)"
)

runtime <- paintr_build_runtime(
  obj,
  list(
    "ggplot+3+2" = "mpg",
    "ggplot+3+3" = "disp",
    "geom_point+2" = 2,
    "labs+2" = "Mtcars scatter",
    "geom_point+checkbox" = TRUE,
    "labs+checkbox" = TRUE
  )
)

runtime$code_text
inherits(runtime$plot, "ggplot")
```

## Launch an app

```r
library(ggpaintr)

ggpaintr_app("
ggplot(data = iris, aes(x = var, y = var)) +
  geom_point(aes(color = var), size = num) +
  labs(title = text) +
  facet_wrap(expr)
")
```

## Export a standalone app

```{r export, eval=FALSE}
obj <- paintr_formula(
  "ggplot(data = upload, aes(x = var, y = var)) + geom_point()"
)

generate_shiny(obj, list(), tempfile(fileext = ".R"))
```

## Runtime behavior

- Structural formula errors fail at parse time.
- `var` with no data source fails while preparing UI.
- Missing local data objects are deferred to draw-time plot errors.
- Render-time ggplot failures are shown through the same inline error channel.

See `vignette("ggpaintr-workflow")` for a fuller walkthrough.
