---
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "80%",
  echo=FALSE,
  warning = FALSE,
  message = FALSE
)
```

## Installation

You can install the development version from Github with:

```{r, eval=FALSE, echo=TRUE}
# install.packages("devtools")
devtools::install_github("willju-wangqian/ggpaintr")
```

## Run the shiny app

```{r, eval=FALSE, echo=TRUE}
library(ggpaintr)

run_ggpaintr_app()
```



## Development

+   location for external data: `inst/extdata`
+   location for example shiny app: `inst/shiny`
+   location for files that need to be uploaded but are not part of the package: `preconsideration/<folder_name>`
+   coding style consistency: please try to use `underscores_in_variable_names` and (maybe) `camelCase` for web-related names...

## Generate plot icons

+   the code that generates the plot icons: `preconsideration/generate_plot_icon`
+   we can save the generated icons into `preconsideration/saved_plot_icon` or copy them to `inst/shiny/ggpaintr_app/www/img_button`
+   different functions for `actionButton`
    +   `shinyWidgets::actionBttn`, [link](https://bttn.surge.sh/)
    +   `actionButton`: this is the default and we can specify the `style` attribute
    +   `bsButton`: this includes both the `actionButton` and `checkboxInput`, but we don't have much room for `style` 




## Current Version

#### version number: 0.0.9.7



#### version number: 0.0.9.6

- alpha as numeric
- n < 3 colors

- added four chart types, next: update functions in the package

- change expression by clicking the plot icon
- test if ui functions can be added
- how to add customized handler functions?

#### version number: 0.0.9.5

- Updates:
  - introduced `paintr_obj` and `paintr()`, which use a grammar similar to `ggplot2` to specify and control all elements or pieces built in the shiny app
  - use one function to build reactivity of `color` or `fill` so that they can work with `paintr_plot_code`

- TODO:
  - documentations and dependencies
  - write-ups
  - download plot
  - test and finalize code then push to main

#### version number: 0.0.9.4

- heuristics:
  - parameters of handlers
  - parameters of ui modules
  - returns of handlers

+ TODO: 
  - documentations 
  - [X] icon position or size 
  - writeup
    - the concept of grammar of graphics
    - why this package can be useful
    - how to use the package
  - [X] how to wrap up the part of scale_color
  - template?
  - download plot
  - [X] aes then comma

#### version number: 0.0.9.3

+   updates:
    +   the app is now able to show code that generates the plot
    +   dynamic color picker
+   TODO:
    +   update ui
    +   implement other plot types
    +   extend the current functionality: added more handlers (for example one for `labs()`)
    +   documentations

#### version number: 0.0.9.2

+   updates:
    +   `*Handler`s are introduced for plot generation
    +   added bar chart

#### version number: 0.0.9.1

+   updates:
    +   new ui
    +   example for box plot finished
+   TODO:
    +   update the geomGenerators
        +   each line of Grammar of Graphics has its onw handler
        +   a function that collects all handlers
        +   find a way to connect handlers to its ui components
    +   allow color pickers to respond to the selected variable
    +   update ui
    +   implement other plot types
    +   display code
    



