---
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "80%",
  echo=FALSE,
  warning = FALSE,
  message = FALSE
)
```

## Installation

You can install the development version from Github with:

```{r, eval=FALSE, echo=TRUE}
# install.packages("devtools")
devtools::install_github("willju-wangqian/ggpaintr")
```

## Run the shiny app

```{r, eval=FALSE, echo=TRUE}
library(ggpaintr)

run_ggpaintr_app()
```



## Development

+   location for external data: `inst/extdata`
+   location for example shiny app: `inst/shiny`
+   location for files that need to be uploaded but are not part of the package: `preconsideration/<folder_name>`
+   coding style consistency: please try to use `underscores_in_variable_names` and (maybe) `camelCase` for web-related names...

## Generate plot icons

+   the code that generates the plot icons: `preconsideration/generate_plot_icon`
+   we can save the generated icons into `preconsideration/saved_plot_icon` or copy them to `inst/shiny/ggpaintr_app/www/img_button`
+   different functions for `actionButton`
    +   `shinyWidgets::actionBttn`, [link](https://bttn.surge.sh/)
    +   `actionButton`: this is the default and we can specify the `style` attribute
    +   `bsButton`: this includes both the `actionButton` and `checkboxInput`, but we don't have much room for `style` 




## Current Version

#### version number: 0.0.9.2

+   updates:
    +   `*Handler`s are introduced for plot generation
    +   added bar chart

#### version number: 0.0.9.1

+   updates:
    +   new ui
    +   example for box plot finished
+   TODO:
    +   update the geomGenerators
        +   each line of Grammar of Graphics has its onw handler
        +   a function that collects all handlers
        +   find a way to connect handlers to its ui components
    +   allow color pickers to respond to the selected variable
    +   update ui
    +   implement other plot types
    +   display code
    



