---
output: github_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

# ggpaintr

`ggpaintr` turns one ggplot-like formula string into a small Shiny app with
generated controls, rendered plots, generated code, upload support, copy
customization, custom Shiny integration hooks, custom placeholder registries,
and export helpers.

## Installation

```r
# install.packages("pak")
pak::pkg_install("willju-wangqian/ggpaintr")
```

## Core concepts

`ggpaintr` treats a ggplot-like formula string as a template with placeholders
that become Shiny inputs.

- `var` selects a data column in the generated UI
- `text` collects free text
- `num` collects numeric input
- `expr` collects raw R code for places like faceting or labels
- `upload` lets the app use uploaded data
- `ggpaintr_normalize_column_names()` cleans local column names before `var`
  selection when the source data is not already syntactic
- you can register your own placeholder types per app with
  `ggpaintr_placeholder()`

`upload` currently supports `.csv` and `.rds`.

```r
"ggplot(data = iris, aes(x = var, y = var)) +
  geom_point(size = num, color = text) +
  labs(title = text) +
  facet_wrap(expr)"
```

## Why this design

`ggpaintr` keeps the author-facing input as one formula string on purpose.

- formula strings stay close to the way many R users already sketch ggplot code
- the same parsed object can drive generated controls, runtime completion,
  generated code, and standalone export
- one placeholder registry powers parse, UI, runtime, copy rules, and export
  for both built-in and custom placeholders
- the package gives you both a default wrapper and a supported Shiny embedding
  layer, so you can start simple and grow into a more custom app
- exported apps stay explicit `ui <- ...` and `server <- function(...)` files so
  they remain readable and editable outside the package wrapper

## Quick start

Most users should start with `ggpaintr_app()`.

```r
library(ggpaintr)

ggpaintr_app("
ggplot(data = iris, aes(x = var, y = var)) +
  geom_point(aes(color = var), size = num) +
  labs(title = text) +
  facet_wrap(expr)
")
```

## Supported public API

The maintained public path is intentionally narrow.

- Start with `ggpaintr_app()` for the default app wrapper.
- Use `ggpaintr_server()`, `ggpaintr_server_state()`, `ggpaintr_ids()`, and the
  `ggpaintr_bind_*()` helpers when you need to embed `ggpaintr` in a larger
  Shiny app.
- Use `ggpaintr_placeholder()` and `ggpaintr_effective_placeholders()` for
  custom placeholder types.
- Use `ggpaintr_runtime_input_spec()`, `ggpaintr_formula()`,
  `ggpaintr_build_runtime()`, and `ggpaintr_get_plot()` for advanced runtime or
  testing workflows.
- Use `generate_shiny()` to export an explicit standalone app script.

Other helpers in `R/` are internal implementation support rather than part of
the maintained community-facing API.

## Feature tour

### Work with uploaded data

Use `upload` when the app should let users supply a dataset at runtime.

```r
ggpaintr_app("
ggplot(data = upload, aes(x = var, y = var)) +
  geom_point(size = num) +
  labs(title = text)
")
```

The upload control currently supports `.csv` and `.rds`.

Uploads are normalized through the same column-name helper automatically after
read-in. `.rds` uploads must already be tabular or be coercible with
`as.data.frame()`.

### Prepare local data with non-syntactic column names

If your local data uses spaces or punctuation in column names, normalize it
once before you build the app.

```r
messy_sales <- data.frame(
  left = 1:4,
  right = c(2, 4, 6, 8),
  check.names = FALSE
)
names(messy_sales) <- c("first column", "second-column")

sales <- ggpaintr_normalize_column_names(messy_sales)
names(sales)

ggpaintr_app("
ggplot(data = sales, aes(x = var, y = var)) +
  geom_point()
")
```

### Put transforms around `var` in the formula

`var` is a column picker. When you want derived mappings, wrap the placeholder
in the formula itself and let the selected column slot into that expression.

```r
ggpaintr_app("
ggplot(data = mtcars, aes(x = var + 1, y = log(var))) +
  geom_point()
")
```

### Customize control text with `copy_rules`

Use `copy_rules` to change user-facing labels without changing the runtime
behavior of the generated app.

```r
custom_copy <- list(
  shell = list(
    draw_button = list(label = "Render plot"),
    export_button = list(label = "Save app")
  ),
  params = list(
    x = list(var = list(label = "X axis variable")),
    y = list(var = list(label = "Y axis variable")),
    title = list(text = list(label = "Plot title"))
  )
)

ggpaintr_app(
  "ggplot(data = iris, aes(x = var, y = var)) +
     geom_point() +
     labs(title = text)",
  copy_rules = custom_copy
)
```

### Export a standalone app

Use `generate_shiny()` when you want an explicit, editable Shiny app file that
you can continue customizing outside the package wrapper.

```{r export, eval=FALSE}
obj <- ggpaintr_formula(
  "ggplot(data = upload, aes(x = var, y = var)) +
    geom_point() +
    labs(title = text)"
)

app_file <- tempfile(fileext = ".R")
generate_shiny(obj, app_file)

# The generated file keeps an explicit ui, an explicit server, and visible
# copy_rules/placeholders hooks so you can keep editing it as a normal Shiny app.
```

### Embed `ggpaintr` in your own Shiny app

The supported integration layer lets you keep your own layout while reusing the
`ggpaintr` state and bind helpers.

```{r integrate, eval=FALSE}
library(shiny)

ids <- ggpaintr_ids(
  control_panel = "builder_controls",
  draw_button = "render_plot",
  export_button = "export_app",
  plot_output = "main_plot",
  error_output = "main_error",
  code_output = "main_code"
)

ui <- fluidPage(
  sidebarLayout(
    sidebarPanel(ggpaintr_controls_ui(ids = ids)),
    mainPanel(ggpaintr_outputs_ui(ids = ids))
  )
)

server <- function(input, output, session) {
  ggpaintr_state <- ggpaintr_server_state(
    "ggplot(data = iris, aes(x = var, y = var)) +
      geom_point() +
      labs(title = text)",
    ids = ids
  )

  ggpaintr_bind_control_panel(input, output, ggpaintr_state, ids = ids)
  ggpaintr_bind_draw(input, ggpaintr_state, ids = ids)
  ggpaintr_bind_export(output, ggpaintr_state, ids = ids)
  ggpaintr_bind_plot(output, ggpaintr_state, ids = ids)
  ggpaintr_bind_error(output, ggpaintr_state, ids = ids)
  ggpaintr_bind_code(output, ggpaintr_state, ids = ids)
}

shinyApp(ui, server)
```

### Customize the plot in your own `renderPlot()`

If you want to keep the `ggpaintr` runtime but take over plot rendering, use
`ggpaintr_plot_value()` inside your own `renderPlot()`.

```{r customize-plot, eval=FALSE}
server <- function(input, output, session) {
  ggpaintr_state <- ggpaintr_server_state(
    "ggplot(data = iris, aes(x = var, y = var)) + geom_point()"
  )

  ggpaintr_bind_control_panel(input, output, ggpaintr_state)
  ggpaintr_bind_draw(input, ggpaintr_state)
  ggpaintr_bind_export(output, ggpaintr_state)
  ggpaintr_bind_error(output, ggpaintr_state)
  ggpaintr_bind_code(output, ggpaintr_state)

  output$outputPlot <- renderPlot({
    plot_obj <- ggpaintr_plot_value(ggpaintr_state$runtime())

    if (is.null(plot_obj)) {
      plot.new()
      return(invisible(NULL))
    }

    plot_obj + ggplot2::theme_minimal()
  })
}
```

### Register a custom placeholder type

Use `ggpaintr_placeholder()` and `ggpaintr_effective_placeholders()` when you
want to add a new control type without editing package internals.

```{r custom-placeholder}
sales <- data.frame(
  day = as.Date("2024-01-01") + 0:4,
  value = c(10, 13, 12, 16, 18)
)

date_placeholder <- ggpaintr_placeholder(
  keyword = "date",
  build_ui = function(id, copy, meta, context) {
    shiny::dateInput(id, copy$label)
  },
  resolve_expr = function(value, meta, context) {
    if (is.null(value) || identical(as.character(value), "")) {
      return(ggpaintr_missing_expr())
    }

    rlang::expr(as.Date(!!as.character(value)))
  },
  copy_defaults = list(label = "Choose a date for {param}")
)

placeholders <- ggpaintr_effective_placeholders(
  list(date = date_placeholder)
)

obj <- ggpaintr_formula(
  "ggplot(data = sales, aes(x = day, y = value)) +
    geom_line() +
    geom_vline(xintercept = date)",
  placeholders = placeholders
)

names(obj$placeholders)
```

```r
ggpaintr_app(
  "ggplot(data = sales, aes(x = day, y = value)) +
    geom_line() +
    geom_vline(xintercept = date)",
  placeholders = placeholders
)
```

If you want to export a standalone app with a custom placeholder, define the
placeholder hooks inline inside the `ggpaintr_placeholder()` call. The export
path serializes that stored call into the generated app, so helper functions
that only exist in your current session are not exportable today.

### Advanced developer workflow

Use the low-level runtime helpers directly when you want to inspect generated
code, write tests, or build developer tooling around parsed formulas.

`ggpaintr_runtime_input_spec()` is the supported way to discover the runtime
input ids needed by `ggpaintr_build_runtime()`.

```{r advanced-runtime}
obj <- ggpaintr_formula(
  "ggplot(data = mtcars, aes(x = var, y = var)) +
    geom_point(size = num) +
    labs(title = text)"
)

spec <- ggpaintr_runtime_input_spec(obj)
spec

inputs <- setNames(vector("list", nrow(spec)), spec$input_id)
checkbox_rows <- spec$role == "layer_checkbox"
inputs[checkbox_rows] <- rep(list(TRUE), sum(checkbox_rows))
inputs[["ggplot+3+2"]] <- "mpg"
inputs[["ggplot+3+3"]] <- "disp"
inputs[["geom_point+2"]] <- 2
inputs[["labs+2"]] <- "Mtcars scatter"

runtime <- ggpaintr_build_runtime(obj, inputs)

runtime$code_text
inherits(runtime$plot, "ggplot")
```

For upload-backed formulas, the spec also includes the derived dataset-name
input that accompanies each upload control.

```{r advanced-upload-spec}
upload_obj <- ggpaintr_formula(
  "ggplot(data = upload, aes(x = var, y = var)) + geom_point()"
)

ggpaintr_runtime_input_spec(upload_obj)
```

## Current stability guarantees

- built-in placeholders are `var`, `text`, `num`, `expr`, and `upload`
- custom placeholders share the same parse, UI, runtime, copy-rule, and export
  path as built-ins
- `upload` currently supports `.csv` and `.rds`
- the supported app-facing surface is the current `ggpaintr_*` wrapper and
  Shiny-integration layer
- exported apps keep the current explicit-file shape built around
  `ggpaintr_server()`
- only the six top-level ids exposed by `ggpaintr_ids()` are configurable
- runtime failures are labeled as `Input error:` or `Plot error:` and stay on
  the shared inline error path

## Not guaranteed / implementation details

- raw placeholder ids such as `"ggplot+3+2"` are not a stable hand-authored API;
  discover them with `ggpaintr_runtime_input_spec()`
- deeper traversal details such as `index_path` encoding and internal companion
  id conventions remain package internals
- unsupported upload formats are outside the current boundary
- custom placeholders whose hooks are not defined inline inside
  `ggpaintr_placeholder()` are not exportable as standalone apps today
- the formula-string model is the current author-facing interface; future
  hardening should compile it into a richer internal runtime contract rather
  than replace that authoring model outright

## Current behavior boundary

- Structural formula errors fail early during parsing.
- `var` with no data source fails while preparing the UI.
- `var` is a column picker. Formula-level transforms such as `var + 1` or
  `log(var)` are supported inside the formula text, not as direct input values.
- For local data with non-syntactic names, call
  `ggpaintr_normalize_column_names()` first; uploads apply the same
  normalization automatically.
- Missing local data objects are deferred to draw-time inline errors.
- Advanced integrations can customize the plot through `ggpaintr_plot_value()`.

## Where to go next

See `vignette("ggpaintr-workflow")` for the main workflow and
`vignette("ggpaintr-extensibility")` for supported Shiny integration recipes.
See `vignette("ggpaintr-placeholder-registry")` for the placeholder registry
API, hook contract, and export guidance.
