---
title: "Use cases"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Use cases}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment   = "#>",
  warning   = FALSE,
  message   = FALSE,
  eval      = interactive()
)
```

This vignette answers the question *how do I use ggpaintr for my situation?* Companion vignettes cover tailoring defaults (`vignette("ggpaintr-customization")`), sharing safely (`vignette("ggpaintr-safety")`), real ggplot recipes (`vignette("ggpaintr-gallery")`), and LLM integration (`vignette("ggpaintr-llm")`).

Every chunk is marked `eval = interactive()` and is runnable as-is at the R prompt after:

```{r libs}
library(ggpaintr)
```

## Formula tour

A ggpaintr formula is a single `ggplot()` call written as text. Drop one of five placeholder keywords anywhere a value would normally go, and the runtime substitutes the user's input back into the expression at render time. The five keywords are:

| Keyword  | Becomes                                       | Example formula fragment        | Renders as                                       |
|----------|-----------------------------------------------|---------------------------------|--------------------------------------------------|
| `var`    | column picker (data-aware `selectInput`)      | `aes(x = var, y = var)`         | two dropdowns scoped to the upstream data's columns |
| `text`   | free-text input                               | `labs(title = text)`            | one text field                                   |
| `num`    | numeric input                                 | `geom_point(size = num)`        | one numeric input                                |
| `expr`   | code editor (validated against a denylist)    | `facet_wrap(expr)`              | one expression box; see `vignette("ggpaintr-safety")` for the validation model |
| `upload` | file picker (returns a data frame)            | `ggplot(upload, …)`             | one file picker + an optional dataset-name field |

A formula that exercises four of the five:

```{r formula-tour}
ptr_app("
ggplot(iris, aes(x = var, y = var, color = var)) +
  geom_point(size = num) +
  labs(title = text) +
  facet_wrap(expr)
")
```

The full grammar reference — including positional-argument behavior, `shared = "<id>"` annotation, and the empty-call cleanup rule — lives at `?ptr_app`.

## Data pipelines in the formula

A layer's data argument is allowed to be a pipeline. Both `|>` and `%>%` are accepted, and the two can be mixed in the same chain. Each pipeline stage that contains a placeholder gets its own row of widgets under the layer's **Data** subtab, plus a per-stage enable/disable toggle that rewrites the generated code as if that step were never there. Column pickers downstream of a stage lazily re-resolve their column choices against the current Data-subtab inputs, so changing a filter threshold immediately refreshes the picker below it.

```{r pipeline-formula-tour}
ptr_app(
  "mpg |>
     dplyr::filter(displ > num) |>
     dplyr::group_by(class) |>
     dplyr::filter(dplyr::n() > num) |>
     dplyr::ungroup() |>
     ggplot(aes(var, var, color = class)) +
     geom_point(alpha = num)"
)
```

See `vignette(\"ggpaintr-gallery\")` § 4 for the same example with the original `dplyr` pipeline alongside the parameterized version.

### What we expect to work

- **Both pipe operators**, possibly mixed in the same chain: `df %>% filter(...) |> ggplot(...)`.
- **Bare-symbol stage heads**: `filter(...)`, `head(...)`, `transform(...)`.
- **Namespaced stage heads**: `dplyr::filter(...)`, `tidyr:::pivot_longer(...)`.
- **Parenthesised heads**: `(filter)(...)` — used occasionally to disambiguate function lookup.
- **`upload` as the pipeline head**: `upload |> dplyr::filter(...) |> ggplot(...)`. The upload widget feeds the pipeline; every downstream stage and placeholder resolves against the uploaded data.
- **Per-layer pipelines on multi-layer formulas**: each layer can have its own data-arg pipeline (`ggplot(data = df1 |> filter(...)) + geom_smooth(data = df2 |> filter(...))`) — the Data subtabs are scoped per layer.
- **Placeholders in any stage**: `var`, `text`, `num`, `expr` inside any stage's call resolve as if that stage were a regular layer call.

### What we know does **not** work yet

- **Anonymous functions as pipe stages.** `df |> (\(x) x[x$cyl > 4, ])() |> ggplot(...)` and `df |> (function(x) ...)() |> ggplot(...)` are not supported. The runtime resolves each stage's name from its call head; an inline lambda has no name, so the stage cannot be assigned a stable Shiny id or a layer label, and the Data subtab cannot render. The expression-level safety walker still inspects the lambda body, but the surrounding pipeline machinery does not. Workaround: lift the anonymous function to a named helper in the calling environment, then use it as a bare-symbol stage (`df |> my_helper() |> ggplot(...)`).

## L1 — All-in-one entry point

L1 is the dominant use case: hand ggpaintr a formula, get back a running Shiny app. Two entry points cover the common layouts.

| Entry point        | Use when                                                                 |
|--------------------|--------------------------------------------------------------------------|
| `ptr_app()`        | Default layout. The built-in chrome (title, layer picker, plot pane, code pane) is enough. |
| `ptr_app_grid()`   | Multiple formulas, one app — each plot gets its own tile, optionally with shared widgets at the top. |

For a different page shell or theme, write a thin wrapper on top of the public ggpaintr primitives — see `vignette("ggpaintr-customization")` § "Writing your own wrapper" for the recipe and a worked example.

### Single plot — `ptr_app()`

```{r app-basic}
ptr_app(
  "ggplot(iris, aes(var, var, color = var)) + geom_point() + labs(title = text)"
)
```

### Multiple plots and shared widgets — `ptr_app_grid()`

`ptr_app_grid()` is the turn-key entry point for multi-plot apps with shared widgets. It takes a list of formulas in `plots =` and renders one tile per formula. Annotate any placeholder occurrence with `keyword(shared = "<id>")`; pass `shared_ui = list(<id> = function(id) widget_tag)` so the framework knows what tag to render at the top of the page. The `<id>` namespace is shared across every formula in `plots`, so editing the top widget propagates to every plot.

```{r app-grid-shared}
ptr_app_grid(
  plots = list(
    "ggplot(iris, aes(x = var, y = Sepal.Length, fill = Species)) +
       geom_boxplot()",
    "ggplot(iris, aes(x = var, y = Sepal.Width, fill = Species)) +
       geom_violin()"
  ),
  shared_ui = list()  # no shared= annotations above; the first var stays per-tile
)
```

Adding `shared = "metric"` to both `var` occurrences and supplying `shared_ui = list(metric = function(id) selectInput(id, "Metric", names(iris)))` lifts the column choice into a single widget that drives both plots when the user clicks **Draw all**. The shared mechanism works for any keyword — built-in `num(shared = "sz")` paired with `shared_ui = list(sz = function(id) sliderInput(id, "Size", 1, 10, 3))` is the simplest two-line example; custom consumers work the same way (see `vignette("ggpaintr-customization")` for registering custom widget types).

### Empty-call cleanup

A placeholder that resolves to "missing" (an empty `var` pick, a blank `text`, a cleared `num`) drops its argument from the generated code. When that leaves an empty call whose name is in ggpaintr's curated cleanup list, the whole call disappears too:

```r
ptr_app("
ggplot(iris, aes(x = var, y = var)) +
  geom_point() +
  labs(title = text) +              # title empty → labs() empty → labs() dropped
  facet_wrap(expr)                  # expr empty → facet_wrap() empty → facet_wrap() dropped
")
```

The cleanup list covers safe ggplot2 no-ops: `theme()`, `labs()`, `xlab` / `ylab` / `ggtitle`, `facet_wrap` / `facet_grid` / `facet_null`, `xlim` / `ylim` / `lims`, `expand_limits`, `guides`, `annotate`, and the empty mapping helpers (`aes()`, `vars()`). Third-party helpers (e.g. `pcp_theme()` from `ggpcp`) are *not* in the cleanup list — being absent is the "removal safety unknown" signal. Pass `safe_to_remove = c("pcp_theme")` to opt a specific name into the cleanup pass. An `expr` placeholder whose user supplies an expression always wins: whatever was typed is honoured verbatim, even if its top-level name is in `safe_to_remove`.

### Local data with non-syntactic columns

`var` placeholders need syntactic, unique column names — `Sepal.Length` is fine, `"Sepal Length"` is not, and `iris` deliberately ships clean. If your local data has spaces, reserved words, or duplicates in its names, pipe it through `ptr_normalize_column_names()` before referencing it from a formula:

```{r normalize-cols}
messy <- data.frame(
  check.names = FALSE,
  "first column"  = 1:3,
  "if"            = 4:6
)
clean <- ptr_normalize_column_names(messy)
names(clean)
#> "first.column" "if."

ptr_app("ggplot(clean, aes(x = var, y = var)) + geom_point()")
```

Uploaded data goes through the same normalization automatically; the manual call is only needed for data already in your R session. `?ptr_normalize_column_names` documents the exact rules (`make.names()` plus uniqueness + reserved-word collision avoidance).

## L2 — Embed in your own Shiny app

L2 is "I already have a Shiny app and want to drop a ggpaintr-driven block somewhere inside it." Two paths cover the common cases: a Shiny module (one id, two helpers) or a free-form split (UI + outputs placed wherever you want).

### Module path — `ptr_module_ui()` / `ptr_module_server()`

The compact option. One `id`, controls + outputs live next to each other inside the module's namespace.

```{r module-app}
formula <- "ggplot(iris, aes(var, var, color = var)) + geom_point()"

ui <- shiny::fluidPage(
  shiny::titlePanel("My host app"),
  ptr_module_ui("p", formula)
)
server <- function(input, output, session) {
  ptr_module_server("p", formula)
}

shiny::shinyApp(ui, server)
```

`ptr_module_server()` returns the `ptr_state` from `ptr_init_state()` so embedders that want to react to it (e.g. call `ptr_gg_extra()`, read `ptr_extract_*()`) can capture it; otherwise the side effects are all that matter. `...` is forwarded to `ptr_init_state()` — so `ui_text =`, `checkbox_defaults =`, `expr_check =`, `safe_to_remove =`, and `shared =` are all available here too.

### Non-module path — `ptr_controls_ui()` + `ptr_outputs_ui()` + `ptr_server()`

When you want the controls in one part of the page and the plot in another, split the UI:

```{r split-app}
formula <- "ggplot(iris, aes(var, var, color = var)) + geom_point()"

ui <- shiny::fluidPage(
  shiny::titlePanel("Custom layout"),
  shiny::sidebarLayout(
    shiny::sidebarPanel(
      ptr_controls_ui(formula = formula)   # widgets only
    ),
    shiny::mainPanel(
      shiny::tags$h3("Plot + code + errors"),
      ptr_outputs_ui()                     # plot / error / code outputs
    )
  )
)
server <- function(input, output, session) {
  ptr_server(input, output, session, formula)
}

shiny::shinyApp(ui, server)
```

`ptr_controls_ui()` and `ptr_outputs_ui()` are the two halves that `ptr_module_ui()` renders together; each one self-wraps in a `<div class="ptr-app">` so the bundled stylesheet attaches without any host-side scaffolding. `ptr_server()` registers `ptr_register_plot()` / `_error()` / `_code()` for you — `ptr_outputs_ui()` reserves the three canonical output slots (`ptr_plot`, `ptr_error`, `ptr_code`) that those registrations target.

If you need to embed more than one `ggpaintr` formula on the same page — or share a sidebar with hand-written widgets that might collide — pass a matching `id` to all three: `ptr_controls_ui("p", formula)`, `ptr_outputs_ui("p")`, and `shiny::moduleServer("p", function(input, output, session) ptr_server(input, output, session, formula))`. The single-formula case (the example above) needs neither the `id` nor the `moduleServer()` wrapper.

### Splitting outputs further

`ptr_register_plot()`, `ptr_register_error()`, and `ptr_register_code()` are exposed for embedders who want fewer than three outputs (drop the code pane, for example), or who want to drive the outputs from their own `ptr_init_state()` without `ptr_server()`'s opinionated bundle. Each accepts `(output, state)` and writes to a canonical, namespaced output id; pair with `plotOutput()` / `uiOutput()` / `verbatimTextOutput()` at the namespaced output id.

If you need custom output names — your own `plotOutput("my_chart")` rather than ggpaintr's `ptr_plot` — that is L3 below: read the latest plot with `ptr_extract_plot(state)` inside your own `renderPlot()` block.

### `ptr_init_state()` for advanced init

`ptr_init_state(formula, ...)` is what `ptr_server()` and `ptr_module_server()` call under the hood. Use it directly when you want the state object without auto-registered outputs — for example, to drive a single plot but compose your own code pane somewhere else:

```{r init-state}
server <- function(input, output, session) {
  state <- ptr_init_state(formula = "ggplot(iris, aes(var, var)) + geom_point()")
  ptr_register_plot(output, state)  # only the plot; no error / code panes
}
```

`?ptr_init_state` documents the advanced arguments: `shared` (host-level shared bindings), `draw_trigger` (custom redraw reactive), `producer_debounce_ms` (data-source debouncing), and the namespace controls.

### How input ids are built — and how to avoid colliding with them

The framework names every widget it emits from the formula's syntax, so the same formula always produces the same ids. When you mix ggpaintr UI with your own widgets in the same Shiny namespace, the cleanest fix is to wrap the embed in `ptr_module_ui()` / `ptr_module_server()` with a unique `id` — everything ggpaintr emits then lives under `<id>-…` and your top-level inputs cannot collide. If you can't namespace, the surface to avoid is:

**Layer names.** Each top-level call in the formula contributes a layer name equal to the function being called (`ggplot`, `geom_point`, `facet_wrap`, …). Repeats get `-2`, `-3`, … suffixes:

| Formula fragment                          | Layer names                      |
|-------------------------------------------|----------------------------------|
| `geom_point() + geom_smooth()`            | `geom_point`, `geom_smooth`      |
| `geom_point() + geom_point()`             | `geom_point`, `geom_point-2`     |

**Per-placeholder input ids** follow `<layer>_<path>_<keyword>_<shared-or-NA>`, where `<path>` is the underscore-joined positional index path into the call. For example, `aes(x = var, y = var)` inside the `ggplot()` layer yields `ggplot_1_1_var_NA` and `ggplot_1_2_var_NA`; a shared annotation `var(shared = "x_col")` lands at `ggplot_1_1_var_x_col`.

**Layer-derived ids:**

| Pattern                                          | What it is                                                  |
|--------------------------------------------------|-------------------------------------------------------------|
| `<layer>_checkbox`                               | "include this layer" toggle (no checkbox on `ggplot`)       |
| `<layer>_<path>_stage_enabled`                   | per-stage toggle on `%>%` / `|>` data-arg pipelines         |
| `<placeholder-id>_ui`                            | `renderUI` container for a consumer placeholder             |
| `<layer>_subtab`, `ptr_layer_content_<layer>`    | internal layer-nav and panel-container ids                  |

**Top-level package-owned ids.** The framework writes to a fixed set of `ptr_`-prefixed top-level ids and they are **not** user-configurable: `ptr_plot` (plotOutput), `ptr_error` (uiOutput), `ptr_code` (verbatimTextOutput), `ptr_update_plot` (draw button), plus internal layer-nav inputs `ptr_layer_select` / `ptr_layer_tabset`. Treat the whole `ptr_` prefix as reserved.

> Rule of thumb: in a host app's top-level namespace, don't author Shiny inputs or outputs whose ids start with `ptr_`, or match `<layer-name-from-your-formula>_…`. If that feels brittle, wrap the embed in `ptr_module_ui()` / `ptr_module_server()` with a unique `id` and the collision surface disappears entirely.

## L3 — Custom rendering via extract accessors

L3 is "I want my own renderer." The runtime stays inside ggpaintr; you swap the *output* with whatever Plotly / ggiraph / custom host you prefer.

There are two reading paths off a `ptr_state`, and the choice depends on whether you are inside a reactive context:

| Inside a `renderX({...})` / `reactive({...})` / `observeEvent({...})` | Outside reactive contexts (tests, download handlers, snapshots) |
|----|----|
| Read `state$runtime()` and unpack its slots (`$ok`, `$plot`, `$code_text`, `$error`). This takes the reactive dependency that wires re-renders to *Update plot* clicks. | Call `ptr_extract_plot(state)` / `ptr_extract_code(state)` / `ptr_extract_error(state)`. These accessors wrap `shiny::isolate()` — they return the *current* value without registering a reactive dependency. |

Mixing the two in a render block — calling `ptr_extract_plot(state)` inside `renderPlotly({...})` — silently breaks reactivity (the block fires once on mount and never again). Stick to `state$runtime()` inside reactive blocks.

### Custom plot renderer — `state$runtime()$plot` inside `renderPlotly()`

The pattern: wire `ptr_module_server()` for the runtime, then unpack `state$runtime()` inside a `plotly::renderPlotly()` block (or `ggiraph::renderGirafe()`, or any custom renderer). Place your own output container; omit `ptr_outputs_ui()` if you do not want the bundled plot pane (its auto-registered outputs stay inert when no UI consumer exists).

```{r l3-plotly}
formula <- "ggplot(iris, aes(x = var, y = var, color = var)) + geom_point()"

ui <- shiny::fluidPage(
  shiny::fluidRow(
    shiny::column(5, ptr_module_ui("p", formula)),
    shiny::column(7, plotly::plotlyOutput("custom_plot", height = "500px"))
  )
)
server <- function(input, output, session) {
  state <- ptr_module_server("p", formula)
  output$custom_plot <- plotly::renderPlotly({
    res <- state$runtime()
    shiny::req(isTRUE(res$ok), res$plot)
    plotly::ggplotly(res$plot)
  })
}

shiny::shinyApp(ui, server)
```

`req(isTRUE(res$ok), res$plot)` keeps the panel quiet between draws and on transient error states.

`vignette("ggpaintr-gallery")` §5 *Interactive output packages* is the canonical worked-example reference for the L3 pattern — it shows the full plotly and ggiraph setups end-to-end, including the ggiraph variant with click handling and a custom-placeholder colour picker. The chunk above is the minimal scaffold; gallery §5 is the runnable recipe.

### Custom code panel

Bind to `state$runtime()` inside `renderText({...})` and pull `code_text` off it. Inside the same server as the `state <-` assignment above:

```{r l3-code-pane}
output$my_code <- shiny::renderText({
  state$runtime()$code_text %||% ""
})
```

For non-reactive contexts (a download handler, a snapshot test), use the public accessor `ptr_extract_code(state)`. It returns a single string and includes any `ptr_gg_extra()` expressions captured on the state.

### Custom error UI

`state$runtime()$error` is the latest runtime error string (or `NULL` when the last cycle succeeded). Inside the same server:

```{r l3-error-ui}
output$my_status <- shiny::renderUI({
  msg <- state$runtime()$error
  if (is.null(msg)) shiny::tags$div(class = "status-ok", "Plot is up to date")
  else              shiny::tags$div(class = "status-error", msg)
})
```

`ptr_extract_error(state)` is the non-reactive counterpart.

### Programmatic layer injection — `ptr_gg_extra()`

`ptr_gg_extra(state, ...)` evaluates one or more `ggplot2` expressions and attaches the results as "extras" on the state. The runtime folds them into the rendered plot during the next cycle, and `state$runtime()$code_text` (and `ptr_extract_code(state)`) include them in the printable code. Each call replaces the previously captured extras — pass everything you want layered on in a single call.

```{r l3-gg-extra}
formula <- "ggplot(mtcars, aes(x = mpg, y = hp)) + geom_point()"

ui <- shiny::fluidPage(
  shiny::actionButton("add_log", "Toggle log-scale"),
  ptr_module_ui("p", formula)
)
server <- function(input, output, session) {
  state <- ptr_module_server("p", formula)
  shiny::observeEvent(input$add_log, {
    ptr_gg_extra(state, ggplot2::scale_x_log10())
  })
}

shiny::shinyApp(ui, server)
```

Eval errors from the captured expressions leave the existing extras untouched (atomic update). Extras are also suppressed automatically when the underlying runtime reports a failure, so stale extras from a prior successful draw never surface during an error state.
