---
title: "Custom Placeholders with the ggpaintr Registry"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Custom Placeholders with the ggpaintr Registry}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)

library(ggpaintr)
library(ggplot2)
library(rlang)
library(shiny)
```

```{r clipboard, echo = FALSE}
xaringanExtra::use_clipboard()
```

## Why a placeholder registry?

The five built-in placeholder keywords (`var`, `text`, `num`, `expr`, `upload`)
cover the common cases, but real apps need more: date pickers, sliders bound
to data extents, colour pickers, row-filter widgets, and so on. The
**placeholder registry** is the extension point that lets you add such
widgets without patching the parser, the runtime, or the copy system.

Concretely, the registry lets you:

- **Add a new placeholder keyword.** Register a `date` placeholder and use
  `date` inside any formula, e.g. `geom_vline(xintercept = date)`.
- **Replace a built-in widget.** Swap `var`'s `selectInput()` for
  `shinyWidgets::pickerInput()` without touching package internals.
- **Ship placeholders from another package.** A registry is a plain list, so
  an external package can export `placeholders = ...` that downstream apps
  pass into `ptr_app()`.

Before reading further, make sure you are comfortable with the formula
syntax and the built-in placeholders — those are covered in
`vignette("ggpaintr-workflow")`. This vignette focuses on what you add
**on top** of that foundation.

## The placeholder contract at a glance

A placeholder is the value returned by `ptr_define_placeholder()`. It bundles
up to five hooks and a copy-defaults list. Only two hooks are required:

| Hook | Signature | Required? | Purpose |
|------|-----------|-----------|---------|
| `build_ui` | `function(id, copy, meta, context)` | **required** | Return the Shiny tag for one placeholder occurrence. |
| `resolve_expr` | `function(value, meta, context)` | **required** | Turn the current widget value into an R expression spliced into the generated ggplot call. |
| `resolve_input` | `function(input, id, meta, context)` | optional | Pre-process the raw Shiny input before it reaches `resolve_expr`. Defaults to `input[[id]]`. |
| `bind_ui` | `function(input, output, metas, context)` | optional | Reactive wiring inside the server (e.g. `renderUI`, `observe`, `updateSelectInput`). Runs once per keyword with the list of occurrences. |
| `prepare_eval_env` | `function(input, metas, eval_env, context)` | optional | Inject objects into the environment where the completed formula is evaluated; must return `eval_env`. |

Every hook receives two shared records:

**`meta`** — describes the specific occurrence of the placeholder in the
formula. Fields set by the parser:

- `id` — Shiny input id ggpaintr generated for this occurrence.
- `keyword` — the placeholder keyword (e.g. `"date"`).
- `layer_name` — the ggplot layer, e.g. `"geom_vline"`.
- `param` — the argument name (`"xintercept"`), or `""` for positional args.
- `index_path` — integer vector pointing into the parsed expression tree.

**`context`** — shared read-only environment:

- `context$ptr_obj` — the parsed `ptr_obj`, including `placeholder_map`.
- `context$placeholders` — the active placeholder registry.
- `context$ui_text` — merged copy rules.
- `context$envir` — caller environment (used to resolve data symbols).
- `context$eval_env`, `context$var_column_map` — set in Shiny mode; `NULL`
  elsewhere.

The copy-defaults list may set any of the four leaf fields ggpaintr
understands: `label`, `help`, `placeholder`, `empty_text`. Strings support
`{param}` and `{layer}` interpolation. Use `ptr_missing_expr()`
from `resolve_expr` when an occurrence should be **removed** from the
completed expression (rather than rendered as `NULL`).

## Built-in placeholders

The five built-ins use the same registry path as any custom placeholder.
The table below shows which hooks each one defines — a useful reference
when you want to mimic or replace one.

| Keyword | Widget | Value produced | Hooks used |
|---------|--------|----------------|------------|
| `var` | Dynamic `shinyWidgets::pickerInput` of data columns | Column name as a symbol | `build_ui`, `resolve_expr`, `bind_ui` |
| `text` | `textInput` | String literal | `build_ui`, `resolve_expr` |
| `num` | `numericInput` | Numeric scalar | `build_ui`, `resolve_expr` |
| `expr` | Text field parsed as R | Arbitrary expression (safety-checked) | `build_ui`, `resolve_expr` |
| `upload` | File input + format picker | User-supplied data frame bound in eval env | `build_ui`, `resolve_expr`, `resolve_input`, `prepare_eval_env` |

`var` uses `bind_ui` because its choices can only be populated after the
data is known at runtime. `upload` uses `prepare_eval_env` because the
uploaded data frame has to live in the eval environment under the name the
user typed.

## A minimal custom placeholder — a date picker

Here is a complete end-to-end example: define a `date` placeholder, use it
in a formula, inspect the generated metadata, and launch a Shiny app that
drives it.

```{r date-placeholder}
sales <- data.frame(
  day = as.Date("2024-01-01") + 0:4,
  value = c(10, 13, 12, 16, 18)
)

date_placeholder <- ptr_define_placeholder(
  keyword = "date",
  build_ui = function(id, copy, meta, context) {
    shiny::dateInput(id, copy$label)
  },
  resolve_expr = function(value, meta, context) {
    if (is.null(value) || identical(as.character(value), "")) {
      return(ptr_missing_expr())
    }
    rlang::expr(as.Date(!!as.character(value)))
  },
  copy_defaults = list(label = "Choose a date for {param}")
)
```

`ptr_merge_placeholders()` combines your custom entries with the built-ins
so the registry you pass downstream still understands `var`, `text`, etc.
Custom entries with a built-in keyword override the built-in.

```{r merge-placeholders}
placeholders <- ptr_merge_placeholders(
  list(date = date_placeholder)
)

names(placeholders)
```

Parse a formula with the enriched registry and inspect the metadata record
the parser built for the `date` occurrence:

```{r date-parse}
obj <- ptr_parse_formula(
  "ggplot(data = sales, aes(x = day, y = value)) +
    geom_line() +
    geom_vline(xintercept = date)",
  placeholders = placeholders
)

obj$placeholder_map$geom_vline[["geom_vline_2"]]
```

The fields match the `meta` record every hook receives — `id`, `keyword`,
`layer_name`, `param`, `index_path`.

Execute the plot headlessly with a concrete date. Prefer
`ptr_runtime_input_spec(obj)` for discovering the expected input ids
programmatically rather than hard-coding the `"geom_vline_2"` pattern used
below for clarity.

```{r date-exec}
runtime <- ptr_exec(
  obj,
  list(
    "geom_line_checkbox"  = TRUE,
    "geom_vline_checkbox" = TRUE,
    "geom_vline_2"        = as.Date("2024-01-03")
  )
)

runtime$code_text
```

To launch a full Shiny app that drives the same placeholder interactively,
use `ptr_app()` (see the final section for every API that accepts
`placeholders = ...`):

```{r date-app, eval = FALSE}
ptr_app(
  "ggplot(data = sales, aes(x = day, y = value)) +
    geom_line() +
    geom_vline(xintercept = date)",
  placeholders = placeholders
)
```

## The five hooks in detail

### `build_ui(id, copy, meta, context)` — required

Returns the Shiny tag for one placeholder occurrence. Called once per
occurrence when ggpaintr constructs the control panel. `id` is the Shiny
input id you must bind; `copy` is a named list with the four leaf fields
(`label`, `help`, `placeholder`, `empty_text`) already resolved for this
occurrence.

```{r build-ui-peek}
date_placeholder$build_ui
```

`build_ui` has no access to reactives. If the widget needs to depend on
current input values, return a container tag (e.g. `shiny::uiOutput()`)
and finish wiring it inside `bind_ui()`.

### `resolve_expr(value, meta, context)` — required

Turns the current widget value into an R expression that replaces the
placeholder token in the completed code. Use `!!` (`rlang` unquoting) to
splice the runtime value into the returned expression. Return
`ptr_missing_expr()` to **drop** the argument from the completed
expression entirely.

The returned value must be one of: a call, symbol, pairlist, character
vector, or a scalar (`numeric`, `logical`, `integer`, `double`,
`complex`, or `NULL`). Anything else is rejected with a clear error.

### `resolve_input(input, id, meta, context)` — optional

Runs before `resolve_expr` and lets you massage the raw Shiny input value.
When omitted, ggpaintr passes `input[[id]]` straight through.

```{r resolve-input-example, eval = FALSE}
trimmed_text <- ptr_define_placeholder(
  keyword = "trimmed_text",
  build_ui = function(id, copy, meta, context) {
    shiny::textInput(id, copy$label)
  },
  resolve_input = function(input, id, meta, context) {
    trimws(input[[id]])
  },
  resolve_expr = function(value, meta, context) {
    if (identical(value, "")) return(ptr_missing_expr())
    rlang::expr(!!value)
  }
)
```

### `bind_ui(input, output, metas, context)` — optional

`bind_ui` is called once per keyword inside the Shiny server with the full
list of occurrences (`metas`). Use it to register `renderUI()` outputs or
`observe()` blocks for widgets that depend on reactive state. The return
value is a named list of `renderUI()`-capable tags keyed by `meta$id`,
which ggpaintr uses for deferred rendering; returning `NULL` or
`invisible(NULL)` is fine.

### `prepare_eval_env(input, metas, eval_env, context)` — optional

Runs immediately before the completed expression is evaluated. Use it to
bind objects into the eval environment — uploaded data frames, helper
functions, lookup tables. **The hook must return the modified
`eval_env`**; ggpaintr does not mutate it in place.

## A more complex example — `bind_ui` + `prepare_eval_env`

This example defines a `log_num` placeholder that chooses a numeric column
from the formula's `data` and wraps it in `log()` in the generated code.
It uses `bind_ui` to populate a `selectInput` with the data's numeric
columns, and `prepare_eval_env` to expose a small safe-log helper so the
generated call stays readable.

```{r log-num-placeholder}
log_num <- ptr_define_placeholder(
  keyword = "log_num",

  # Return a container; fill it in from bind_ui once data is available.
  build_ui = function(id, copy, meta, context) {
    shiny::uiOutput(paste0(id, "_container"))
  },

  # Populate the selectInput with numeric columns drawn from the data
  # symbol referenced in the formula's ggplot() layer.
  bind_ui = function(input, output, metas, context) {
    eval_env <- context$eval_env
    ggplot_expr <- context$ptr_obj$expr_list[["ggplot"]]
    data_expr <- if (is.call(ggplot_expr)) ggplot_expr$data else NULL
    numeric_cols <- character(0)
    if (!is.null(eval_env) && !is.null(data_expr)) {
      d <- tryCatch(eval(data_expr, envir = eval_env), error = function(e) NULL)
      if (is.data.frame(d)) {
        numeric_cols <- names(d)[vapply(d, is.numeric, logical(1))]
      }
    }
    for (meta in metas) {
      local({
        m <- meta
        output[[paste0(m$id, "_container")]] <- shiny::renderUI({
          shiny::selectInput(
            m$id,
            label   = paste("Log-transform column for", m$param),
            choices = numeric_cols
          )
        })
      })
    }
    invisible(NULL)
  },

  # Generate `.log_safe(<col>)` in the completed expression.
  resolve_expr = function(value, meta, context) {
    if (is.null(value) || identical(value, "")) {
      return(ptr_missing_expr())
    }
    rlang::expr(.log_safe(!!rlang::sym(value)))
  },

  # Bind the helper the generated call references by name.
  prepare_eval_env = function(input, metas, eval_env, context) {
    eval_env$.log_safe <- function(x) log(pmax(x, .Machine$double.eps))
    eval_env
  },

  copy_defaults = list(label = "Log-transform column for {param}")
)
```

A formula using it, parsed with the merged registry:

```{r log-num-parse}
registry <- ptr_merge_placeholders(list(log_num = log_num))

obj <- ptr_parse_formula(
  "ggplot(data = mtcars, aes(x = wt, y = log_num)) + geom_point()",
  placeholders = registry
)

obj$formula_text
```

Running the same formula headlessly — note that `.log_safe` resolves
because `prepare_eval_env()` bound it in the eval environment:

```{r log-num-exec}
runtime <- ptr_exec(
  obj,
  list(
    "geom_point_checkbox" = TRUE,
    "ggplot_3_3"          = "mpg"
  )
)

runtime$code_text
```

Launch the live app to exercise `bind_ui`:

```{r log-num-app, eval = FALSE}
ptr_app(
  "ggplot(data = mtcars, aes(x = wt, y = log_num)) + geom_point()",
  placeholders = registry
)
```

## Replacing a built-in placeholder

Any custom keyword that matches a built-in one wins — the built-in is
evicted from the merged registry. Below, we replace `var` with a
`shinyWidgets::pickerInput()` that offers search and tick-boxes while
keeping the original reactive behaviour (columns populated from data).

```{r replace-var}
picker_var <- ptr_define_placeholder(
  keyword = "var",

  # Container filled in by bind_ui once data is known.
  build_ui = function(id, copy, meta, context) {
    shiny::uiOutput(paste0(id, "_container"))
  },

  # Produce the same symbol the default `var` does.
  resolve_expr = function(value, meta, context) {
    if (is.null(value) || identical(value, "") || identical(value, "(none)")) {
      return(ptr_missing_expr())
    }
    rlang::sym(value)
  },

  # Mirror the reactive-column behaviour of the default `var`.
  bind_ui = function(input, output, metas, context) {
    eval_env <- context$eval_env
    ggplot_expr <- context$ptr_obj$expr_list[["ggplot"]]
    data_expr <- if (is.call(ggplot_expr)) ggplot_expr$data else NULL
    cols <- character(0)
    if (!is.null(eval_env) && !is.null(data_expr)) {
      d <- tryCatch(eval(data_expr, envir = eval_env), error = function(e) NULL)
      if (is.data.frame(d)) {
        cols <- names(d)
      }
    }
    for (meta in metas) {
      local({
        m <- meta
        output[[paste0(m$id, "_container")]] <- shiny::renderUI({
          shinyWidgets::pickerInput(
            inputId = m$id,
            label   = paste("Column for", m$param),
            choices = c("(none)", cols),
            options = shinyWidgets::pickerOptions(liveSearch = TRUE)
          )
        })
      })
    }
    invisible(NULL)
  },

  copy_defaults = list(
    label      = "Column for {param}",
    empty_text = "(none)"
  )
)

replacement_registry <- ptr_merge_placeholders(list(var = picker_var))
identical(replacement_registry$var, picker_var)
```

```{r replace-var-app, eval = FALSE}
ptr_app(
  "ggplot(data = mtcars, aes(x = var, y = var)) + geom_point()",
  placeholders = replacement_registry
)
```

## Plugging a registry into every API

Every public entry point that accepts formulas also accepts a
`placeholders = ...` argument holding a merged registry. Pass the same
registry object everywhere so built-in widgets, custom widgets, and copy
rules stay in sync.

### `ptr_app()` and `ptr_app_bslib()` — Level 1

```{r api-ptr-app, eval = FALSE}
ptr_app(
  "ggplot(data = sales, aes(x = day, y = value)) +
    geom_line() +
    geom_vline(xintercept = date)",
  placeholders = placeholders
)

ptr_app_bslib(
  "ggplot(data = sales, aes(x = day, y = value)) +
    geom_line() +
    geom_vline(xintercept = date)",
  placeholders = placeholders
)
```

### `ptr_parse_formula()` + `ptr_exec()` — Level 3 headless

Useful for batch reports, tests, and custom pipelines. The registry is
stored on the returned `ptr_obj`, so downstream calls such as
`ptr_runtime_input_spec()` and `ptr_exec()` pick it up automatically.

```{r api-parse-exec}
obj <- ptr_parse_formula(
  "ggplot(data = sales, aes(x = day, y = value)) +
    geom_line() +
    geom_vline(xintercept = date)",
  placeholders = placeholders
)

spec <- ptr_runtime_input_spec(obj)
names(spec)
```

### Manual embedding with `ptr_input_ui()` / `ptr_server_state()` — Level 2

For embedding ggpaintr inside an existing Shiny app, pass `placeholders`
through `ptr_server_state()` — it flows from there into the controls,
plot, and code registrations automatically.

```{r api-manual-embed, eval = FALSE}
ids <- ptr_build_ids(
  control_panel = "builder_controls",
  draw_button   = "render_plot",
  plot_output   = "main_plot",
  error_output  = "main_error",
  code_output   = "main_code"
)

ui <- shiny::fluidPage(
  shiny::sidebarLayout(
    shiny::sidebarPanel(ptr_input_ui(ids = ids)),
    shiny::mainPanel(ptr_output_ui(ids = ids))
  )
)

server <- function(input, output, session) {
  ptr_state <- ptr_server_state(
    "ggplot(data = sales, aes(x = day, y = value)) +
      geom_line() +
      geom_vline(xintercept = date)",
    ids          = ids,
    placeholders = placeholders
  )

  ptr_setup_controls(input, output, ptr_state, ids = ids)
  ptr_register_draw(input, ptr_state, ids = ids)
  ptr_register_plot(output, ptr_state, ids = ids)
  ptr_register_error(output, ptr_state, ids = ids)
  ptr_register_code(output, ptr_state, ids = ids)
}

shiny::shinyApp(ui, server)
```

See `vignette("ggpaintr-extensibility")` §3 for the full Level-2 embed
walkthrough.

## Copy defaults inside a placeholder definition

`copy_defaults` ships default copy with the placeholder itself, so every
downstream app that uses it gets sensible labels without extra work. The
supported fields match the four leaf fields in `ptr_resolve_ui_text()`:
`label`, `help`, `placeholder`, `empty_text`. Strings may reference
`{param}` and `{layer}`.

```{r copy-defaults-example}
annotated_date <- ptr_define_placeholder(
  keyword  = "date",
  build_ui = function(id, copy, meta, context) {
    shiny::tagList(
      shiny::dateInput(id, copy$label),
      if (nzchar(copy$help)) shiny::helpText(copy$help)
    )
  },
  resolve_expr = function(value, meta, context) {
    if (is.null(value)) return(ptr_missing_expr())
    rlang::expr(as.Date(!!as.character(value)))
  },
  copy_defaults = list(
    label = "Choose a {param} date",
    help  = "Used by {layer}."
  )
)

# ptr_resolve_ui_text() shows the final text a `build_ui` call would receive.
ptr_resolve_ui_text(
  component    = "control",
  keyword      = "date",
  layer_name   = "geom_vline",
  param        = "xintercept",
  placeholders = ptr_merge_placeholders(list(date = annotated_date))
)
```

Everything beyond `copy_defaults` — the three-layer merge precedence
(`defaults` / `params` / `layers`), `ptr_merge_ui_text()` and the `shell`
chrome copy — is covered in `vignette("ggpaintr-extensibility")` §3d. That
section walks through the full override catalogue and shows a worked
example of a single placeholder receiving different copy under each layer.

## Summary

1. A placeholder is a bundle of two required hooks (`build_ui`,
   `resolve_expr`) and three optional ones (`resolve_input`, `bind_ui`,
   `prepare_eval_env`) plus `copy_defaults`.
2. `ptr_merge_placeholders()` turns your custom entries into a registry
   that still knows the built-ins, and custom entries win on name clash.
3. Every public API that accepts formulas also accepts a
   `placeholders = ...` argument.
4. For copy customisation beyond `copy_defaults`, use the `ui_text`
   system — see `vignette("ggpaintr-extensibility")` §3d.
