---
title: "ggpaintr Extensibility"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ggpaintr Extensibility}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)

library(ggpaintr)
```

## Overview

`ggpaintr` still ships the default `ptr_app()` and `ptr_server()`
wrappers, but it now also exposes a supported integration layer for users who
already have a Shiny app and want to embed the generated `ggpaintr` controls,
runtime, and export behavior into their own UI.

The supported integration pieces are:

- `ptr_build_ids()`
- `ptr_server_state()`
- `ptr_register_controls()`
- `ptr_register_draw()`
- `ptr_register_export()`
- `ptr_register_plot()`
- `ptr_register_error()`
- `ptr_register_code()`
- `ptr_extract_plot()`
- `ptr_extract_error()`
- `ptr_extract_code()`
- `ptr_input_ui()`
- `ptr_output_ui()`

## Choosing the right surface

The current package surface is intentionally layered.

- use `ptr_app()` or `ptr_server()` for most apps
- use `ptr_server_state()` plus the `ptr_bind_*()` helpers when you
  already own a larger Shiny layout and want to embed `ggpaintr`
- use the low-level `ptr_*` runtime helpers, together with
  `ptr_runtime_input_spec()`, when you are writing tests, tooling, or
  package-level extensions around parsed formulas

That split is a little opinionated, but it keeps the beginner path small while
still exposing a supported integration seam for more advanced Shiny work.

## Recipe 1: Embed `ggpaintr` with the default binders

```{r, eval = FALSE}
library(ggpaintr)
library(shiny)

ui <- fluidPage(
  titlePanel("Embedded ggpaintr"),
  sidebarLayout(
    sidebarPanel(
      ptr_input_ui()
    ),
    mainPanel(
      ptr_output_ui()
    )
  )
)

server <- function(input, output, session) {
  ptr_state <- ptr_server_state(
    "ggplot(data = iris, aes(x = var, y = var)) +
      geom_point() +
      labs(title = text)"
  )

  ptr_register_controls(input, output, ptr_state)
  ptr_register_draw(input, ptr_state)
  ptr_register_export(output, ptr_state)
  ptr_register_plot(output, ptr_state)
  ptr_register_error(output, ptr_state)
  ptr_register_code(output, ptr_state)
}

shinyApp(ui, server)
```

This is the closest supported equivalent to `ptr_server()`, but it lets
you place the controls and outputs inside your own page layout.

## Recipe 2: Customize top-level ids

Use `ptr_build_ids()` when your app already has its own naming scheme or you want
to mount `ggpaintr` outputs in specific UI containers.

```{r, eval = FALSE}
ids <- ptr_build_ids(
  control_panel = "builder_controls",
  draw_button = "render_plot",
  export_button = "export_app",
  plot_output = "main_plot",
  error_output = "main_error",
  code_output = "main_code"
)

ids
```

```{r, eval = FALSE}
ui <- fluidPage(
  sidebarLayout(
    sidebarPanel(
      ptr_input_ui(ids = ids)
    ),
    mainPanel(
      ptr_output_ui(ids = ids)
    )
  )
)

server <- function(input, output, session) {
  ptr_state <- ptr_server_state(
    "ggplot(data = iris, aes(x = var, y = var)) + geom_point()",
    ids = ids
  )

  ptr_register_controls(input, output, ptr_state, ids = ids)
  ptr_register_draw(input, ptr_state, ids = ids)
  ptr_register_export(output, ptr_state, ids = ids)
  ptr_register_plot(output, ptr_state, ids = ids)
  ptr_register_error(output, ptr_state, ids = ids)
  ptr_register_code(output, ptr_state, ids = ids)
}
```

Only the six top-level ids are configurable in this phase. Internal placeholder
ids and dynamic `var-*` output ids remain package-owned.

## Recipe 3: Customize the returned plot in `renderPlot()`

The default `ptr_register_plot()` binder preserves the current wrapper
behavior. If you want to modify the built plot before rendering, write your own
`renderPlot()` and use `ptr_extract_plot()`.

```{r, eval = FALSE}
server <- function(input, output, session) {
  ptr_state <- ptr_server_state(
    "ggplot(data = iris, aes(x = var, y = var)) + geom_point()"
  )

  ptr_register_controls(input, output, ptr_state)
  ptr_register_draw(input, ptr_state)
  ptr_register_export(output, ptr_state)
  ptr_register_error(output, ptr_state)
  ptr_register_code(output, ptr_state)

  output$outputPlot <- renderPlot({
    plot_obj <- ptr_extract_plot(ptr_state$runtime())

    if (is.null(plot_obj)) {
      plot.new()
      return(invisible(NULL))
    }

    plot_obj + ggplot2::theme_minimal()
  })
}
```

`ptr_extract_plot()` returns the raw `ggplot` object on success and `NULL`
otherwise. That keeps the advanced customization seam side-effect free and lets
you decide how to render failure states in your own app.

## Pure helpers versus bind helpers

Use the bind helpers when you want the standard `ggpaintr` behavior with custom
layout or custom top-level ids.

Use the pure value helpers when you want to own the rendering details:

- `ptr_extract_plot()` for plot customization
- `ptr_extract_error()` for custom `renderUI()` workflows
- `ptr_extract_code()` for custom `renderText()` or downstream processing

## How to improve the advanced surface

The current advanced surface is powerful, but it is still closer to package
developer tooling than to a polished app-builder DSL. The next improvements
should stay incremental:

- improve docs and helper discoverability first
- add higher-level helper constructors for common placeholder patterns only
  after the current low-level registry contract has settled a bit more
- consider an optional Shiny module wrapper on top of the current id-based
  helpers rather than replacing the existing supported surface
- do not introduce a separate declarative end-user spec in this phase

## Roadmap for the formula-string runtime

The formula-string model is still the author-facing interface. To make the
runtime easier to reason about without giving up that authoring model, the next
hardening step should happen internally:

- keep formula strings as the public authoring format
- compile each parsed `ptr_obj` once into a richer internal runtime contract
- have runtime completion consume that compiled contract instead of repeatedly
  relying on raw expression walks and companion-id conventions

That approach would tighten the runtime semantics without forcing users to
rewrite formulas into a separate declarative language.
