---
title: "ggpaintr Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ggpaintr Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

## Overview

`ggpaintr` now focuses on the `ggpaintr` workflow. A plot is described as one
ggplot-like formula string. Placeholder tokens inside that string are converted
into Shiny controls, completed with current inputs, evaluated into a plot, and
rendered with generated code and readable runtime feedback.

Supported placeholders are:

- `var`
- `text`
- `num`
- `expr`
- `upload`

`upload` currently supports `.csv` and `.rds`.

## Parsing a formula

```{r}
library(ggpaintr)
library(ggplot2)

obj <- paintr_formula(
  "ggplot(data = iris, aes(x = var, y = var)) +
    geom_point(aes(color = var), size = num) +
    labs(title = text) +
    facet_wrap(expr)"
)

names(obj$expr_list)
names(obj$keywords_list$ggplot)
```

## Building runtime output

```{r}
spec <- ggpaintr_runtime_input_spec(obj)
spec

inputs <- setNames(vector("list", nrow(spec)), spec$input_id)
inputs[spec$role == "layer_checkbox"] <- rep(list(TRUE), sum(spec$role == "layer_checkbox"))
inputs[[spec$input_id[spec$layer_name == "ggplot" & spec$param_key == "x"]]] <- "Sepal.Length"
inputs[[spec$input_id[spec$layer_name == "ggplot" & spec$param_key == "y"]]] <- "Sepal.Width"
inputs[[spec$input_id[spec$layer_name == "geom_point" & spec$param_key == "color"]]] <- "Species"
inputs[[spec$input_id[spec$layer_name == "geom_point" & spec$keyword == "num"]]] <- 2.5
inputs[[spec$input_id[spec$layer_name == "labs" & spec$param_key == "title"]]] <- "Iris scatter"
inputs[[spec$input_id[spec$layer_name == "facet_wrap" & spec$keyword == "expr"]]] <- "~ Species"

runtime <- paintr_build_runtime(
  obj,
  inputs
)

runtime$code_text
inherits(runtime$plot, "ggplot")
```

`ggpaintr_runtime_input_spec()` is the supported way to discover the low-level
runtime inputs consumed by `paintr_build_runtime()`. Raw ids still appear in the
returned data, but they should be treated as implementation details rather than
as hand-authored examples.

## Launching an app

```{r, eval = FALSE}
ggpaintr_app("
ggplot(data = iris, aes(x = var, y = var)) +
  geom_point(aes(color = var), size = num) +
  labs(title = text) +
  facet_wrap(expr)
")
```

The generated app:

- renders one control tab per layer
- delays `var` selectors until upload-backed data is available
- clears the plot on failure
- shows `Input error:` and `Plot error:` messages inline
- preserves generated code when completion succeeds but plotting fails

## Exporting a standalone app

```{r}
out_file <- tempfile(fileext = ".R")
generate_shiny(obj, list(), out_file, style = FALSE)
file.exists(out_file)
```

## Current behavior boundary

- Structural parse failures still stop the app before launch.
- `var` placeholders still require a data source during UI preparation.
- Missing local data objects such as `unknown_object` are deferred to draw time.
- Render-time ggplot failures such as missing faceting variables are surfaced by
  the same runtime error path as other plot failures.

For recipes that embed `ggpaintr` into an existing Shiny app with custom ids or
custom plot rendering, see `vignette("ggpaintr-extensibility")`.
