---
title: "Customization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Customization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment   = "#>",
  warning   = FALSE,
  message   = FALSE,
  eval      = interactive()
)
```

This vignette answers the question *how do I tailor ggpaintr's defaults?* Companion vignettes cover the use-case ladder (`vignette("ggpaintr-use-cases")`), the safety story (`vignette("ggpaintr-safety")`), real ggplot recipes (`vignette("ggpaintr-gallery")`), and LLM integration (`vignette("ggpaintr-llm")`).

Every chunk is marked `eval = interactive()` and is runnable as-is at the R prompt after:

```{r libs}
library(ggpaintr)
```

## Rewriting copy with `ui_text`

Every label, placeholder, and help string in the generated UI comes from a deep-merged copy-rules tree. ggpaintr ships a set of defaults; you override any subset by passing `ui_text =` to the entry point. Only the keys you spell out change — everything else stays at its built-in value.

### The six override sections

| Section          | What it controls |
|------------------|------------------|
| `shell`          | Outer chrome — app title, draw / draw-all buttons, layer picker label, data / controls subtab labels. |
| `upload`         | The `upload` placeholder widget — file-picker label, optional-dataset-name label / placeholder / help text. |
| `layer_checkbox` | The "Include this layer in the plot" toggle on each layer panel. |
| `defaults`       | Per-keyword copy defaults — one entry per registered placeholder, including custom ones (auto-populated from `copy_defaults` passed at registration time). |
| `params`         | Per-parameter overrides — copy that applies when a placeholder appears as the `x`, `color`, `title`, `alpha`, … argument of a layer. |
| `layers`         | Per-layer overrides — copy that applies to one placeholder occurrence inside one specific layer call (e.g. the positional `expr` of `facet_wrap()`). |

The first three sections are flat: one fixed lookup path per UI slot. The last three are nested by keyword — and, for `layers`, by layer name × keyword × parameter — letting you scope an override to a single placeholder, a single parameter, or a single layer.

### Merge precedence

When the runtime needs the copy for a *control* (the widget rendered for one placeholder occurrence in one layer), it deep-merges three rules in order:

1. `defaults[[keyword]]` — most general.
2. `params[[param_key]][[keyword]]` — keyed by the surrounding parameter name.
3. `layers[[layer_name]][[keyword]][[param_key]]` — keyed by the layer the placeholder lives in.

Each later rule wins on fields it specifies; fields it omits fall back to the earlier rule. For the other UI slots (chrome, upload widget, layer checkbox) the resolution path is fixed — the override either replaces a leaf field or does not.

### Building overrides

```{r ui-text-example}
custom_text <- ptr_ui_text(list(
  shell = list(
    title       = list(label = "Iris explorer"),
    draw_button = list(label = "Render")
  ),
  params = list(
    x     = list(var  = list(label = "X variable")),
    title = list(text = list(label = "Plot heading"))
  )
))

ptr_app(
  "ggplot(iris, aes(x = var, y = var, color = var)) +
     geom_point() +
     labs(title = text)",
  ui_text = custom_text
)
```

`ptr_ui_text()` validates the partial override and returns a fully-merged `ptr_ui_text` object. Passing the bare list (without wrapping it in `ptr_ui_text()`) works too — the entry points call it internally — but using it explicitly catches typos in section keys at registration time rather than at render time.

`ptr_resolve_ui_text(component, keyword, param, layer_name, ui_text)` is the lower-level lookup the runtime uses to fetch the resolved copy for a single UI slot. Embedders writing custom UIs that mix ggpaintr-driven labels with their own copy can call it directly; most users do not need it.

## Styling with CSS

ggpaintr ships a bundled stylesheet (`ggpaintr.css`) and a stable, BEM-style class vocabulary under the `ptr-*` prefix. Three customization surfaces, in order of how invasive they are:

### Surface 1 — class hooks ggpaintr already emits

Every region of the default shell carries a stable class you can target from your own CSS. The vocabulary is BEM (`block`, `block__element`, `block--modifier`) and is committed surface; future releases will not silently rename these.

| Region                  | Block             | Notable elements                                                                                              |
|-------------------------|-------------------|---------------------------------------------------------------------------------------------------------------|
| Outer shell             | `.ptr-app`        | `.ptr-app__header`, `.ptr-app__mark`, `.ptr-app__title`                                                       |
| Plot card               | `.ptr-card`       | `.ptr-card--plot`, `.ptr-card__head`, `.ptr-card__title`, `.ptr-card__body`                                   |
| Output area             | `.ptr-output`     | —                                                                                                             |
| Generated-code window   | `.ptr-code-window`| `.ptr-code-window__head`, `.ptr-code-window__title`, `.ptr-code-window__actions`, `.ptr-code-window__close`, `.ptr-code-window__body`, `.ptr-copy-btn`, `.ptr-code-toggle`, `.ptr-icon-btn` |
| Shared-controls panel   | `.ptr-shared-panel` | `.ptr-shared-panel__title`, `.ptr-shared-panel__hint`                                                       |
| Pipeline stage rows     | `.ptr-stage`      | `.ptr-stage-head`, `.ptr-stage-fields`, `.ptr-stage-row`                                                      |
| Inline errors           | `.ptr-alert`      | `.ptr-alert--error`, `.ptr-alert__detail`                                                                     |

### Surface 2 — injecting your own stylesheet with `css =`

Every public entry point (`ptr_app()`, `ptr_app_grid()`, `ptr_module_ui()`, `ptr_controls_ui()`, `ptr_outputs_ui()`) accepts a `css =` argument: a character vector of paths to `.css` files. Each file is registered as a static resource and linked **after** the bundled stylesheet, so your rules win at equal specificity. Relative `url(...)` references resolve against each file's own directory, so you can ship images next to the stylesheet.

```{r css-example, eval = FALSE}
ptr_app(
  "ggplot(mtcars, aes(var, var)) + geom_point()",
  css = "my-theme.css"
)
```

A minimal `my-theme.css` that recolours the title bar and tightens the card padding:

```css
.ptr-app__header { background: #1f2937; color: #f9fafb; }
.ptr-app__title  { font-family: "Inter", sans-serif; letter-spacing: -0.01em; }
.ptr-card__body  { padding: 0.5rem 0.75rem; }
```

The `css =` argument validates each path: missing files and non-`.css` extensions abort at construction time with a clear message — typos surface before first browser load.

### What is **not** stable surface

The `.ptr-*` table above is the committed contract. Outside that:

- Internal markup details — element order, nesting depth, descendant tag names — may change between releases even when class names stay. Style by class, not by structural selectors like `.ptr-card > div:nth-child(2)`.
- Classes emitted by upstream Shiny, shinyWidgets, and bslib (`.shiny-input-container`, `.selectize-control`, `.form-control`, `.btn-primary`, …) belong to those packages, not ggpaintr — their stability is governed there.
- Anything matching `^ptr_` is an **id** (Shiny input/output id), not a class; styling against Shiny ids is allowed but brittle, since ids are namespaced by host modules.

## Writing your own wrapper

When CSS alone is not enough — you want a different page shell, a theme system like `bslib`, or a layout that does not map onto `ptr_app()`'s defaults — write a thin wrapper around the public ggpaintr primitives. The three pieces a single-formula wrapper composes are:

- `ptr_controls_ui(id, formula, ...)` — the layer picker, per-layer parameter panels, and the "Update plot" button, packaged in a `<div class="ptr-app">` so the bundled stylesheet attaches.
- `ptr_outputs_ui(id)` — the plot pane, inline error, and generated-code reveal, same wrapper.
- `ptr_module_server(id, formula, ...)` — the server side. Forwards `ui_text` / `checkbox_defaults` / `expr_check` / `safe_to_remove` / `envir` to [ptr_init_state()] and wires all the reactivity, including shared widgets declared with `var(shared = "...")` inside the same formula.

The package ships one worked example as an exported function, `ptr_app_bslib()`. It is not a recommended primary entry point — `ptr_app()` and `ptr_app_grid()` are. Its purpose is to *demonstrate* the wrapper pattern. The entire source is short enough to read in one screen:

```{r ptr-app-bslib-source, eval = FALSE}
ptr_app_bslib <- function(formula,
                          envir = parent.frame(),
                          ui_text = NULL,
                          checkbox_defaults = NULL,
                          expr_check = TRUE,
                          safe_to_remove = character(),
                          theme = NULL) {
  if (!requireNamespace("bslib", quietly = TRUE)) {
    rlang::abort("Package 'bslib' is required for ptr_app_bslib().")
  }
  if (is.null(theme)) {
    theme <- bslib::bs_theme(version = 5, bootswatch = "flatly")
  }
  id <- "ptr"
  title <- ptr_resolve_ui_text("title", ui_text = ui_text)$label %||% "ggpaintr"

  ui <- bslib::page_sidebar(
    title = title,
    theme = theme,
    sidebar = bslib::sidebar(
      title = "Controls",
      ptr_controls_ui(id = id, formula = formula, ui_text = ui_text,
                      checkbox_defaults = checkbox_defaults,
                      expr_check = expr_check)
    ),
    bslib::card(full_screen = TRUE, ptr_outputs_ui(id = id))
  )

  server <- function(input, output, session) {
    ptr_module_server(id, formula = formula, envir = envir,
                      ui_text = ui_text,
                      checkbox_defaults = checkbox_defaults,
                      expr_check = expr_check,
                      safe_to_remove = safe_to_remove)
  }

  shiny::shinyApp(ui = ui, server = server)
}
```

Everything ggpaintr-specific in that source is on the public API. The only non-ggpaintr pieces are the `bslib::page_sidebar()` shell and the `theme =` passthrough to `bslib::bs_theme()`; those belong to `bslib`, not to ggpaintr. A wrapper is free to expose any downstream-library args alongside the ggpaintr ones it forwards.

### Wrappers compose

Wrappers built this way compose freely. A "dark mode" variant is one line:

```{r ptr-app-dark, eval = FALSE}
ptr_app_dark <- function(formula, ...) {
  ptr_app_bslib(
    formula,
    theme = bslib::bs_theme(version = 5, bootswatch = "darkly"),
    ...
  )
}
```

The same template extends to other shells (`bslib::page_navbar()`, a custom `shiny::fluidPage()` with bespoke chrome, a `flexdashboard`-style page) — swap the UI half and leave the `ptr_module_server()` half alone.

### Public-API limits

Two cases the wrapper pattern above does **not** cover on the public API today:

- **Split output layouts.** `ptr_outputs_ui()` bundles the plot, inline error, and generated-code panes as a single block. Layouts that need to place the plot and code in separate containers (for example, two side-by-side `bslib::card()`s) are not expressible without reaching into unexported helpers.
- **Single-formula `var(shared = "...")` coordination.** When one formula declares the same `shared` key on multiple placeholders, `ptr_app()` auto-builds a combined widget at the top of the sidebar. The wiring for this lives in an internal helper; `ptr_module_server()` (the public path) aborts unless you pass a pre-built `ptr_shared_state`. The recommended path for this case is `ptr_app()` itself, or `ptr_app_grid()` with explicit `shared_ui =` builders.

Both gaps are tracked for a future API cycle.

### Background

`ptr_app_bslib()` was originally documented as a co-equal entry point alongside `ptr_app()` and `ptr_app_grid()`. Repositioning it as a demonstration wrapper makes the function's purpose honest — the value it carries is not "the official bslib variant of ggpaintr" but "a worked example of how short a wrapper can be when the public API does the heavy lifting." Users who want a `bslib` theme can still call `ptr_app_bslib()` directly; users who want a different theme, layout, or shell should write their own wrapper using the same recipe.

## Adding a new widget type

ggpaintr ships five built-in placeholder keywords (`var`, `text`, `num`, `expr`, `upload`). Three constructors let you register your own. They write to a **process-global** registry, so call each one once per R session before launching any app that uses the new keyword. Re-registering an existing keyword warns and overwrites; `ptr_clear_placeholder()` (below) removes registrations between sessions.

| Constructor | Use when the widget … |
|---|---|
| `ptr_define_placeholder_value()`    | … is not data-aware. Sliders, text, numeric, date, color pickers. |
| `ptr_define_placeholder_consumer()` | … needs upstream column names. Column pickers, multi-column selectors. |
| `ptr_define_placeholder_source()`   | … *produces* a data frame. Database queries, built-in datasets, custom readers. |

Each constructor takes a `build_ui` hook (the Shiny tag) and a resolver hook (how to turn the user's input into a substituted expression). `resolve_expr()` returning `NULL` is the *drop this argument* signal — the substitute pass strips the corresponding argument from the generated code, exactly like the built-in `var` keyword returning `NULL` for an empty pick.

The three subsections below introduce one constructor at a time with a worked example. Skim "How a placeholder runs" first — the mental model makes the hook signatures land.

### How a placeholder runs

A placeholder is *not* a widget; it is a **rule** that ggpaintr applies twice each time the user touches the app:

1. **Parse.** When `ptr_app("...formula...")` starts, ggpaintr parses the formula and, for every token whose name matches a registered keyword (`var`, `pct`, `range`, …), creates a *node*. A node remembers where the token appeared and gets a namespaced Shiny `id`. You never construct nodes by hand — the framework hands one to your hook.
2. **Render UI (your `build_ui`).** For each node, ggpaintr calls your `build_ui(node, label, …)`. Whatever Shiny tag you return is dropped into the layer's control panel. You **must** set the widget's `inputId` to `node$id` — that is the wire connecting the widget to step 3.
3. **User types or clicks.** Standard Shiny: the widget's current value is `input[[node$id]]`. The framework watches every node's input and triggers step 4 whenever something changes.
4. **Resolve to code (your `resolve_expr`).** ggpaintr calls `resolve_expr(value, node)` with the widget's current value. Your job is to return an **R expression** — not a value — that will be spliced into the rendered call at the placeholder's position. The framework then runs the substituted call to produce the plot, and prints the rendered call to the code window so the user sees a reproducible snippet.
5. **Render data (sources only — your `resolve_data`).** Source placeholders have a third hook that returns the actual data frame downstream consumers read column names from. It runs alongside `resolve_expr`.

So the two hooks do different things at different times. `build_ui` runs once per node at app start. `resolve_expr` runs every time the widget value changes. The split is what lets the same registration drive both the UI *and* the generated code.

#### What `node` carries

Inside either hook, `node` is a small list. The fields you can read:

| Field              | Meaning                                                                                                                                                |
|--------------------|--------------------------------------------------------------------------------------------------------------------------------------------------------|
| `node$id`          | Namespaced Shiny input id. Pass as `inputId` of your widget in `build_ui`. In `resolve_expr` it just identifies which node you are resolving.          |
| `node$keyword`     | The registered keyword (`"pct"`, `"range"`, …). Mostly useful when one `build_ui` is shared across keywords.                                          |
| `node$param`       | The formal argument the placeholder is bound to (e.g. `"alpha"` for `geom_point(alpha = pct)`), or `NA_character_` for unnamed positional uses.       |
| `node$companion_id`| Source placeholders only, when `companion_id_fn` is set. Namespaced id of the sibling input. See *Source placeholders* below.                          |

You should not mutate `node` — copy out the fields you need.

#### What `value` looks like

`value` is whatever Shiny stores at `input[[node$id]]` — typed exactly as the widget emits it. The common shapes:

| Widget                                    | Type of `value`                                                       |
|-------------------------------------------|-----------------------------------------------------------------------|
| `textInput`, `selectInput` (single)       | length-1 character; `""` when empty.                                  |
| `numericInput`                            | length-1 numeric; can be `NA_real_` or `NULL` before first input.     |
| `selectInput(multiple = TRUE)`, picker    | character vector; `character(0)` when nothing is selected.            |
| `sliderInput(value = c(lo, hi))`          | length-2 numeric.                                                     |
| `checkboxInput`                           | length-1 logical.                                                     |
| `fileInput`                               | one-row data frame with columns `name`, `size`, `type`, `datapath`.   |
| Custom widget                             | whatever you push via `session$sendInputMessage()`.                   |

**Always validate `value` first.** The framework calls `resolve_expr` on every change, including before the user has touched anything. Treat empty / `NULL` / wrong-length input as "no value yet" and return `NULL` (see next).

#### What `resolve_expr` must return

The framework splices your return value into the call at the placeholder's position. Allowed return types:

| Return                                  | Splices as                                          | Build with                                  |
|-----------------------------------------|-----------------------------------------------------|---------------------------------------------|
| Scalar atomic (numeric/character/…)     | A literal (`0.4`, `"red"`, `TRUE`)                  | `value`, or `as.numeric(value)`, …          |
| `name` / `symbol`                       | A bare identifier (`cyl`)                           | `rlang::sym(value)`                         |
| `call` / `language`                     | A call (`factor(cyl)`, `c(0, 100)`)                 | `rlang::expr()`, `rlang::call2()`, `!!`     |
| `NULL`                                  | **Prunes** this argument from the call entirely     | `return(NULL)`                              |

The `NULL`-prunes-the-argument rule is central. It is how the built-in `var` placeholder makes `geom_point(color = var)` render as `geom_point()` until the user actually picks a column — *not* as `geom_point(color = NULL)`. Use it generously: any time the input is empty, blank, or otherwise unusable, return `NULL` and let the framework drop the argument. For inputs that are *malformed* (a non-parseable expression, a non-numeric string in a numeric slot), throw with `rlang::abort()` instead — the runtime surfaces the message inline.

#### Where copy comes from

Your registration's `copy_defaults` becomes `defaults[[keyword]]` in the `ui_text` tree (covered in the first half of this vignette). At render time the framework merges `defaults` → `params[[param]]` → `layers[[layer]][[keyword]][[param]]` and passes the result into `build_ui`:

- The resolved `label` is passed as the `label` argument — always declare it.
- Any other leaf field you put in `copy_defaults` (`help`, `placeholder`, `empty_text`) is offered as a same-named argument if `build_ui` declares it. The convention used by the built-ins is to also accept a `copy = NULL` list and read `copy$help` etc. off it — both styles work.
- Always end the signature with `...` so future fields pass through harmlessly.

Strings in `copy_defaults` may contain `{param}`; the framework interpolates it with the surrounding formal-argument name (e.g. `"Enter a value for {param}"` becomes `"Enter a value for alpha"`).

### Value placeholders — `ptr_define_placeholder_value()`

The simplest case: a widget whose value depends on nothing outside itself. Here is a `range` keyword that renders a `sliderInput()` and substitutes a `c(lo, hi)` call into the formula.

```{r value-range}
ptr_define_placeholder_value(
  keyword       = "range",
  build_ui      = function(node, label = NULL, ...) {
    sliderInput(node$id, label = label %||% "Range",
                min = 0, max = 100, value = c(0, 100), step = 0.1)
  },
  resolve_expr  = function(value, node, ...) {
    if (is.null(value) || length(value) != 2L) return(NULL)
    rlang::expr(c(!!value[1], !!value[2]))
  },
  copy_defaults = list(label = "Range for {param}")
)
```

`node$id` is the rendered widget's Shiny input id (already namespaced for you); `label` is the resolved `ui_text` label for this slot, so any host-supplied override flows through automatically. `copy_defaults` becomes `defaults$range` in the `ui_text` tree; the `{param}` token interpolates the surrounding parameter name at render time.

Now use it. Drop a `range` token into a formula and launch:

```{r value-range-app}
ptr_app("ggplot(mtcars, aes(mpg, hp)) + geom_point() + xlim(range)")
```

The slider appears in the layer panel; dragging it re-renders the plot and updates the code window to `... + xlim(c(15, 30))`. If your hook returns `NULL` for some inputs (not the case for a slider, but the convention applies to all keywords), the framework drops the corresponding argument from the rendered call rather than splicing `NULL` literally.

### Consumer placeholders — `ptr_define_placeholder_consumer()`

A consumer is data-aware: its `build_ui` hook receives `cols` (the upstream column-name vector) and, if needed, `data` (the resolved upstream data frame — `NULL` until upstream resolves). The runtime walks the formula's pipeline and resolves the data context for you, including through `|>` chains; there is no walker to write.

```{r consumer-colvars}
ptr_define_placeholder_consumer(
  keyword       = "colvars",
  build_ui      = function(node, cols = character(), label = NULL,
                           selected = character(0), ...) {
    selectInput(node$id, label = label %||% "Columns",
                choices = cols, selected = intersect(selected, cols),
                multiple = TRUE)
  },
  resolve_expr  = function(value, node, ...) {
    if (length(value) == 0L) return(NULL)
    rlang::call2("c", !!!as.list(value))
  },
  copy_defaults = list(label = "Columns for {param}")
)
```

Consumer constructors also accept an optional `validate_input = function(value, upstream_cols)` callback. It returns `TRUE` (or `NULL`) when the input is valid, or an error-message string otherwise. The built-in `var` consumer uses it to surface "Column `foo` was removed from the upstream data" messages when the data shape changes mid-session.

Wire it into an app to see the multi-select picker react to the upstream data:

```{r consumer-colvars-app}
ptr_app("mtcars |> dplyr::select(colvars) |> 
        ggplot(aes(x = var, y = var)) + geom_point()")
```

The picker's choices are populated from `mtcars`'s column names — that is `cols` in `build_ui`. Pick two or three columns; the rendered code becomes `mtcars |> dplyr::select(c('mpg', 'hp', 'wt'))`. Clear the picker and `resolve_expr` returns `NULL`, so the `c(...)` arg is pruned rather than spliced as `NULL`.

#### Reading values, not just column names — the `data` argument

`cols` is enough when the widget only needs to know *which columns exist*. When the widget's bounds, choices, or defaults depend on the **values** themselves, declare `data` in the signature — the framework hands you the upstream data frame, refreshed whenever upstream resolves. A range slider whose `min`/`max` come from the column itself is the prototypical case:

```{r consumer-numrange}
ptr_define_placeholder_consumer(
  keyword       = "numrange",
  build_ui      = function(node, cols = character(), data = NULL,
                           label = NULL, ...) {
    num_cols <- cols[vapply(data[cols], is.numeric, logical(1))]
    col      <- num_cols[1] %||% ""
    rng      <- if (nzchar(col)) range(data[[col]], na.rm = TRUE) else c(0, 1)
    sliderInput(node$id, label = label %||% "Range",
                min = rng[1], max = rng[2], value = rng)
  },
  resolve_expr  = function(value, ...) {
    if (is.null(value)) return(NULL)
    rlang::call2("c", value[1], value[2])
  },
  copy_defaults = list(label = "Range for {param}")
)
```

```{r consumer-numrange-app}
ptr_app("
mtcars |> 
  dplyr::filter(mpg >= numrange(shared = 'n')[1],
                mpg <= numrange(shared = 'n')[2]) |> 
  ggplot(aes(x = var, y = var)) +
  geom_point()
")
```

The slider's bounds are `range(mtcars$mpg)` — not a hard-coded guess — because `data` carries the actual frame. `data` is `NULL` before upstream resolves, so guard for it (the `if (nzchar(col))` fallback above does this). Note that `numrange` splices `c(lo, hi)` and the formula reads `numrange[1]` / `numrange[2]` to pull the bounds; `%in%` would do set membership, not a range check, so don't reach for it here.

### Source placeholders — `ptr_define_placeholder_source()`

A source placeholder *produces* a data frame. Use this for widgets the rest of the formula reads from — built-in datasets, database queries, programmatic samples — without an `upload` placeholder doing the work.

```{r source-dataset}
ptr_define_placeholder_source(
  keyword       = "dataset",
  build_ui      = function(node, label = NULL, ...) {
    selectInput(node$id, label = label %||% "Built-in dataset",
                choices = c("iris", "mtcars", "diamonds", "economics"))
  },
  resolve_data  = function(value, node, ...) {
    if (is.null(value) || !nzchar(value)) return(NULL)
    switch(value,
           iris      = iris,
           mtcars    = mtcars,
           diamonds  = ggplot2::diamonds,
           economics = ggplot2::economics,
           NULL)
  },
  copy_defaults = list(label = "Built-in dataset for {param}")
)
```

`resolve_data` returns the data frame to feed downstream consumers. `resolve_expr` defaults to substituting the bare keyword (`rlang::sym(value)`) into the generated code, which is right when `value` is the name of a global object. Override `resolve_expr` to substitute something else (e.g. `expr(read.csv(!!path))`) when the generated code should re-fetch the data rather than carry a reference.

`companion_id_fn = function(id) <companion_id>` pairs the primary widget with a sibling — the built-in `upload` source uses it for the "Optional dataset name" textbox. Most sources do not need it.

Drop the `dataset` keyword in as the first stage of a pipeline:

```{r source-dataset-app}
ptr_app("ggplot(dataset, aes(var, var)) + geom_point()")
```

Pick *iris* in the dropdown — that runs `resolve_data("iris")`, which returns `iris`. The two downstream `var` consumers receive the iris columns in `cols`, populate their pickers, and the rendered code becomes `ggplot(iris, aes(Sepal.Length, Petal.Length)) + geom_point()` once the user picks columns. `resolve_expr` was not overridden, so the default `rlang::sym(value)` splices the bare symbol `iris`. If you wanted the code window to show `ggplot(read.csv("path/to/iris.csv"), …)` instead, you would override `resolve_expr` to return that `expr()` — the data the app reads from is unchanged, only the spliced text differs.

### Common pitfalls

- **Forgetting `inputId = node$id`.** Without it the widget exists but its value never reaches `resolve_expr`. Symptom: the widget renders, the user types, nothing happens. Always copy `node$id` straight into the underlying Shiny constructor.
- **Returning a `character` when you wanted a symbol.** `resolve_expr = function(value, ...) value` for a column picker splices `"cyl"` (a string literal) into the call, which then errors as `geom_point(color = "cyl")` — `"cyl"` is treated as a constant color, not a column. Wrap with `rlang::sym(value)` to splice the bare identifier.
- **Forgetting to handle empty input.** Shiny calls your hook before the user has touched anything. Treat `NULL`, `character(0)`, `NA`, and empty strings as "no value yet" and `return(NULL)` — the framework will prune the argument, and the layer continues to work.
- **Confusing `resolve_expr` with `resolve_data`.** For source placeholders the two run together but mean different things: `resolve_data` produces the **actual data frame** downstream consumers read; `resolve_expr` produces the **text that goes in the rendered call**. They can disagree on purpose (e.g. show `read.csv("…")` in the code while reading uploaded bytes at runtime).
- **Dropping `...` from the signature.** The framework may add new named arguments to its `build_ui` invocation in future versions; without `...` the call fails. Always end every hook signature with `...`.
- **Re-registering inside a reactive.** The registry is process-global; calling `ptr_define_placeholder_*()` repeatedly inside a Shiny observer wastes work and emits an "Overwriting placeholder registry entry" notice each time. Register once at script top level, before `ptr_app()`.
- **Returning the data instead of an expression from `resolve_expr`.** A common confusion for source placeholders: `resolve_expr` does *not* return the data frame — it returns the snippet that *names* the data in the rendered code. `resolve_data` returns the actual data.

## Unregistering

Custom placeholders live in a process-global registry. Call `ptr_clear_placeholder()` between sessions, inside tests, or when re-running this vignette interactively, to wipe user-registered keywords without restarting R:

```{r unregister, eval = FALSE}
ptr_clear_placeholder("range")  # remove one keyword
ptr_clear_placeholder()         # remove every user-registered keyword
```

The five built-ins (`var`, `text`, `num`, `expr`, `upload`) are protected — passing one to `ptr_clear_placeholder()` errors. Re-registering an existing keyword via any of the three constructors above overwrites the previous entry (with a `cli` notice); you do not need to clear first.
