---
title: "ggpaintr in the Wild"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ggpaintr in the Wild}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment   = "#>",
  warning   = FALSE,
  message   = FALSE,
  eval      = interactive()
)
```

```{r clipboard, echo = FALSE, eval = TRUE}
xaringanExtra::use_clipboard()
```

## 1. Setup

Every chunk in this vignette is marked `eval = interactive()` — they will not run during vignette build, but each block is runnable as-is at the R prompt after:

```{r libs, eval = TRUE}
library(ggpaintr)   # also attaches ggplot2
library(shiny)
library(rlang)
```

Some sections also load `dplyr`, `tidyr`, `purrr`, `broom`, `slider`, `palmerpenguins`, `plotly`, `ggiraph`, `colourpicker`, or one of the ggplot2 extension packages (`ggpcp`, `ggridges`, `ggrepel`, `ggalluvial`, `ggdist`). Each section names what it needs.

## 2. What this vignette is

The other vignettes teach the workflow piece by piece. This one shows what the pieces add up to. Every section pairs a complete `ggplot2` (or extension-package) graphic with the `ggpaintr` formula that turns it into an interactive Shiny app. Read the original to know what is being built; read the formula to see which arguments become widgets.

Copy a chunk into a session, install the extension package named at the top of the section if you do not have it, and run it. The examples assume the setup chunks above have already executed.

Several gallery sections below reference custom placeholders — `range` (a slider returning a length-2 numeric) and `colvars` (a multi-column picker) — alongside the five built-ins (`var`, `text`, `num`, `expr`, `upload`). The three constructors that register such widgets (`ptr_define_placeholder_value()`, `ptr_define_placeholder_consumer()`, `ptr_define_placeholder_source()`), the full hook contract, and a worked example for each are covered in `vignette("ggpaintr-customization")`. Run the setup chunk below once per R session before launching any gallery example that uses these keywords.

```{r custom-placeholders-setup}
ptr_define_placeholder_value(
  keyword = "range",
  build_ui = function(node, label = NULL, ...) {
    shiny::sliderInput(node$id, label = label %||% "Range",
                       min = -100, max = 100, value = c(0, 50), step = 0.1)
  },
  resolve_expr = function(value, node, ...) {
    if (is.null(value) || length(value) != 2L) return(NULL)
    rlang::expr(c(!!value[1], !!value[2]))
  },
  copy_defaults = list(label = "Range for {param}")
)

ptr_define_placeholder_consumer(
  keyword = "colvars",
  build_ui = function(node, cols = character(), label = NULL,
                      selected = character(0), ...) {
    shiny::selectInput(node$id, label = label %||% "Columns",
                       choices = cols, selected = intersect(selected, cols),
                       multiple = TRUE)
  },
  resolve_expr = function(value, node, ...) {
    if (length(value) == 0L) return(NULL)
    rlang::call2("c", !!!as.list(value))
  },
  copy_defaults = list(label = "Columns for {param}")
)
```

## 3. A realistic ggplot2 graphic, parameterized

A typical exploratory plot stacks several layers and reads several hard-coded constants. Here is a multi-layer fuel-economy chart from the `mpg` dataset:

```{r mpg-original}
ggplot(mpg, aes(displ, hwy, color = class)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_smooth(method = "loess", se = FALSE, span = 0.75) +
  facet_wrap(~ drv) +
  scale_color_brewer(palette = "Set2") +
  labs(title = "Highway MPG vs engine displacement",
       x = "Engine displacement (L)", y = "Highway MPG") +
  coord_cartesian(xlim = c(1, 7), ylim = c(10, 45))
```

Replace each tuning knob with a placeholder token. `var` becomes a column picker populated from the data; `num` becomes `numericInput`s; `text` becomes `textInput`s; the custom `range` becomes a slider:

```{r mpg-paintr}
ptr_app(
  "ggplot(data = mpg, aes(x = var, y = var, color = var)) +
     geom_point(alpha = num, size = num) +
     geom_smooth(method = 'loess', se = FALSE, span = num) +
     facet_wrap(~ var) +
     scale_color_brewer(palette = 'Set2') +
     labs(title = text, x = text, y = text) +
     coord_cartesian(xlim = range, ylim = range)"
)
```

Every aesthetic, alpha, size, span, and the two coordinate ranges is now a widget. Rebuilding the original graphic from the running app is a matter of picking those values; the code panel under the plot lifts the selections back into a static script.

## 4. Data-cleaning pipelines into `ggplot()`

`ggpaintr` parses any expression that evaluates to a `ggplot` object, including the common "clean, then plot" pipeline where the data is piped into `ggplot()`. Pipeline stages with placeholders get a per-layer **Data** sub-tab in the sidebar; column pickers further down the pipeline lazily resolve their upstream against the current Data-tab inputs, so picking a column in stage one immediately refreshes the choices for stage two without requiring an explicit "apply" click.

The six sub-sections below scale from a one-line filter to full broom-driven analyses. Each pairs the static `ggplot2` graphic an analyst would actually write with the `ptr_app()` formula that exposes its knobs.

### 4.1 Filter, group, and plot

```{r pipe-original}
library(dplyr)

mpg |>
  dplyr::filter(displ > 1.5) |>
  dplyr::group_by(class) |>
  dplyr::filter(dplyr::n() > 5) |>
  dplyr::ungroup() |>
  ggplot(aes(displ, hwy, color = class)) +
  geom_point(alpha = 0.5)
```

The same pipeline parameterized — the two `filter` thresholds and the point alpha become widgets:

```{r pipe-paintr}
library(dplyr)

ptr_app(
  "mpg |>
     dplyr::filter(displ > num) |>
     dplyr::group_by(class) |>
     dplyr::filter(dplyr::n() > num) |>
     dplyr::ungroup() |>
     ggplot(aes(var, var, color = class)) +
     geom_point(alpha = num) +
     labs(title = text)"
)
```

Each pipeline stage gets its own enable/disable checkbox (G11). Toggling a stage off rewrites the generated code as if that step were never there; toggle it back on and the threshold widget resurfaces. Note also that `var` (and any custom consumer like `colvars`) introspects the upstream data automatically — pipe-walking through `|>` chains is part of the runtime, not your placeholder code.

### 4.2 PCA scores with 95% confidence ellipses

A classic exploratory move — project numeric columns onto their first two principal components, colour by a grouping factor, and overlay confidence ellipses. The `do_pca()` helper keeps the lambda out of the pipe (anonymous-function pipe stages are unsupported by the runtime — see `vignette("ggpaintr-use-cases")`).

```{r pca-original}
library(dplyr)
library(broom)

do_pca <- function(d, cols) {
  broom::augment(prcomp(d[, cols], scale. = TRUE), d)
}

iris |>
  do_pca(c("Sepal.Length", "Sepal.Width", "Petal.Length", "Petal.Width")) |>
  ggplot(aes(.fittedPC1, .fittedPC2, color = Species)) +
    stat_ellipse(level = 0.95, linewidth = 0.8) +
    geom_point(alpha = 0.8, size = 2) +
    scale_color_brewer(palette = "Dark2") +
    labs(title = "Iris in PC space, with 95% ellipses",
         x = "PC1", y = "PC2") +
    theme_minimal() +
    theme(legend.position = "bottom")
```

The ggpaintr version introduces a small custom **palette** placeholder so the ColorBrewer palette becomes a dropdown, lets the user pick which numeric columns enter PCA via `colvars`, and re-uses `num(shared = "n")` across the ellipse `level` and the title — moving one slider redraws both.

```{r pca-paintr}
library(dplyr)
library(broom)

ptr_define_placeholder_value(
  keyword       = "palette",
  build_ui      = function(node, label = NULL, ...) {
    shiny::selectInput(
      node$id, label = label %||% "ColorBrewer palette",
      choices = c("Dark2", "Set1", "Set2", "Set3", "Paired", "Accent"),
      selected = "Dark2"
    )
  },
  resolve_expr  = function(value, node, ...) {
    if (is.null(value) || identical(value, "")) return(NULL)
    as.character(value)
  },
  copy_defaults = list(label = "Palette for {param}")
)

do_pca <- function(d, cols) {
  broom::augment(prcomp(d[, cols], scale. = TRUE), d)
}

ptr_app(
  'iris |>
     do_pca(colvars) |>
     ggplot(aes(.fittedPC1, .fittedPC2, color = Species)) +
       stat_ellipse(level = num(shared = "n"), linewidth = num) +
       geom_point(alpha = num, size = num) +
       scale_color_brewer(palette = palette) +
       labs(title = paste0("Iris in PC space, with ",
                           as.character(100 * num(shared = "n")),
                           "% ellipses"),
            x = text, y = text) +
       theme_minimal() +
       theme(legend.position = "bottom")'
)
```

### 4.3 K-means elbow plot

Tibble of `k`, map `kmeans()` over it, glance for `tot.withinss`, plot. `expr(shared = "k")` is the single source of truth for both the cluster range and the x-axis breaks — type `1:9` or `c(2, 3, 5, 8)` and the chart redraws end-to-end.

```{r kmeans-original}
library(dplyr)
library(tidyr)
library(purrr)
library(broom)

tibble(k = 1:9) |>
  mutate(
    kmeans  = map(k, function(.k) kmeans(
      scale(iris[, c("Sepal.Length", "Sepal.Width",
                     "Petal.Length", "Petal.Width")]),
      centers = .k, nstart = 25)),
    glanced = map(kmeans, broom::glance)
  ) |>
  unnest(glanced) |>
  ggplot(aes(k, tot.withinss)) +
    geom_line(color = "gray50") +
    geom_point(size = 3, color = "#0072B2") +
    scale_x_continuous(breaks = 1:9) +
    labs(title = "Elbow plot for k-means on iris",
         x = "Number of clusters", y = "Total within-cluster SS") +
    theme_minimal()
```

The clustered columns are fixed here rather than a `colvars` picker: `colvars` is a *consumer* placeholder and resolves its choices against the pipeline's upstream data, which at this point is `tibble(k = ...)` — not `iris`. The parameterized knobs are the cluster range (`expr(shared = "k")`, reused for the x-axis breaks) and the point size.

```{r kmeans-paintr}
library(dplyr)
library(tidyr)
library(purrr)
library(broom)

ptr_app(
  'tibble(k = expr(shared = "k")) |>
     mutate(
       kmeans  = map(k, function(.k) kmeans(
         scale(iris[, c("Sepal.Length", "Sepal.Width",
                        "Petal.Length", "Petal.Width")]),
         centers = .k, nstart = 25)),
       glanced = map(kmeans, broom::glance)
     ) |>
     unnest(glanced) |>
     ggplot(aes(k, tot.withinss)) +
       geom_line(color = "gray50") +
       geom_point(size = num, color = "#0072B2") +
       scale_x_continuous(breaks = expr(shared = "k")) +
       labs(title = text, x = text, y = text) +
       theme_minimal()'
)
```

### 4.4 Regression diagnostics

`broom::augment()` turns an `lm` fit into a tidy data frame whose `.fitted`, `.resid`, `.hat`, `.std.resid`, and `.cooksd` columns are the staples of regression diagnostics. Make the model formula an `expr` widget and the diagnostic axes a pair of `var` pickers — one app, every standard diagnostic plot (residuals vs fitted, scale-location, leverage).

```{r regress-original}
library(broom)

broom::augment(lm(mpg ~ wt + cyl + hp, data = mtcars)) |>
  ggplot(aes(.fitted, .resid)) +
    geom_hline(yintercept = 0, color = "gray60", linetype = 2) +
    geom_point(alpha = 0.7, size = 2.5, color = "#D55E00") +
    geom_smooth(method = "loess", se = FALSE, span = 0.75) +
    labs(title = "Residuals vs fitted (mpg ~ wt + cyl + hp)",
         x = "Fitted", y = "Residual") +
    theme_minimal()
```

```{r regress-paintr}
library(broom)

ptr_app(
  'broom::augment(lm(expr, data = mtcars)) |>
     ggplot(aes(x = var, y = var)) +
       geom_hline(yintercept = 0, color = "gray60", linetype = 2) +
       geom_point(alpha = num, size = num, color = "#D55E00") +
       geom_smooth(method = "loess", se = FALSE, span = num) +
       labs(title = text, x = text, y = text) +
       theme_minimal()'
)
```

The `expr` widget accepts the full formula `mpg ~ wt + cyl + hp`; the `var` pickers downstream populate from the augmented frame so `.fitted`, `.resid`, `.hat`, `.std.resid`, `.cooksd` all appear as choices.

### 4.5 Rolling time-series

`economics` (shipped with `ggplot2`) holds 40+ years of monthly US macro indicators. The standard recipe — pick a series, overlay an *n*-month rolling mean — turns into two widgets: `var(shared = "y")` chooses the series (used both raw and in the rolling computation) and `num` picks the window size.

```{r rolling-original}
library(dplyr)

roll_mean <- function(x, n) {
  as.numeric(stats::filter(x, rep(1 / n, n), sides = 1))
}

economics |>
  mutate(roll = roll_mean(unemploy, 12)) |>
  ggplot(aes(date, unemploy)) +
    geom_line(color = "gray70") +
    geom_line(aes(y = roll), color = "#0072B2", linewidth = 1) +
    labs(title = "US unemployment with 12-month rolling mean",
         x = NULL, y = "Persons (thousands)") +
    theme_minimal()
```

```{r rolling-paintr}
library(dplyr)

roll_mean <- function(x, n) {
  as.numeric(stats::filter(x, rep(1 / n, n), sides = 1))
}

ptr_app(
  'economics |>
     mutate(roll = roll_mean(var(shared = "y"), num)) |>
     ggplot(aes(date, var(shared = "y"))) +
       geom_line(color = "gray70") +
       geom_line(aes(y = roll), color = "#0072B2", linewidth = num) +
       labs(title = text, x = text, y = text) +
       theme_minimal()'
)
```

`var(shared = "y")` is the canonical pattern when the same column drives both the data step *and* the plot aesthetic — without it, you would have two pickers the user has to keep in sync by hand.

### 4.6 Group-wise regression coefficients

A small-multiples regression: nest by a grouping factor, fit one `lm` per group, tidy the coefficients with confidence intervals, plot as a forest. The model formula stays as `expr` so a reader can swap responses or predictors at will.

```{r coefs-original}
library(palmerpenguins)
library(dplyr)
library(tidyr)
library(purrr)
library(broom)

penguins |>
  tidyr::drop_na(bill_depth_mm, bill_length_mm) |>
  group_by(species) |>
  tidyr::nest() |>
  mutate(
    fit  = map(data, function(.d) lm(bill_depth_mm ~ bill_length_mm, data = .d)),
    tidy = map(fit,  function(.f) broom::tidy(.f, conf.int = TRUE))
  ) |>
  unnest(tidy) |>
  filter(term == "bill_length_mm") |>
  ggplot(aes(estimate, species, color = species)) +
    geom_vline(xintercept = 0, color = "gray60", linetype = 2) +
    geom_pointrange(aes(xmin = conf.low, xmax = conf.high), size = 0.8) +
    scale_color_brewer(palette = "Dark2") +
    labs(title = "Slope of bill depth ~ bill length, by species",
         x = "Estimate (95% CI)", y = NULL) +
    theme_minimal() +
    theme(legend.position = "none")
```

```{r coefs-paintr}
library(palmerpenguins)
library(dplyr)
library(tidyr)
library(purrr)
library(broom)

ptr_app(
  'penguins |>
     tidyr::drop_na(bill_depth_mm, bill_length_mm) |>
     group_by(species) |>
     tidyr::nest() |>
     mutate(
       fit  = map(data, function(.d) lm(expr, data = .d)),
       tidy = map(fit,  function(.f) broom::tidy(.f, conf.int = TRUE))
     ) |>
     unnest(tidy) |>
     filter(term == text) |>
     ggplot(aes(estimate, species, color = species)) +
       geom_vline(xintercept = 0, color = "gray60", linetype = 2) +
       geom_pointrange(aes(xmin = conf.low, xmax = conf.high),
                       size = num) +
       scale_color_brewer(palette = "Dark2") +
       labs(title = text, x = text, y = NULL) +
       theme_minimal() +
       theme(legend.position = "none")'
)
```

`filter(term == text)` is how you let the user pick which coefficient to display — type `bill_length_mm` for slopes, `(Intercept)` for intercepts.

## 5. Interactive output packages

`ggpaintr` does not lock you into `renderPlot`. The runtime's `state$runtime()` reactive returns a list whose `$plot` slot is the fitted `ggplot` object, suitable for any backend that accepts a `ggplot` — `plotly` and `ggiraph` are the two most common. `ptr_extract_plot(state)` is the same accessor in non-reactive contexts.

The recipe is the same in both subsections: lay out `ptr_ui(formula, "id")` for the sidebar + default plot panel, add a second column with the alternative output, and in the server use `state <- ptr_server(formula, "id")` to wire everything up, then bind your custom output to `state$runtime()`.

### 5.1 plotly tooltips

Original — a hover-tooltipped scatter:

```{r plotly-original}
library(plotly)

p <- ggplot(mpg, aes(displ, hwy, color = class,
                     text = paste(manufacturer, model, sep = " "))) +
       geom_point(size = 3, alpha = 0.7) +
       coord_cartesian(xlim = c(1, 7), ylim = c(10, 45))

plotly::ggplotly(p, tooltip = "text")
```

ggpaintr version — embed via the module helpers and own the plotly output:

```{r plotly-paintr}
library(plotly)

formula <- "ggplot(data = mpg,
                   aes(x = var, y = var, color = var,
                       text = paste(manufacturer, model, sep = ' '))) +
              geom_point(size = num, alpha = num) +
              coord_cartesian(xlim = range, ylim = range)"

ui <- fluidPage(
  fluidRow(
    column(5, ptr_ui(formula, "plotly_demo")),
    column(7, plotly::plotlyOutput("interactive_plot", height = "500px"))
  )
)

server <- function(input, output, session) {
  state <- ptr_server(formula, "plotly_demo")
  output$interactive_plot <- plotly::renderPlotly({
    res <- state$runtime()
    shiny::req(isTRUE(res$ok), res$plot)
    plotly::ggplotly(res$plot, tooltip = "text")
  })
}

shinyApp(ui, server)
```

Reading `state$runtime()` outside `isolate()` is what wires the plotly re-render to Update Plot clicks; `req(isTRUE(res$ok), res$plot)` keeps the panel quiet between draws and on transient error states.

### 5.2 ggiraph tooltips and click handling

Original — interactive points with a colour-picked highlight:

```{r ggiraph-original}
library(ggiraph)

ggplot(mpg, aes(displ, hwy)) +
  geom_point_interactive(aes(tooltip = model), size = 3, color = "#e63946")
```

ggpaintr version with a colour-picker placeholder so the user can recolour the layer:

```{r ggiraph-paintr}
library(ggiraph)
library(colourpicker)

ptr_define_placeholder_value(
  keyword       = "color",
  build_ui      = function(node, label = NULL, ...) {
    colourpicker::colourInput(node$id, label = label %||% "Highlight colour",
                              value = "#e63946")
  },
  resolve_expr  = function(value, node, ...) {
    if (is.null(value) || identical(value, "")) return(NULL)
    as.character(value)
  },
  copy_defaults = list(label = "Highlight colour for {param}")
)

formula <- "ggplot(data = mpg, aes(x = var, y = var)) +
              geom_point_interactive(aes(tooltip = var),
                                     size = num, color = color)"

ui <- fluidPage(
  fluidRow(
    column(5, ptr_ui(formula, "ggiraph_demo")),
    column(7, ggiraph::girafeOutput("interactive_plot", height = "500px"))
  )
)

server <- function(input, output, session) {
  state <- ptr_server(formula, "ggiraph_demo")
  output$interactive_plot <- ggiraph::renderGirafe({
    res <- state$runtime()
    shiny::req(isTRUE(res$ok), res$plot)
    ggiraph::girafe(code = print(res$plot))
  })
}

shinyApp(ui, server)
```

## 6. ggplot2 extension packages

`ggpaintr` does not know about extension geoms — it sees them as ordinary layer calls and parameterizes their arguments the same way it parameterizes core geoms. The five examples below cover a parallel-coordinates plot, ridge densities, repelled labels, alluvial flows, and `ggdist` distribution slabs.

### 6.1 ggpcp — parallel coordinates

Original (uses the `flea` dataset shipped by `GGally`):

```{r pcp-original}
library(ggpcp)
data(flea, package = "GGally")

flea |>
  pcp_select(species, tars1, tars2, head, aede1, aede2, aede3) |>
  pcp_scale(method = "uniminmax") |>
  pcp_arrange() |>
  ggplot(aes_pcp()) +
    geom_pcp_axes() +
    geom_pcp(aes(colour = species))
```

ggpaintr version — let the user pick which columns to put on the parallel axes via the `colvars` placeholder:

```{r pcp-paintr}
library(ggpcp)
data(flea, package = "GGally")

ptr_app(
  "ggplot(data = flea |>
                  pcp_select(colvars) |>
                  pcp_scale(method = 'uniminmax') |>
                  pcp_arrange(),
          mapping = aes_pcp()) +
     geom_pcp_axes() +
     geom_pcp(aes(colour = var))"
)
```

### 6.2 ggridges — ridge densities

Original (uses `lincoln_weather` shipped by `ggridges`):

```{r ridges-original}
library(ggridges)

ggplot(lincoln_weather,
       aes(x = `Mean Temperature [F]`, y = Month, fill = ..x..)) +
  geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01,
                               bandwidth = 3) +
  scale_fill_viridis_c(option = "C")
```

ggpaintr version — bandwidth, scale, and the colour aesthetic become widgets:

`lincoln_weather` ships with non-syntactic column names (`Mean Temperature [F]`, …). The `var` placeholder needs syntactic identifiers, so we normalize first; see `vignette("ggpaintr-use-cases")` §"Local data with non-syntactic columns".

```{r ridges-paintr}
library(ggridges)

lincoln_weather_clean <- ptr_normalize_column_names(lincoln_weather)

ptr_app(
  "ggplot(data = lincoln_weather_clean, aes(x = var, y = var, fill = ..x..)) +
     geom_density_ridges_gradient(scale = num,
                                  rel_min_height = num,
                                  bandwidth = num) +
     scale_fill_viridis_c(option = 'C') +
     labs(title = text)"
)
```

### 6.3 ggrepel — labelled scatter

Original — label cars by row name, repelling overlaps:

```{r repel-original}
library(ggrepel)

mt <- data.frame(mtcars, model = rownames(mtcars))

ggplot(mt, aes(wt, mpg, label = model)) +
  geom_point(size = 3, color = "steelblue") +
  geom_text_repel(size = 3.5, max.overlaps = 12, box.padding = 0.3)
```

ggpaintr version — `max.overlaps`, `size`, and `box.padding` all turn into numeric inputs. The label aesthetic is bound by `var`:

```{r repel-paintr}
library(ggrepel)

mt <- data.frame(mtcars, model = rownames(mtcars))

ptr_app(
  "ggplot(data = mt, aes(x = var, y = var, label = var)) +
     geom_point(size = num, color = 'steelblue') +
     geom_text_repel(size = num,
                     max.overlaps = num,
                     box.padding = num) +
     labs(title = text)",
  envir = environment()
)
```

`envir = environment()` is the seam that lets the runtime see the local `mt` data frame. `ptr_app()` defaults to `parent.frame()`, which would pick up `mt` here automatically; passing `environment()` is explicit.

### 6.4 ggalluvial — flow diagrams

Original (uses the `vaccinations` dataset shipped by `ggalluvial`):

```{r alluvial-original}
library(ggalluvial)

ggplot(vaccinations,
       aes(x = survey, stratum = response, alluvium = subject,
           y = freq, fill = response)) +
  geom_flow(alpha = 0.7) +
  geom_stratum() +
  scale_x_discrete(expand = c(0.1, 0.1))
```

ggpaintr version — alpha and the fill aesthetic become widgets:

```{r alluvial-paintr}
library(ggalluvial)

ptr_app(
  "ggplot(data = vaccinations,
          aes(x = survey, stratum = response, alluvium = subject,
              y = freq, fill = var)) +
     geom_flow(alpha = num) +
     geom_stratum() +
     scale_x_discrete(expand = c(0.1, 0.1)) +
     labs(title = text)"
)
```

### 6.5 ggdist — distribution visualizations

Original — half-eye distributions per cylinder count:

```{r dist-original}
library(ggdist)

ggplot(mtcars, aes(x = factor(cyl), y = mpg)) +
  stat_halfeye(adjust = 0.5, justification = -0.2,
               .width = 0, point_colour = NA, slab_alpha = 0.6) +
  geom_boxplot(width = 0.15, outlier.shape = NA) +
  coord_flip()
```

ggpaintr version — every visual constant of `stat_halfeye()` becomes a slider or numeric input:

```{r dist-paintr}
library(ggdist)

ptr_app(
  "ggplot(data = mtcars, aes(x = factor(cyl), y = var)) +
     stat_halfeye(adjust = num,
                  justification = num,
                  .width = num,
                  point_colour = NA,
                  slab_alpha = num) +
     geom_boxplot(width = num, outlier.shape = NA) +
     coord_flip() +
     labs(title = text)"
)
```

## 7. Where to go next

This gallery is the recipe book. The other vignettes answer different user questions:

- **`vignette("ggpaintr-use-cases")`** — the three-level ladder. L1 all-in-one (`ptr_app()` / `ptr_app_grid()`); L2 embed in your own Shiny app keeping the default layout (the self-contained pair `ptr_ui()` / `ptr_server()`, plus the shared trio `ptr_shared()` / `ptr_shared_panel()` / `ptr_shared_server()` for two or more linked plots); L3 own every piece of the UI — one bare exported function per pane (`ptr_ui_header()` / `ptr_ui_controls()` / `ptr_ui_plot()` / `ptr_ui_error()` / `ptr_ui_code()` / `ptr_ui_shared_panel()`), the two combinators (`ptr_ui_inline_error()` / `ptr_ui_toggle_code()`), and the optional `ptr_ui_page()` shell — including custom rendering off the `state` that the single `ptr_server()` returns (read `state$runtime()` inside a render block, or the `ptr_extract_plot()` / `_code()` / `_error()` accessors plus `ptr_gg_extra()` outside reactive contexts). The plotly / ggiraph examples in §5 above are the canonical worked examples for the L3 custom-renderer pattern — use-cases cross-links back here. For a custom page shell or theme, see `vignette("ggpaintr-customization")` § "Writing your own wrapper".
- **`vignette("ggpaintr-customization")`** — `ui_text` overrides, the three custom-placeholder constructors (full hook contract and worked examples — the `range` / `colvars` constructors used throughout this gallery live there), and the shared-placeholder pattern for one widget driving multiple plots.
- **`vignette("ggpaintr-safety")`** — `expr_check`, the denylist + AST-walker safety model, upload trust boundary, public-deployment checklist.
- **`vignette("ggpaintr-llm")`** — wiring ggpaintr up as an LLM-callable tool with ellmer.

For canonical API reference, follow the `?fn` links:

- `?ptr_app`, `?ptr_ui`, `?ptr_server`, `?ptr_app_grid` — entry points.
- `?ptr_define_placeholder_value`, `?ptr_define_placeholder_consumer`, `?ptr_define_placeholder_source` — the three constructors that extend the registry.
- `?ptr_extract_plot`, `?ptr_extract_code`, `?ptr_extract_error`, `?ptr_gg_extra` — the embed-time accessors used by the plotly / ggiraph sections.
