---
title: "Use cases"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Use cases}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment   = "#>",
  warning   = FALSE,
  message   = FALSE,
  eval      = interactive()
)
```

This vignette answers the question *how do I use ggpaintr for my situation?* Companion vignettes cover tailoring defaults (`vignette("ggpaintr-customization")`), sharing safely (`vignette("ggpaintr-safety")`), real ggplot recipes (`vignette("ggpaintr-gallery")`), and LLM integration (`vignette("ggpaintr-llm")`).

Every chunk is marked `eval = interactive()` and is runnable as-is at the R prompt after:

```{r libs}
library(ggpaintr)
```

## Formula tour

A ggpaintr formula is a single `ggplot()` call written as text. Drop one of five placeholder keywords anywhere a value would normally go, and the runtime substitutes the user's input back into the expression at render time. The five keywords are:

| Keyword  | Becomes                                       | Example formula fragment        | Renders as                                       |
|----------|-----------------------------------------------|---------------------------------|--------------------------------------------------|
| `var`    | column picker (data-aware `selectInput`)      | `aes(x = var, y = var)`         | two dropdowns scoped to the upstream data's columns |
| `text`   | free-text input                               | `labs(title = text)`            | one text field                                   |
| `num`    | numeric input                                 | `geom_point(size = num)`        | one numeric input                                |
| `expr`   | code editor (validated against a denylist)    | `facet_wrap(expr)`              | one expression box; see `vignette("ggpaintr-safety")` for the validation model |
| `upload` | file picker (returns a data frame)            | `ggplot(upload, …)`             | one file picker + an optional dataset-name field |

A formula that exercises four of the five:

```{r formula-tour}
ptr_app("
ggplot(iris, aes(x = var, y = var, color = var)) +
  geom_point(size = num) +
  labs(title = text) +
  facet_wrap(expr)
")
```

The full grammar reference — including positional-argument behavior, `shared = "<id>"` annotation, and the empty-call cleanup rule — lives at `?ptr_app`.

## Data pipelines in the formula

A layer's data argument is allowed to be a pipeline. Both `|>` and `%>%` are accepted, and the two can be mixed in the same chain. Each pipeline stage that contains a placeholder gets its own row of widgets under the layer's **Data** subtab, plus a per-stage enable/disable toggle that rewrites the generated code as if that step were never there. Column pickers downstream of a stage lazily re-resolve their column choices against the current Data-subtab inputs, so changing a filter threshold immediately refreshes the picker below it.

```{r pipeline-formula-tour}
ptr_app(
  "mpg |>
     dplyr::filter(displ > num) |>
     dplyr::group_by(class) |>
     dplyr::filter(dplyr::n() > num) |>
     dplyr::ungroup() |>
     ggplot(aes(var, var, color = class)) +
     geom_point(alpha = num)"
)
```

See `vignette(\"ggpaintr-gallery\")` § 4 for the same example with the original `dplyr` pipeline alongside the parameterized version.

### What we expect to work

- **Both pipe operators**, possibly mixed in the same chain: `df %>% filter(...) |> ggplot(...)`.
- **Bare-symbol stage heads**: `filter(...)`, `head(...)`, `transform(...)`.
- **Namespaced stage heads**: `dplyr::filter(...)`, `tidyr:::pivot_longer(...)`.
- **Parenthesised heads**: `(filter)(...)` — used occasionally to disambiguate function lookup.
- **`upload` as the pipeline head**: `upload |> dplyr::filter(...) |> ggplot(...)`. The upload widget feeds the pipeline; every downstream stage and placeholder resolves against the uploaded data.
- **Per-layer pipelines on multi-layer formulas**: each layer can have its own data-arg pipeline (`ggplot(data = df1 |> filter(...)) + geom_smooth(data = df2 |> filter(...))`) — the Data subtabs are scoped per layer.
- **Placeholders in any stage**: `var`, `text`, `num`, `expr` inside any stage's call resolve as if that stage were a regular layer call.

### What we know does **not** work yet

- **Anonymous functions as pipe stages.** `df |> (\(x) x[x$cyl > 4, ])() |> ggplot(...)` and `df |> (function(x) ...)() |> ggplot(...)` are not supported. The runtime resolves each stage's name from its call head; an inline lambda has no name, so the stage cannot be assigned a stable Shiny id or a layer label, and the Data subtab cannot render. The expression-level safety walker still inspects the lambda body, but the surrounding pipeline machinery does not. Workaround: lift the anonymous function to a named helper in the calling environment, then use it as a bare-symbol stage (`df |> my_helper() |> ggplot(...)`).

## L1 — All-in-one entry point

L1 is the dominant use case: hand ggpaintr a formula, get back a running Shiny app. Two entry points cover the common layouts.

| Entry point        | Use when                                                                 |
|--------------------|--------------------------------------------------------------------------|
| `ptr_app()`        | Default layout. The built-in chrome (title, layer picker, plot pane, code pane) is enough. |
| `ptr_app_grid()`   | Multiple formulas, one app — each plot gets its own tile, optionally with shared widgets at the top. |

For a different page shell or theme, write a thin wrapper on top of the public ggpaintr primitives — see `vignette("ggpaintr-customization")` § "Writing your own wrapper" for the recipe and a worked example.

### Single plot — `ptr_app()`

```{r app-basic}
ptr_app(
  "ggplot(iris, aes(var, var, color = var)) + geom_point() + labs(title = text)"
)
```

### Multiple plots and shared widgets — `ptr_app_grid()`

`ptr_app_grid()` is the turn-key entry point for multi-plot apps with shared widgets. It takes a list of formulas in `plots =` and renders one tile per formula. Annotate any placeholder occurrence with `keyword(shared = "<id>")`; pass `shared_ui = list(<id> = function(id) widget_tag)` so the framework knows what tag to render at the top of the page. The `<id>` namespace is shared across every formula in `plots`, so editing the top widget propagates to every plot. (Internally `ptr_app_grid()` now builds the same shared coordinator the L2 trio exposes — see *Multiple linked instances* below — but as an L1 entry point it hides that machinery behind `plots =` / `shared_ui =`.)

```{r app-grid-shared}
ptr_app_grid(
  plots = list(
    "ggplot(iris, aes(x = var, y = Sepal.Length, fill = Species)) +
       geom_boxplot()",
    "ggplot(iris, aes(x = var, y = Sepal.Width, fill = Species)) +
       geom_violin()"
  ),
  shared_ui = list()  # no shared= annotations above; the first var stays per-tile
)
```

Adding `shared = 'metric'` to both `var` occurrences and supplying `shared_ui = list(metric = function(id) selectInput(id, "Metric", names(iris)))` lifts the column choice into a single widget that drives both plots when the user clicks **Draw all**. The shared mechanism works for any keyword — built-in `num(shared = "sz")` paired with `shared_ui = list(sz = function(id) sliderInput(id, "Size", 1, 10, 3))` is the simplest two-line example; custom consumers work the same way (see `vignette("ggpaintr-customization")` for registering custom widget types).

```{r app-grid-shared-added}
ptr_app_grid(
  plots = list(
    "ggplot(iris, aes(x = var(shared = 'metric'), y = Sepal.Length, fill = Species)) +
       geom_boxplot()",
    "ggplot(iris, aes(x = var(shared = 'metric'), y = Sepal.Width, fill = Species)) +
       geom_violin()"
  ),
  shared_ui = list(metric = function(id) selectInput(id, "Metric", names(iris)))
)
```

### Empty-call cleanup

A placeholder that resolves to "missing" (an empty `var` pick, a blank `text`, a cleared `num`) drops its argument from the generated code. When that leaves an empty call whose name is in ggpaintr's curated cleanup list, the whole call disappears too:

```r
ptr_app("
ggplot(iris, aes(x = var, y = var)) +
  geom_point() +
  labs(title = text) +              # title empty → labs() empty → labs() dropped
  facet_wrap(expr)                  # expr empty → facet_wrap() empty → facet_wrap() dropped
")
```

The cleanup list covers safe ggplot2 no-ops: `theme()`, `labs()`, `xlab` / `ylab` / `ggtitle`, `facet_wrap` / `facet_grid` / `facet_null`, `xlim` / `ylim` / `lims`, `expand_limits`, `guides`, `annotate`, and the empty mapping helper (`aes()`). Third-party helpers (e.g. `pcp_theme()` from `ggpcp`) are *not* in the cleanup list — being absent is the "removal safety unknown" signal. Pass `safe_to_remove = c("pcp_theme")` to opt a specific name into the cleanup pass. An `expr` placeholder whose user supplies an expression always wins: whatever was typed is honoured verbatim, even if its top-level name is in `safe_to_remove`.

### Local data with non-syntactic columns

`var` placeholders need syntactic, unique column names — `Sepal.Length` is fine, `"Sepal Length"` is not, and `iris` deliberately ships clean. If your local data has spaces, reserved words, or duplicates in its names, pipe it through `ptr_normalize_column_names()` before referencing it from a formula:

```{r normalize-cols}
messy <- data.frame(
  check.names = FALSE,
  "first column"  = 1:3,
  "if"            = 4:6
)
clean <- ptr_normalize_column_names(messy)
names(clean)
#> "first_column" "if_"

ptr_app("ggplot(clean, aes(x = var, y = var)) + geom_point()")
```

Uploaded data goes through the same normalization automatically; the manual call is only needed for data already in your R session. `?ptr_normalize_column_names` documents the exact rules (`make.names()` plus uniqueness + reserved-word collision avoidance).

## L2 — Embed in your own Shiny app

L2 is "I already have a Shiny app and want to drop a ggpaintr-driven block somewhere inside it, keeping ggpaintr's default layout." The L2 surface is **self-contained** — every L2 function owns its own `.ptr-app` theme scope and bundled asset dependency, so it drops straight into a host page with no scaffolding (this is the `ptr_<x>` half of the naming convention; `ptr_ui_<x>` bare pieces are L3). There is exactly one default-layout block — the module pair — plus, for two or more linked plots, the shared trio:

| L2 function | Role |
|---|---|
| `ptr_module_ui(id, formula)` / `ptr_module_server(id, formula)` | One self-contained default-layout block (sidebar controls + plot/code/error), namespaced by `id`. |
| `ptr_shared(formulas, …)` → `obj` | The shared **coordinator**: a pure object built once from the full formula set. Multi-instance only. |
| `ptr_shared_panel(obj)` | The one standalone cross-formula shared panel (self-contained). |
| `ptr_shared_server(obj)` | The reactive bundle threaded into each module via `shared_state =`. |

There is **no** free-form controls-here / plot-there split at L2: the only default-layout block is the self-contained module pair. Placing the controls in one region and the plot in another — or replacing a built-in pane with your own renderer — is **L3** (own the layout / render path), covered below. L2 is strictly "default layout, dropped into your app."

### Module path — `ptr_module_ui()` / `ptr_module_server()`

The compact option. One `id`, controls + outputs live next to each other inside the module's namespace.

```{r module-app}
formula <- "ggplot(iris, aes(var, var, color = var)) + geom_point()"

ui <- shiny::fluidPage(
  shiny::titlePanel("My host app"),
  ptr_module_ui("p", formula)
)
server <- function(input, output, session) {
  ptr_module_server("p", formula)
}

shiny::shinyApp(ui, server)
```

`ptr_module_server()` returns the `ptr_state` from `ptr_init_state()` so embedders that want to react to it (e.g. call `ptr_gg_extra()`, read `ptr_extract_*()`) can capture it; otherwise the side effects are all that matter. `...` is forwarded to `ptr_init_state()` — so `ui_text =`, `checkbox_defaults =`, `expr_check =`, `safe_to_remove =`, and the shared-binding arguments are all available here too.

### Shared widgets: single instance vs. multiple instances

A placeholder annotated `keyword(shared = "<key>")` is driven by **one** widget in lockstep wherever that key appears — across `aes()`, pipeline stages, and (in the multi-plot case) across formulas. *How* that widget surfaces is decided **per key by how many formulas reference it** — the **partition rule**:

> A shared key referenced in **exactly one** formula → that formula's inline **shared section** (rendered inside its own control panel).
> A shared key referenced in **two or more** formulas → the one standalone **shared panel**.
>
> *Example*: `f1 = sharedA + sharedA + sharedB`, `f2 = sharedC + sharedC + sharedB` ⟹ f1's section holds `sharedA`, f2's section holds `sharedC`, the standalone panel holds `sharedB`.

The practical consequence is a hard split by **instance count**:

- **One ggpaintr instance** ⟹ no coordinator, no panel, *ever*. Every shared key is formula-local by definition, so it auto-renders in that instance's inline shared section. The single-instance realization is L1 `ptr_app()` — write the `shared = "..."` annotation and you are done:

```{r single-instance-shared}
# One instance, one key used twice: the widget auto-renders inline. No coordinator.
ptr_app(
  "ggplot(iris, aes(x = var(shared = 'col'), y = var(shared = 'col'),
                    color = Species)) + geom_point()"
)
```

  `ptr_shared()` / `ptr_shared_panel()` / `ptr_shared_server()` are **strictly multi-instance API** — a single-instance embedder never sees them. This is enforced, not just advised: `ptr_module_server()` given a `shared = "..."` formula with no coordinator wiring **errors on purpose**, telling you to build the coordinator. (One ggpaintr block driven entirely by one shared key is L1's `ptr_app()`, above; the module path is for *embedding alongside other instances*.)

- **Multiple instances** ⟹ you *must* build the coordinator once with `ptr_shared(formulas = list(…))`. It computes the partition: cross-formula keys (≥2 formulas) go to the one `ptr_shared_panel(obj)`; each formula's formula-local keys still render inline in that module. `obj` is a pure, non-reactive single source of truth, so the UI and server can never disagree about which key lives where.

### Multiple linked instances — the shared trio

When you embed two or more ggpaintr blocks in your own app and want one widget to drive several of them, build the coordinator once and feed it to the three consumers. The contract:

- `ptr_shared(formulas, shared_ui = list(), ui_text = NULL, expr_check = TRUE, draw_all_label = "Draw all")` → `obj`. Pass the **same list of formula strings** you will hand to the modules. `shared_ui` maps each non-auto key to a `function(id) -> shiny.tag`; a `var(shared = "...")` consumer needs no entry (its picker is built for you), and a built-in `num`/`text` shared key also auto-builds if you omit it. `obj` carries the computed partition (`obj$panel_keys`, `obj$local_keys_by_formula`).
- `ptr_shared_panel(obj, css = NULL)` — the **one** standalone panel, holding exactly `obj`'s cross-formula keys. Self-contained (owns its `.ptr-app` scope + assets); place it once, wherever you want on the page. `css =` is the L2 restyle hook (see `?ptr_app` / `?ptr_css`).
- `ptr_shared_server(obj)` — call **once at the top level** of your server (never inside `moduleServer()`). Returns a `ptr_shared_state` bundle.
- `ptr_module_server(id, formula, shared_state = <bundle>)` — pass the bundle to every per-plot module. Each module renders its own formula-local shared keys inline and reads the panel keys from the bundle.

```{r l2-shared}
plots <- list(
  "ggplot(iris, aes(x = var(shared = 'metric'), y = Sepal.Length, fill = Species)) + geom_boxplot()",
  "ggplot(iris, aes(x = var(shared = 'metric'), y = Sepal.Width,  fill = Species)) + geom_violin()"
)

obj <- ptr_shared(formulas = plots,
                   shared_ui = list(metric = function(id) shiny::selectInput(id, "Metric", names(iris))))

ui <- shiny::fluidPage(
  shiny::titlePanel("My host app"),
  ptr_shared_panel(obj),
  shiny::fluidRow(
    shiny::column(6, ptr_module_ui("plot_1", plots[[1]], shared = obj)),
    shiny::column(6, ptr_module_ui("plot_2", plots[[2]], shared = obj))
  )
)
server <- function(input, output, session) {
  sh <- ptr_shared_server(obj)
  ptr_module_server("plot_1", plots[[1]], shared_state = sh)
  ptr_module_server("plot_2", plots[[2]], shared_state = sh)
}

shiny::shinyApp(ui, server)
```

`ptr_shared_panel()` emits a **Draw all** button (top-level input id `ptr_shared_draw_all`) whenever the coordinator spans two or more formulas; the per-module **Update plot** buttons still work independently. Building `ptr_shared()` from a `formulas` list that declares no `shared = "..."` annotation is an error — drop the coordinator and use the plain module path if nothing is shared.

Here `metric` is used in both formulas, so the partition rule sends it to the standalone `ptr_shared_panel(obj)`. To see the partition in action, mix a cross-formula key with formula-local ones:

```{r l2-shared-partition}
plots <- list(
  "ggplot(iris, aes(x = var(shared = 'ax1'), y = var(shared = 'ax1'),
                    color = Species)) + geom_point(size = num(shared = 'sz'))",
  "ggplot(iris, aes(x = var(shared = 'ax2'), y = Sepal.Width,
                    color = Species)) + geom_point(size = num(shared = 'sz'))"
)

obj <- ptr_shared(
  formulas  = plots,
  # shared_ui = list(sz = function(id) shiny::sliderInput(id, "Size", 1, 6, 3))
)
# sz → both formulas → standalone panel.  ax1 → only plot_1's inline section.
# ax2 → only plot_2's inline section.
obj$panel_keys                       # "sz"

ui <- shiny::fluidPage(
  ptr_shared_panel(obj),             # holds sz only
  shiny::fluidRow(
    shiny::column(6, ptr_module_ui("plot_1", plots[[1]], shared = obj)),  # ax1 inline
    shiny::column(6, ptr_module_ui("plot_2", plots[[2]], shared = obj))   # ax2 inline
  )
)
server <- function(input, output, session) {
  sh <- ptr_shared_server(obj)
  ptr_module_server("plot_1", plots[[1]], shared_state = sh)
  ptr_module_server("plot_2", plots[[2]], shared_state = sh)
}

shiny::shinyApp(ui, server)
```

### How input ids are built — and how to avoid colliding with them

The framework names every widget it emits from the formula's syntax, so the same formula always produces the same ids. When you mix ggpaintr UI with your own widgets in the same Shiny namespace, the cleanest fix is to wrap the embed in `ptr_module_ui()` / `ptr_module_server()` with a unique `id` — everything ggpaintr emits then lives under `<id>-…` and your top-level inputs cannot collide. If you can't namespace, the surface to avoid is:

**Layer names.** Each top-level call in the formula contributes a layer name equal to the function being called (`ggplot`, `geom_point`, `facet_wrap`, …). Repeats get `-2`, `-3`, … suffixes:

| Formula fragment                          | Layer names                      |
|-------------------------------------------|----------------------------------|
| `geom_point() + geom_smooth()`            | `geom_point`, `geom_smooth`      |
| `geom_point() + geom_point()`             | `geom_point`, `geom_point-2`     |

**Per-placeholder input ids** follow `<layer>_<path>_<keyword>_<shared-or-NA>`, where `<path>` is the underscore-joined positional index path into the call. For example, `aes(x = var, y = var)` inside the `ggplot()` layer yields `ggplot_1_1_var_NA` and `ggplot_1_2_var_NA`; a shared annotation `var(shared = "x_col")` lands at `ggplot_1_1_var_x_col`.

**Layer-derived ids:**

| Pattern                                          | What it is                                                  |
|--------------------------------------------------|-------------------------------------------------------------|
| `<layer>_checkbox`                               | "include this layer" toggle (no checkbox on `ggplot`)       |
| `<layer>_<path>_stage_enabled`                   | per-stage toggle on `%>%` / `|>` data-arg pipelines         |
| `<placeholder-id>_ui`                            | `renderUI` container for a consumer placeholder             |
| `<layer>_subtab`, `ptr_layer_content_<layer>`    | internal layer-nav and panel-container ids                  |

**Top-level package-owned ids.** The framework writes to a fixed set of `ptr_`-prefixed top-level ids and they are **not** user-configurable: `ptr_plot` (plotOutput), `ptr_error` (uiOutput), `ptr_code` (verbatimTextOutput), `ptr_update_plot` (draw button), the cross-formula panel ids `ptr_shared_draw_all` / `ptr_shared_errors`, plus internal layer-nav inputs `ptr_layer_select` / `ptr_layer_tabset`. Treat the whole `ptr_` prefix as reserved.

> Rule of thumb: in a host app's top-level namespace, don't author Shiny inputs or outputs whose ids start with `ptr_`, or match `<layer-name-from-your-formula>_…`. If that feels brittle, wrap the embed in `ptr_module_ui()` / `ptr_module_server()` with a unique `id` and the collision surface disappears entirely.

## L3 — Own the layout and the render path

L3 is the level finer than L2: instead of one self-contained default-layout block, **every piece of ggpaintr's UI has its own bare exported function** (the `ptr_ui_<x>` half of the naming convention — emits only its widgets, no wrapper, no assets), so you can pick exactly the pieces you want, place each one anywhere in your own Shiny, and — because the server returns the `ptr_state` — swap ggpaintr's renderer for your own. The server contract is unchanged from L2: the pieces write to / read from the same canonical ids `ptr_server()` already targets, so a piece you never place is simply a no-op.

L3 owns two things L2 does not: the **markup and placement** of every pane, and the **render path** (your own `plotly` / `ggiraph` / custom renderer off `state`). Both are covered below.

### The bare pieces and combinators

The pieces are **truly orthogonal** — they take an `id` (and, where relevant, a `formula`) and nothing else. There are **no** `error=` / `code_toggle=` flags on `ptr_ui_plot()`; layered behaviour is added by **combinators** that wrap already-built pieces.

| Function | What it emits | Key arguments |
|---|---|---|
| `ptr_ui_page(…, page, css)` | The optional page shell: a Bootstrap-3 page + the single `.ptr-app` theme scope + the deduped asset bundle, wrapping the pieces you pass in `…` | `page` (default `shiny::fluidPage`), `css` |
| `ptr_ui_header(title)` | The branded header bar (logo + title) | `title` (default `"ggpaintr"`) |
| `ptr_ui_controls(id, formula, …)` | The generated control widgets (layer picker, per-layer panels, *Update plot*, inline shared section) | `id`, `formula`, `ui_text`, `checkbox_defaults`, `expr_check`, `shared` |
| `ptr_ui_plot(id)` | The bare plot card (`ptr_plot`) — no error slot, no toggle | `id` |
| `ptr_ui_error(id)` | The bare inline error slot (`ptr_error`) | `id` |
| `ptr_ui_code(id, style)` | The bare generated-code pane (`ptr_code`) | `style = "panel"` (plain, always visible — default) or `"window"` (slide-out chrome, only meaningful via the toggle combinator) |
| `ptr_ui_inline_error(plot, error)` | **Combinator**: nests an error piece inside a plot piece's card body | `plot`, `error` |
| `ptr_ui_toggle_code(plotish, code)` | **Combinator**: wraps a plot-ish piece + code piece into the `</>` slide-out toggle layout | `plotish`, `code` |
| `ptr_ui_shared_panel(obj)` | The bare cross-formula shared panel (L3 counterpart of `ptr_shared_panel()`; no `css`) | `obj` |
| `ptr_ui_assets(css)` | The CSS/JS bundle, **escape hatch only** for roots `ptr_ui_page()` can't cover (`navbarPage`, bslib/BS5) | `css` |

Two facts shape how the bare pieces behave:

1. **`ptr_ui_page()` is the only scaffolding to remember — and it is optional.** ggpaintr's controls use a `shinyWidgets::pickerInput()` and the Bootstrap grid, which need Bootstrap's own CSS/JS — Shiny loads those only when the outermost UI object is a Bootstrap page builder. Separately, the bundled theme is scoped under a single `.ptr-app`, and the pieces are deliberately bare (a per-piece self-wrap would stack N×`100vh` of empty background when pieces are vertical siblings). `ptr_ui_page()` resolves both at once: it *is* the Bootstrap page and owns the single `.ptr-app` scope + the deduped assets. You may instead write your own bare Shiny — but then you own that scaffolding (the `navbarPage`/bslib recipe below shows exactly what to reproduce).
2. **Combinators add behaviour; flags do not.** A standalone `ptr_ui_code()` is a plain, always-visible panel that needs no wiring. The inline error and the draggable slide-out code window are produced by composing pieces through `ptr_ui_inline_error()` and `ptr_ui_toggle_code()` — they are pure DOM-structure helpers with no server coupling (the server registers `ptr_plot`/`ptr_error`/`ptr_code` regardless). They **nest**: `ptr_ui_toggle_code(ptr_ui_inline_error(ptr_ui_plot(id), ptr_ui_error(id)), ptr_ui_code(id))` is byte-for-byte the output block `ptr_app()` and `ptr_module_ui()` render internally.

Ids must line up: pass the same `id` to every piece and to the server (a `shiny::moduleServer(id, …)` wrapping `ptr_server()`, or `ptr_module_server(id, …)`). The single-embedding case needs neither the `id` nor the wrapper. Hand-placing pieces makes id collisions most likely here, so keep the whole `ptr_`-prefixed top-level id set reserved — the full list and the namespacing fix are the single source of truth in the *"How input ids are built"* section above (not duplicated here).

### A fully hand-laid-out page

Header, controls, plot, error, and code each in their own region — bare pieces, no flags:

```{r l3-pieces}
formula <- "ggplot(iris, aes(x = var, y = var, color = var)) + geom_point()"

ui <- ptr_ui_page(                          # Bootstrap page + single .ptr-app + assets
  ptr_ui_header("Iris explorer"),
  shiny::fluidRow(
    shiny::column(4, ptr_ui_controls(formula = formula)),
    shiny::column(8, ptr_ui_plot())         # bare plot card; error placed below
  ),
  ptr_ui_error(),                           # error banner in its own row
  ptr_ui_code()                             # plain, always-visible code card
                                            #   (carries its own "Generated code" header)
)
server <- function(input, output, session) {
  ptr_server(input, output, session, formula)   # binds ptr_plot / ptr_error / ptr_code
}

shiny::shinyApp(ui, server)
```

Swap the page builder with `page =` when you want a different Bootstrap-3 root — `ptr_ui_page(…, page = shiny::fillPage)` for an edge-to-edge layout, for instance. The contract is *any* BS3 page builder whose `…` are tag children (`fluidPage` (default), `fixedPage`, `fillPage`, `bootstrapPage`, `basicPage`). It is **not** `navbarPage` (that needs a positional `title` + `tabPanel()` children, not free tag children) and **not** a bslib/BS5 page (the bundled CSS is Bootstrap-3-scoped — that is `ptr_app_bslib()`'s domain). For those roots, decompose by hand (next).

`ptr_ui_code()` (style `"panel"`, the default) already renders its own bordered card with a "Generated code" header — drop it straight into your layout; do **not** bury it inside a collapsed `shiny::tags$details()`, or only the disclosure summary shows. (If you *want* it collapsible, pass `open = NA` to `tags$details()` so it starts expanded, or accept that it is hidden until the user clicks.) All three panes — plot, error, code — are empty until the first *Update plot* click, exactly as in the bundled app.

### The familiar slide-out code window — the combinator recipe

Want the slide-out code window and inline error while still owning the surrounding layout? Compose the pieces with the two combinators — this nested call is exactly what `ptr_app()` / `ptr_module_ui()` render internally, so behaviour and performance match the all-in-one path:

```{r l3-pieces-toggle}
formula <- "ggplot(iris, aes(x = var, y = var, color = var)) + geom_point()"

ui <- ptr_ui_page(
  ptr_ui_header("Iris explorer"),
  shiny::sidebarLayout(
    shiny::sidebarPanel(ptr_ui_controls(formula = formula)),
    shiny::mainPanel(
      ptr_ui_toggle_code(                                  # </> slide-out toggle …
        ptr_ui_inline_error(ptr_ui_plot(), ptr_ui_error()),# … around plot + inline error
        ptr_ui_code()                                      # … wrapped as the slide-out window
      )
    )
  )
)
# server unchanged
server <- function(input, output, session) {
  ptr_server(input, output, session, formula)   # binds ptr_plot / ptr_error / ptr_code
}

shiny::shinyApp(ui, server)
```

`ptr_ui_inline_error()` nests the error slot in the plot card body; `ptr_ui_toggle_code()` injects the `</>` button and wraps the code piece in the draggable `.ptr-code-window`. The combinators are DOM-only — no extra server wiring; `ptr_server()` already registers all three outputs.

### Custom or `navbarPage` roots — decompose by hand

`ptr_ui_page()` covers every Bootstrap-3 page builder whose `…` are tag children. Two roots it deliberately does **not** cover: `navbarPage()` (positional `title` then `tabPanel()` children, not free tags) and bslib/BS5 pages (the bundled CSS is Bootstrap-3-scoped — use `ptr_app_bslib()` for a bslib theme). For those, hand-build what `ptr_ui_page()` expands to — a Bootstrap page, `ptr_ui_assets()` once (this is its **only** sanctioned use — the navbar/bslib escape hatch, never part of normal L2/L3 composition), and one `div(class = "ptr-app")` around the pieces:

```{r l3-decompose}
ui <- shiny::navbarPage(
  "Iris explorer",                            # navbarPage needs a positional title
  shiny::tabPanel(
    "Plot",
    ptr_ui_assets(),                          # the bundle, once (self-deduping) — escape hatch
    shiny::tags$div(
      class = "ptr-app",                      # the single theme scope
      shiny::sidebarLayout(
        shiny::sidebarPanel(ptr_ui_controls(formula = formula)),
        shiny::mainPanel(
          ptr_ui_toggle_code(
            ptr_ui_inline_error(ptr_ui_plot(), ptr_ui_error()),
            ptr_ui_code()
          )
        )
      )
    )
  ),
  shiny::tabPanel("About", "Built with ggpaintr.")
)
# server unchanged
server <- function(input, output, session) {
  ptr_server(input, output, session, formula)   # binds ptr_plot / ptr_error / ptr_code
}

shiny::shinyApp(ui, server)
```

This is exactly the body of `ptr_ui_page()` (`page(ptr_ui_assets(css), div(class = "ptr-app", …))`) inlined into the one tab — so it doubles as a transparent view of what the shell does for you. Keep `ptr_ui_assets()` and the `.ptr-app` wrapper *inside* each tab that hosts ggpaintr pieces; the asset bundle is self-deduping (htmlDependency), so repeating it across tabs costs nothing.

### Own the render path — custom renderers off `state` (L3)

"I want my own renderer." Both `ptr_server()` and `ptr_module_server()` **return** the `ptr_state`, so you can swap ggpaintr's renderer for Plotly / ggiraph / any custom host. The runtime stays inside ggpaintr; you replace only the *output*. Owning the render path is definitionally L3 (the *enabling* return value lives at the L2 server boundary, but using it to replace a pane is owning the render path).

**The canonical pattern** nests `ptr_server()` inside *your own* `shiny::moduleServer(id, …)`, and places your output widget at `shiny::NS(id)(…)` in the UI. Namespacing is then automatic and plain-Shiny-idiomatic — you never touch any internal namespace plumbing:

```r
# UI:     plotly::plotlyOutput(shiny::NS("plot1")("my_plot"))   + ptr_ui_controls("plot1", formula)
# server: shiny::moduleServer("plot1", function(input, output, session) {
#           state <- ptr_server(input, output, session, formula)
#           output$my_plot <- plotly::renderPlotly({ ... state$runtime() ... })
#         })
```

There are two reading paths off a `ptr_state`, and the choice depends on whether you are inside a reactive context:

| Inside a `renderX({...})` / `reactive({...})` / `observeEvent({...})` | Outside reactive contexts (download handlers, snapshots) |
|----|----|
| Read `state$runtime()` and unpack its slots (`$ok`, `$plot`, `$code_text`, `$error`). This takes the reactive dependency that wires re-renders to *Update plot* clicks. | Call `ptr_extract_plot(state)` / `ptr_extract_code(state)` / `ptr_extract_error(state)`. These wrap `shiny::isolate()` — current value, no reactive dependency. |

Mixing the two in a render block — calling `ptr_extract_plot(state)` inside `renderPlotly({...})` — silently breaks reactivity (the block fires once on mount and never again). Stick to `state$runtime()` inside reactive blocks.

#### Custom plot renderer — `state$runtime()$plot` inside `renderPlotly()`

Render `ptr_ui_controls(id, formula)` for the widgets, place your own output container at `shiny::NS(id)(...)`, and never place `ptr_ui_plot(id)` — with no UI slot bound to it the built-in `ptr_plot` output stays inert, so there is no bundled pane at all. The server is the canonical pattern: `ptr_server()` inside your own `moduleServer(id)`.

```{r l3-plotly}
formula <- "ggplot(iris, aes(x = var, y = var, color = var)) + geom_point()"

ui <- ptr_ui_page(
  shiny::fluidRow(
    shiny::column(5, ptr_ui_controls("plot1", formula)),                   # widgets only
    shiny::column(7, plotly::plotlyOutput(shiny::NS("plot1")("custom_plot"),
                                          height = "500px") |>
                    ptr_ui_toggle_code(ptr_ui_code("plot1")))               # your own output
  )
)
server <- function(input, output, session) {
  shiny::moduleServer("plot1", function(input, output, session) {
    state <- ptr_server(input, output, session, formula)   # runtime wired; ptr_plot unbound = inert
    output$custom_plot <- plotly::renderPlotly({
      res <- state$runtime()
      shiny::req(isTRUE(res$ok), res$plot)
      plotly::ggplotly(res$plot)
    })
  })
}

shiny::shinyApp(ui, server)
```

`req(isTRUE(res$ok), res$plot)` keeps the panel quiet between draws and on transient error states. `vignette("ggpaintr-gallery")` §5 *Interactive output packages* is the canonical worked-example reference for this pattern — the full plotly and ggiraph setups end-to-end, including the ggiraph variant with click handling and a custom-placeholder colour picker. The chunk above is the minimal scaffold; gallery §5 is the runnable recipe.

If you would rather keep ggpaintr's chrome *and* add a second custom view, capture the module's state and render a second output off it — `ptr_module_ui("p", formula)` in the UI plus `state <- ptr_module_server("p", formula)` in the server — but `ptr_module_ui()` always renders the bundled plot pane alongside your custom one. Use the bare-piece composition above when you want exactly one (custom) plot.

#### Custom code panel

Bind to `state$runtime()` inside `renderText({...})` and pull `code_text` off it. Inside the same `moduleServer(id)` as the `state <-` assignment above:

```{r l3-code-pane}
output$my_code <- shiny::renderText({
  state$runtime()$code_text %||% ""
})
```

For non-reactive contexts (a download handler, a snapshot test), use the public accessor `ptr_extract_code(state)`. It returns a single string and includes any `ptr_gg_extra()` expressions captured on the state.

#### Custom error UI

`state$runtime()$error` is the latest runtime error string (or `NULL` when the last cycle succeeded). Inside the same `moduleServer(id)`:

```{r l3-error-ui}
output$my_status <- shiny::renderUI({
  msg <- state$runtime()$error
  if (is.null(msg)) shiny::tags$div(class = "status-ok", "Plot is up to date")
  else              shiny::tags$div(class = "status-error", msg)
})
```

`ptr_extract_error(state)` is the non-reactive counterpart.

#### Programmatic layer injection — `ptr_gg_extra()`

`ptr_gg_extra(state, ...)` evaluates one or more `ggplot2` expressions and attaches the results as "extras" on the state. The runtime folds them into the rendered plot during the next cycle, and `state$runtime()$code_text` (and `ptr_extract_code(state)`) include them in the printable code. Each call replaces the previously captured extras — pass everything you want layered on in a single call.

```{r l3-gg-extra}
formula <- "ggplot(mtcars, aes(x = mpg, y = hp)) + geom_point()"

ui <- shiny::fluidPage(
  shiny::actionButton("add_log", "Toggle log-scale"),
  ptr_module_ui("p", formula)
)
server <- function(input, output, session) {
  state <- ptr_module_server("p", formula)
  shiny::observeEvent(input$add_log, {
    ptr_gg_extra(state, ggplot2::scale_x_log10())
  })
}

shiny::shinyApp(ui, server)
```

Eval errors from the captured expressions leave the existing extras untouched (atomic update). Extras are also suppressed automatically when the underlying runtime reports a failure, so stale extras from a prior successful draw never surface during an error state.

#### Shared widgets driving custom renderers

Combine the multi-instance shared coordinator with custom outputs: one `ptr_shared_server(obj)` bundle feeds every module's `ptr_server()`, and each module's returned `state` feeds its own `plotly::renderPlotly()`. Each plot gets its own `moduleServer(id)` so the per-placeholder ids never collide; the cross-formula key lives in the standalone `ptr_ui_shared_panel(obj)` (its keys are excluded from each `ptr_ui_controls(id, formula, shared = obj)`).

```{r l3-shared}
plots <- list(
  "ggplot(iris, aes(x = var(shared = 'metric'), y = Sepal.Length, fill = Species)) + geom_boxplot()",
  "ggplot(iris, aes(x = var(shared = 'metric'), y = Sepal.Width,  fill = Species)) + geom_violin()"
)

obj <- ptr_shared(formulas = plots,
                   shared_ui = list(metric = function(id) shiny::selectInput(id, "Metric", names(iris))))

ui <- ptr_ui_page(
  ptr_ui_shared_panel(obj),                                  # cross-formula panel (bare, L3)
  shiny::fluidRow(
    shiny::column(3, ptr_ui_controls("plot_1", plots[[1]], shared = obj)),
    shiny::column(9, plotly::plotlyOutput(shiny::NS("plot_1")("custom")))
  ),
  shiny::fluidRow(
    shiny::column(3, ptr_ui_controls("plot_2", plots[[2]], shared = obj)),
    shiny::column(9, plotly::plotlyOutput(shiny::NS("plot_2")("custom")))
  )
)
server <- function(input, output, session) {
  sh <- ptr_shared_server(obj)                               # top level, once
  shiny::moduleServer("plot_1", function(input, output, session) {
    state <- ptr_server(input, output, session, plots[[1]], shared_state = sh)
    output$custom <- plotly::renderPlotly({
      res <- state$runtime(); shiny::req(isTRUE(res$ok), res$plot)
      plotly::ggplotly(res$plot)
    })
  })
  shiny::moduleServer("plot_2", function(input, output, session) {
    state <- ptr_server(input, output, session, plots[[2]], shared_state = sh)
    output$custom <- plotly::renderPlotly({
      res <- state$runtime(); shiny::req(isTRUE(res$ok), res$plot)
      plotly::ggplotly(res$plot)
    })
  })
}

shiny::shinyApp(ui, server)
```

The coordinator feeds the modules; the modules feed your custom renderers. `ptr_shared_server(obj)` runs once at the top level (never inside `moduleServer()`); its bundle is threaded into each plot's `ptr_server()` via `shared_state =`.

### L3 with shared widgets — default panes

If you want the bare default panes (not a custom renderer) under L3, the shape is the same minus the custom outputs: build `obj`, place `ptr_ui_shared_panel(obj)` once, give each plot its own `ptr_ui_controls(id, formula, shared = obj)` + bare plot piece, and run one top-level `ptr_shared_server(obj)` threaded into each module's `ptr_server()`:

```{r l3-pieces-shared}
plots <- list(
  "ggplot(iris, aes(x = var(shared = 'metric'), y = Sepal.Length, fill = Species)) + geom_boxplot()",
  "ggplot(iris, aes(x = var(shared = 'metric'), y = Sepal.Width,  fill = Species)) + geom_violin()"
)

obj <- ptr_shared(formulas = plots,
                   shared_ui = list(metric = function(id) shiny::selectInput(id, "Metric", names(iris))))

ui <- ptr_ui_page(
  ptr_ui_shared_panel(obj),
  shiny::fluidRow(
    shiny::column(6, ptr_ui_controls("p1", plots[[1]], shared = obj), ptr_ui_plot("p1")),
    shiny::column(6, ptr_ui_controls("p2", plots[[2]], shared = obj), ptr_ui_plot("p2"))
  )
)
server <- function(input, output, session) {
  sh <- ptr_shared_server(obj)
  shiny::moduleServer("p1", function(input, output, session) {
    ptr_server(input, output, session, plots[[1]], shared_state = sh)
  })
  shiny::moduleServer("p2", function(input, output, session) {
    ptr_server(input, output, session, plots[[2]], shared_state = sh)
  })
}

shiny::shinyApp(ui, server)
```

### Placing individual placeholder widgets — `build_ui_for()`

`ptr_ui_controls()` is the whole controls panel, and it is the supported way to render the generated widgets: ggpaintr does not currently expose a public accessor for the parsed node tree, so hand-placing the *individual* placeholder widgets is not a supported public workflow. The `build_ui_for()` S3 generic — the one ggpaintr uses internally to turn one parsed node into its widget — is exported and its per-node contract is documented at `?build_ui_for` for advanced users who already hold node objects, but the means of obtaining those nodes is internal and may change between versions. To add custom widget *types*, register them instead — see `vignette("ggpaintr-customization")`.
