---
title: "ggpaintr Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ggpaintr Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

## Overview

`ggpaintr` now focuses on the `ggpaintr` workflow. A plot is described as one
ggplot-like formula string. Placeholder tokens inside that string are converted
into Shiny controls, completed with current inputs, evaluated into a plot, and
rendered with generated code and readable runtime feedback.

Supported placeholders are:

- `var`
- `text`
- `num`
- `expr`
- `upload`

`upload` currently supports `.csv` and `.rds`.

## Parsing a formula

```{r}
library(ggpaintr)
library(ggplot2)

obj <- paintr_formula(
  "ggplot(data = iris, aes(x = var, y = var)) +
    geom_point(aes(color = var), size = num) +
    labs(title = text) +
    facet_wrap(expr)"
)

names(obj$expr_list)
names(obj$keywords_list$ggplot)
```

## Building runtime output

```{r}
runtime <- paintr_build_runtime(
  obj,
  list(
    "ggplot+3+2" = "Sepal.Length",
    "ggplot+3+3" = "Sepal.Width",
    "ggplot+3+4" = "Species",
    "geom_point+2" = 2.5,
    "labs+2" = "Iris scatter",
    "facet_wrap+2" = "~ Species",
    "geom_point+checkbox" = TRUE,
    "labs+checkbox" = TRUE,
    "facet_wrap+checkbox" = TRUE
  )
)

runtime$code_text
inherits(runtime$plot, "ggplot")
```

## Launching an app

```{r, eval = FALSE}
ggpaintr_app("
ggplot(data = iris, aes(x = var, y = var)) +
  geom_point(aes(color = var), size = num) +
  labs(title = text) +
  facet_wrap(expr)
")
```

The generated app:

- renders one control tab per layer
- delays `var` selectors until upload-backed data is available
- clears the plot on failure
- shows `Input error:` and `Plot error:` messages inline
- preserves generated code when completion succeeds but plotting fails

## Exporting a standalone app

```{r}
out_file <- tempfile(fileext = ".R")
generate_shiny(obj, list(), out_file, style = FALSE)
file.exists(out_file)
```

## Current behavior boundary

- Structural parse failures still stop the app before launch.
- `var` placeholders still require a data source during UI preparation.
- Missing local data objects such as `unknown_object` are deferred to draw time.
- Render-time ggplot failures such as missing faceting variables are surfaced by
  the same runtime error path as other plot failures.
