---
title: "ggpaintr Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ggpaintr Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

```{r clipboard, echo = FALSE}
xaringanExtra::use_clipboard()
```

```{r libs}
library(ggpaintr)
library(ggplot2)
```

## 1. Motivation

Teaching ggplot, exploring a new dataset, or running a quick demo usually
means writing a small Shiny app so the audience can click through
variables. That app is almost always boilerplate: a sidebar of
`selectInput()`s, a `renderPlot()`, and a harness that rebuilds the plot
whenever inputs change.

`ggpaintr` removes the boilerplate. You describe the plot once with a
**formula string** that looks like ordinary ggplot code, and `ggpaintr`
turns each placeholder token in the string into the right Shiny input
widget, evaluates the completed call, and renders both the plot and the
generated code.

The workflow is aimed at three situations:

- **Teaching** — show how changing one aesthetic (`color = var`) changes
  the plot without writing UI code.
- **Exploration** — point the app at a dataset and try several
  aesthetic-to-column assignments before committing to one.
- **Prototyping** — stand up an interactive plot in minutes to share
  with collaborators, then later migrate the generated ggplot call into
  a reproducible script or report.

This vignette covers **Level 1** of `ggpaintr`: spinning up a one-call
app. Levels 2 and 3 (embedding ggpaintr into your own Shiny app, and
driving the runtime directly without Shiny) are covered in
`vignette("ggpaintr-extensibility")` and
`vignette("ggpaintr-placeholder-registry")`.

## 2. The formula syntax

A ggpaintr formula is a single string that looks like a ggplot call,
except that some argument positions hold a **placeholder keyword**
instead of a literal value. When the app starts, each keyword is
replaced by a Shiny input widget whose current value is substituted
back into the call at draw time.

The `+` operator still separates ggplot layers. Each call to
`ggplot()`, `geom_*()`, `labs()`, `facet_*()`, `scale_*()`, etc. is one
layer; layers other than `ggplot()` get an on/off checkbox in the UI so
the viewer can add or drop them.

### 2.1 The five placeholder keywords

`ggpaintr` ships with five built-in keywords. They differ in what
widget they produce and what kind of value they feed back into the
ggplot call.

| Keyword  | Widget                                | Produces at runtime          | Typical use                                 |
|----------|---------------------------------------|------------------------------|---------------------------------------------|
| `var`    | `shinyWidgets::pickerInput()` of data-frame columns | column name as a symbol      | mappings like `x = var`, `color = var`      |
| `text`   | `textInput()`                         | string literal               | `labs(title = text)`, category filters       |
| `num`    | `numericInput()`                      | numeric value                | `size = num`, `alpha = num`                  |
| `expr`   | `textInput()` (parsed as R expression)| arbitrary R expression       | `facet_wrap(expr)`, `aes(x = expr)` maths    |
| `upload` | `fileInput()` + name `textInput()`    | a named data frame in memory | supply the dataset from the browser         |

Each subsection below shows the minimum formula that exercises the
keyword.

#### `var` — choose a column

`var` binds to a column of the active dataset. In the UI it becomes a
`selectInput()` whose choices are the column names of the data frame
(either the one you passed as `data =` or one uploaded via `upload`).
At runtime the selected name is injected as a symbol, so the expression
`aes(x = var)` becomes `aes(x = Sepal.Length)`.

```{r var-example, eval = FALSE}
ptr_app(
  "ggplot(data = mtcars, aes(x = var, y = var)) +
     geom_point()"
)
```

A few behaviors worth knowing:

- When the dataset is supplied via `data =`, `var` choices are
  populated up front.
- When the dataset is supplied by an `upload` placeholder, `var`
  widgets are **deferred** until the upload resolves. You will see a
  short placeholder message in the UI in the meantime.
- Non-syntactic column names (spaces, leading digits, symbols) are
  normalized to syntactic identifiers — see
  [§4. Data sources](#4-data-sources).

#### `text` — a string literal

`text` is a free-form string. The value is injected verbatim as a
character literal — it is **not** parsed as R code. Use it for titles,
labels, and other pure string inputs.

```{r text-example, eval = FALSE}
ptr_app(
  "ggplot(data = mtcars, aes(x = var, y = var)) +
     geom_point() +
     labs(title = text)"
)
```

#### `num` — a numeric value

`num` becomes a `numericInput()`. The value is a plain number, suitable
for arguments like `size`, `alpha`, `width`, or `binwidth`.

```{r num-example, eval = FALSE}
ptr_app(
  "ggplot(data = mtcars, aes(x = var, y = var)) +
     geom_point(size = num)"
)
```

#### `expr` — an R expression

`expr` is the escape hatch. The value you type in the widget is
**parsed as an R expression** at draw time, so you can pass things
like `log(mpg)`, `~ cyl`, or `factor(gear)`.

Because `expr` evaluates user input, it is gated by an expression
safety check. The check is on by default; see the `expr_check`
argument in §3.

```{r expr-example, eval = FALSE}
ptr_app(
  "ggplot(data = mtcars, aes(x = var, y = var)) +
     geom_point() +
     facet_wrap(expr)"
)
```

#### `upload` — load a dataset from the browser

`upload` lets the viewer supply the dataset themselves. It renders a
`fileInput()` that accepts `.csv`, `.tsv`, `.rds`, `.xlsx`, `.xls`, and `.json` files (Excel and JSON read via the suggested packages `readxl` and `jsonlite`), plus a companion
`textInput()` where the user gives the uploaded object a name
(e.g. `penguins`). Any `var` placeholders that reference that name are
wired to the uploaded data frame once the upload completes.

```{r upload-example, eval = FALSE}
ptr_app(
  "ggplot(data = upload, aes(x = var, y = var)) +
     geom_point()"
)
```

### 2.2 Combining placeholders

Placeholders compose. Here is a formula that uses all five:

```{r all-placeholders, eval = FALSE}
ptr_app(
  "ggplot(data = upload, aes(x = var, y = var)) +
     geom_point(aes(color = var), size = num) +
     labs(title = text) +
     facet_wrap(expr)"
)
```

The resulting app has:

- one upload widget (with a name field) for the data,
- three column selectors (`x`, `y`, `color`),
- one numeric input (`size`),
- one text input (title),
- one expression input (faceting formula),
- one "Update plot" button,
- one checkbox per non-`ggplot()` layer to toggle it on and off.

## 3. Level 1 — launching an app

Two entry points cover the common cases.

### 3.1 `ptr_app()`

`ptr_app()` produces a minimal Shiny app with a fluid-page shell: a
single column of controls, a draw button, and the plot + code output
below. It has no extra runtime dependencies beyond `shiny`.

Full signature:

```{r ptr-app-sig, eval = FALSE}
ptr_app(
  formula,
  envir = parent.frame(),
  ui_text = NULL,
  placeholders = NULL,
  expr_check = TRUE
)
```

| Argument       | Default            | What it does                                                                                      |
|----------------|--------------------|---------------------------------------------------------------------------------------------------|
| `formula`      | —                  | A single formula string using ggpaintr placeholders. Required.                                    |
| `envir`        | `parent.frame()`   | Environment used to resolve bare data object names (e.g. `data = mtcars`) in the formula.         |
| `ui_text`      | `NULL`             | Optional named list of UI copy overrides (labels, help text, placeholders). See extensibility vignette. |
| `placeholders` | `NULL`             | Optional custom placeholder definitions or a full registry. See placeholder-registry vignette.   |
| `expr_check`   | `TRUE`             | Controls `expr` placeholder safety validation. See §3.3.                                          |

Smallest working example:

```{r ptr-app-min, eval = FALSE}
ptr_app(
  "ggplot(data = mtcars, aes(x = var, y = var)) + geom_point()"
)
```

### 3.2 `ptr_app_bslib()`

`ptr_app_bslib()` produces the same functional app but inside a
[bslib](https://rstudio.github.io/bslib/) `page_sidebar()` shell with
cards for the plot and generated code. Use it when you want a themed
look without writing any Shiny UI code yourself.

Full signature:

```{r bslib-sig, eval = FALSE}
ptr_app_bslib(
  formula,
  envir = parent.frame(),
  ui_text = NULL,
  placeholders = NULL,
  expr_check = TRUE,
  theme = NULL,
  title = "ggpaintr"
)
```

| Argument       | Default                                              | What it does                                                                                     |
|----------------|------------------------------------------------------|--------------------------------------------------------------------------------------------------|
| `formula`      | —                                                    | A single formula string using ggpaintr placeholders. Required.                                   |
| `envir`        | `parent.frame()`                                     | Environment used to resolve bare data object names in the formula.                               |
| `ui_text`      | `NULL`                                               | Optional named list of UI copy overrides. Same contract as `ptr_app()`.                          |
| `placeholders` | `NULL`                                               | Optional custom placeholder definitions or a registry. Same contract as `ptr_app()`.             |
| `expr_check`   | `TRUE`                                               | Controls `expr` placeholder validation. See §3.3.                                                |
| `theme`        | `NULL` (→ `bslib::bs_theme(version = 5, "flatly")`)  | A `bslib::bs_theme()` object. Pass your own to restyle.                                          |
| `title`        | `"ggpaintr"`                                         | The page header title. Also overridable via `ui_text$shell$title$label`.                         |

Requires `bslib`. If the package is not installed, the function errors
at call time.

Example with a different theme:

```{r bslib-example, eval = FALSE}
ptr_app_bslib(
  "ggplot(data = mtcars, aes(x = var, y = var)) + geom_point()",
  theme = bslib::bs_theme(version = 5, bootswatch = "minty"),
  title = "mtcars explorer"
)
```

### 3.3 When to use which

- Use `ptr_app()` for teaching, quick demos, and tests — it has the
  lightest dependency footprint and matches typical RMarkdown output.
- Use `ptr_app_bslib()` when the app will be shared with non-R users
  or shown as part of a finished deliverable — the themed layout looks
  more polished with no extra effort.

Both functions return a `shiny.appobj`, so at the top level of an R
session they launch the app immediately; inside a function or script
you can assign the result and pass it to `shiny::runApp()`.

### 3.4 The `expr_check` argument

`expr` inputs are evaluated as R expressions. Because that is
effectively `eval(parse(text = input))`, `ggpaintr` runs a safety check
against a curated denylist of dangerous functions (file I/O, system
calls, environment manipulation). The `expr_check` argument controls
that check:

| `expr_check` value                                        | Behavior                                                                                     |
|-----------------------------------------------------------|----------------------------------------------------------------------------------------------|
| `TRUE` (default)                                          | Use the built-in denylist — recommended for interactive use.                                  |
| `FALSE`                                                   | Disable all checking. Only safe when the app is running locally and you trust every user.    |
| A list with `deny_list` and/or `allow_list` character vectors | Replace / augment the denylist. If both are given, denied entries are removed from the allowlist. |

Leave it at `TRUE` unless you have a specific reason to change it.

### 3.5 Choosing which layers start checked

By default every non-`ggplot()` layer in the formula starts checked. When a formula declares mutually-incompatible layers — for example, two competing geoms or alternative coordinate systems — that default produces a broken plot at app startup. The `checkbox_defaults` argument sets the initial checked state per layer, so users land on a sensible starting view and can toggle alternatives in.

```{r checkbox-defaults, eval = FALSE}
ptr_app(
  formula = paste(
    "ggplot(data = mtcars, aes(x = var, y = var)) +",
    "geom_point() +",
    "geom_smooth(method = text) +",
    "geom_line()"
  ),
  checkbox_defaults = list(
    geom_smooth = FALSE,
    geom_line   = FALSE
  )
)
```

Pass a named list. Names match layer names from the formula — use `names(ptr_parse_formula(formula)$expr_list)` to inspect them. **Duplicate layers** receive a hyphen-numeric suffix starting at `-2`: a formula with two `geom_point()` calls produces layer names `geom_point` and `geom_point-2`. To address them as a group, pass a logical vector at the bare key:

```{r checkbox-defaults-dup, eval = FALSE}
list(geom_point = c(TRUE, FALSE))      # 1st on, 2nd off
list(geom_point = FALSE)               # 1st off, 2nd defaults to TRUE (padded)
list(`geom_point-2` = FALSE)           # address one specific instance (backticks required)
```

Vectors shorter than the layer count pad with `TRUE`; longer vectors warn and truncate. Unrecognized names raise a warning and are ignored. `NA` and non-logical values raise an error. The same argument is available on `ptr_app()`, `ptr_app_bslib()`, `ptr_server()`, and `ptr_server_state()`.

### Package-global settings via `ptr_options()`

`ggpaintr` exposes a small number of session-wide flags through `ptr_options()`, which mirrors base `options()`: call with no arguments to read every current value, or with one or more named logicals to set them. Setters invisibly return the previous values, so the standard `withr::with_options()` / `on.exit()` round-trip works (`old <- ptr_options(verbose = FALSE); on.exit(do.call(ptr_options, old), add = TRUE)`).

The two settings currently exposed are:

- `verbose` (default `FALSE`) — when `TRUE`, ggpaintr emits the "Layer foo() removed (no arguments provided)." informational notice and similar pipeline diagnostics. Intended for debugging the formula pipeline; off by default for a quiet user experience.
- `checkbox_default_all_other_layer` (default `TRUE`) — fallback initial state for layer checkboxes not explicitly named in `checkbox_defaults =`. For apps with many alternative layers, invert this with `ptr_options(checkbox_default_all_other_layer = FALSE)` so every layer starts unchecked unless explicitly opted in.

See `?ptr_options` for the canonical reference.

## 4. Data sources {#4-data-sources}

There are three ways to get a data frame into the running app.

### 4.1 Pass `data = <object>` directly in the formula

The simplest option. The name on the right of `data =` is resolved
against `envir` (by default the calling frame). Any variable visible
when you call `ptr_app()` works.

```{r data-direct, eval = FALSE}
mtcars2 <- dplyr::mutate(mtcars, cyl = factor(cyl))
ptr_app(
  "ggplot(data = mtcars2, aes(x = var, y = var)) + geom_point()"
)
```

`var` choices are the columns of the named data frame and are
populated as soon as the app starts.

### 4.2 Use the `upload` placeholder

Replace `data = <object>` with `data = upload` to let the viewer
supply the dataset. The `upload` widget exposes a file picker
(accepting `.csv`, `.tsv`, `.rds`, `.xlsx`, `.xls`, and `.json`) plus a name field; the name is the
symbol the uploaded data binds to inside the generated ggplot call.

```{r data-upload, eval = FALSE}
ptr_app(
  "ggplot(data = upload, aes(x = var, y = var)) + geom_point()"
)
```

Until a file is uploaded and named, `var` widgets stay deferred and
the draw button produces an inline "data not ready" message instead of
a broken plot.

### 4.3 Non-syntactic column names

Column names that are not valid R identifiers (contain spaces, start
with a digit, use punctuation) would break `aes(x = that name)` when
injected as a symbol. `ggpaintr` normalizes them automatically:

- Data passed through an `upload` placeholder is normalized at load
  time by `ptr_normalize_tabular_data()`.
- Data passed directly with `data = <object>` is **not** normalized
  automatically — you are expected to pass a clean data frame.

If you have a data frame with problematic names and want to drive it
through `data =`, call `ptr_normalize_column_names()` yourself first:

```{r normalize, eval = TRUE}
dirty <- data.frame(`miles per gallon` = mtcars$mpg, `# cyl` = mtcars$cyl,
                    check.names = FALSE)
names(dirty)

clean <- ptr_normalize_column_names(dirty)
names(clean)
```

The signature is simply `ptr_normalize_column_names(data)`. Any object
that `as.data.frame()` accepts is allowed; the function returns a data
frame with names run through `make.names(unique = TRUE)` (with empty
names filled in as `V1`, `V2`, …).

## 5. What happens when you click "Update plot"

Knowing the pipeline at a high level helps when an app misbehaves. The
full sequence is:

```
formula string
      │
      ▼
ptr_parse_formula()        ── parses formula, finds placeholders,
      │                       builds an internal ptr_obj
      ▼
ptr_runtime_input_spec()   ── enumerates every Shiny input the app
      │                       needs (one row per placeholder, one per
      │                       layer checkbox, one per upload-name field)
      ▼
UI is built                ── each spec row becomes a widget via the
      │                       placeholder registry's build_ui hook
      ▼
user clicks "Update plot"
      │
      ▼
ptr_exec(ptr_obj, input)   ── substitutes input values into the
      │                       expression list, assembles ggplot
      │                       layers, renders the plot
      ▼
plot + code text output
```

Two of those steps are public functions you can call directly —
`ptr_parse_formula()` and `ptr_exec()`. That is the entry point to
Level 3 (driving the runtime without Shiny) covered in the
extensibility vignette. For Level 1 usage you do not need to touch
either one.

A tiny illustration of the first step:

```{r parse-demo, eval = TRUE}
obj <- ptr_parse_formula(
  "ggplot(data = mtcars, aes(x = var, y = var)) +
     geom_point(aes(color = var), size = num) +
     labs(title = text)"
)
class(obj)
names(obj$expr_list)
```

Each element of `obj$expr_list` is one ggplot layer. Each entry of
`obj$keywords_list` lists which placeholder keywords appear inside the
corresponding layer.

And the second step — inspecting the inputs the runtime expects:

```{r spec-demo, eval = TRUE}
spec <- ptr_runtime_input_spec(obj)
spec
```

Columns in the returned data frame:

- `input_id` — the Shiny input id. Treat as opaque; do not hand-author.
- `role` — `"placeholder"` (a widget value), `"layer_checkbox"` (a
  layer on/off toggle), or `"upload_name"` (the name field that
  accompanies an `upload` placeholder).
- `layer_name` — the ggplot layer the input belongs to
  (`"ggplot"`, `"geom_point"`, …).
- `param_key` — the ggplot argument name (`"x"`, `"color"`, …);
  `NA_character_` for unnamed positional placeholders.
- `keyword` — the placeholder type (`"var"`, `"num"`, …).
- `source_id` — for `upload_name` rows, the `input_id` of the
  matching `upload` placeholder row.

## 6. Going further

### Level 2 — embed ggpaintr in your own Shiny app

If you already have a Shiny app and want to drop ggpaintr controls
into one of its tabs (or override `ui_text` for a single label), the
Level 2 API (`ptr_input_ui()`, `ptr_output_ui()`, `ptr_register_*()`)
is what you want.

After reading `vignette("ggpaintr-extensibility")` you will know how
to embed ggpaintr controls in your own Shiny app and override any
label or help text.

### Level 3 — drive ggpaintr without Shiny

Every step in the §5 pipeline is a public function, so you can parse a
formula, fill in inputs from any source, and render a plot to a file
without launching a Shiny server. This is useful for batch reports,
test fixtures, and programmatic code generation.

After reading `vignette("ggpaintr-extensibility")` you will know how
to drive `ptr_exec()` directly, post-process the plot object, and
generate ggplot code strings from a formula.

### Adding new placeholder types

If none of the five built-in keywords fit your input (say you want a
date picker or a slider), the placeholder registry lets you add a new
keyword and plug it into any ggpaintr app.

After reading `vignette("ggpaintr-placeholder-registry")` you will
know how to add a new placeholder type end-to-end — every required
and optional hook, the metadata contract, and how to plug your
placeholder into `ptr_app()`, `ptr_app_bslib()`, and manual Shiny
embeds.

## 7. Current behavior boundary

A few things to keep in mind while using Level 1:

- Structural parse failures (unbalanced parens, invalid R syntax)
  stop the app before it launches.
- `var` placeholders require a data source to finish UI preparation.
  With `data = <object>` that happens immediately; with `data = upload`
  it is deferred until the file is uploaded.
- Missing local data objects (e.g. `data = unknown_object`) are
  deferred to draw time — the app launches, and the error surfaces
  inline after clicking "Update plot".
- Render-time ggplot failures (missing columns, invalid aesthetics,
  bad `expr` values) are surfaced by the same inline error path as
  other plot failures.
- Generated code is preserved when input completion succeeds but
  plotting fails, so you can inspect what ggpaintr tried to run.
