---
title: "ggpaintr in the Wild"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ggpaintr in the Wild}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment   = "#>",
  warning   = FALSE,
  message   = FALSE,
  eval      = interactive()
)
```

```{r clipboard, echo = FALSE, eval = TRUE}
xaringanExtra::use_clipboard()
```

## 1. Setup

Every chunk in this vignette is marked `eval = interactive()` — they will not run during vignette build, but each block is runnable as-is at the R prompt. Before launching any of them, load the development copy of the package and bind its internals into your global environment so S3 dispatch through internal generics (`prune_walk`, `render_walk`, `classify_walk`, `build_ui_for`, …) resolves to the namespace where the method tables actually live. This is the same setup used by `dev/scripts/feature-sweep.R`:

```{r dev-setup, eval = TRUE}
devtools::load_all(".", export_all = FALSE)

local({
  ns <- asNamespace("ggpaintr")
  for (nm in ls(ns, all.names = FALSE)) {
    assign(nm, get(nm, envir = ns), envir = globalenv())
  }
})
```

```{r libs, eval = TRUE}
library(shiny)
library(ggplot2)
library(rlang)
library(tidyverse)
```

## 2. What this vignette is

The other vignettes teach the workflow piece by piece. This one shows what the pieces add up to. Every section pairs a complete `ggplot2` (or extension-package) graphic with the `ggpaintr` formula that turns it into an interactive Shiny app. Read the original to know what is being built; read the formula to see which arguments become widgets.

Copy a chunk into a session, install the extension package named at the top of the section if you do not have it, and run it. The examples assume the setup chunks above have already executed.

## 3. Three reusable custom placeholders

ggpaintr ships five built-in placeholder keywords (`var`, `text`, `num`, `expr`, `upload`). Anything else — date pickers, sliders bound to a fixed range, multi-column selectors — is registered through one of three constructors:

- `ptr_define_placeholder_value()` — non-data-aware widgets (sliders, text, numeric, date).
- `ptr_define_placeholder_consumer()` — data-aware widgets that need the upstream column-name vector (column pickers).
- `ptr_define_placeholder_source()` — data-aware widgets that *produce* a data frame (file uploads, datasets).

Registration is process-global: call each constructor once per R session. Re-running a chunk warns about the duplicate keyword and overwrites the previous entry.

Three custom placeholders show up across the gallery: `range` (a slider returning a length-2 numeric), `colvars` (a multi-select for column names), and `date` (a date-range picker).

```{r placeholders}
ptr_define_placeholder_value(
  keyword       = "range",
  build_ui      = function(node, label = NULL, ...) {
    sliderInput(node$id, label = label %||% "Range",
                min = 0, max = 100, value = c(0, 100), step = 0.1)
  },
  resolve_expr  = function(value, node, ...) {
    if (is.null(value) || length(value) != 2L) return(NULL)
    rlang::expr(c(!!value[1], !!value[2]))
  },
  copy_defaults = list(label = "Range for {param}")
)

ptr_define_placeholder_consumer(
  keyword       = "colvars",
  build_ui      = function(node, cols = character(), label = NULL,
                           selected = character(0), ...) {
    selectInput(node$id, label = label %||% "Columns",
                choices = cols, selected = intersect(selected, cols),
                multiple = TRUE)
  },
  resolve_expr  = function(value, node, ...) {
    if (length(value) == 0L) return(NULL)
    rlang::call2("c", !!!as.list(value))
  },
  copy_defaults = list(label = "Columns for {param}")
)

ptr_define_placeholder_value(
  keyword       = "date",
  build_ui      = function(node, label = NULL, ...) {
    dateRangeInput(node$id, label = label %||% "Date range",
                   start = "1970-01-01", end = "2015-04-01")
  },
  resolve_expr  = function(value, node, ...) {
    if (is.null(value) || length(value) != 2L) return(NULL)
    rlang::expr(c(as.Date(!!as.character(value[1])),
                  as.Date(!!as.character(value[2]))))
  },
  copy_defaults = list(label = "Date range for {param}")
)
```

The consumer's `cols` argument arrives populated with the upstream column-name vector — the runtime resolves the data context (including through any `|>` chain) before invoking `build_ui`. There is no walker to write; the framework handles it.

`resolve_expr()` returning `NULL` is the "missing" signal: the substitute pass drops the corresponding argument from the generated code, exactly like `var` returning `NULL` for an empty pick.

## 4. A realistic ggplot2 graphic, parameterized

A typical exploratory plot stacks several layers and reads several hard-coded constants. Here is a multi-layer fuel-economy chart from the `mpg` dataset:

```{r mpg-original}
ggplot(mpg, aes(displ, hwy, color = class)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_smooth(method = "loess", se = FALSE, span = 0.75) +
  facet_wrap(~ drv) +
  scale_color_brewer(palette = "Set2") +
  labs(title = "Highway MPG vs engine displacement",
       x = "Engine displacement (L)", y = "Highway MPG") +
  coord_cartesian(xlim = c(1, 7), ylim = c(10, 45))
```

Replace each tuning knob with a placeholder token. `var` becomes a column picker populated from the data; `num` becomes `numericInput`s; `text` becomes `textInput`s; the custom `range` becomes a slider:

```{r mpg-paintr}
ptr_app(
  "ggplot(data = mpg, aes(x = var, y = var, color = var)) +
     geom_point(alpha = num, size = num) +
     geom_smooth(method = 'loess', se = FALSE, span = num) +
     facet_wrap(~ var) +
     scale_color_brewer(palette = 'Set2') +
     labs(title = text, x = text, y = text) +
     coord_cartesian(xlim = range, ylim = range)"
)
```

Every aesthetic, alpha, size, span, and the two coordinate ranges is now a widget. Rebuilding the original graphic from the running app is a matter of picking those values; the code panel under the plot lifts the selections back into a static script.

## 5. Data-cleaning pipelines into `ggplot()`

`ggpaintr` parses any expression that evaluates to a `ggplot` object, including the common "clean, then plot" pipeline where the data is piped into `ggplot()`. Pipeline stages with placeholders get a per-layer **Data** sub-tab in the sidebar; column pickers further down the pipeline lazily resolve their upstream against the current Data-tab inputs, so picking a column in stage one immediately refreshes the choices for stage two without requiring an explicit "apply" click.

```{r pipe-original}
library(dplyr)

mpg |>
  dplyr::filter(displ > 1.5) |>
  dplyr::group_by(class) |>
  dplyr::filter(dplyr::n() > 5) |>
  dplyr::ungroup() |>
  ggplot(aes(displ, hwy, color = class)) +
  geom_point(alpha = 0.5)
```

```{r}
lm(expr, data = mtcars)
```


The same pipeline parameterized — the two `filter` thresholds and the point alpha become widgets:

```{r pipe-paintr}
library(dplyr)

ptr_app(
  "mpg |>
     dplyr::filter(displ > num) |>
     dplyr::group_by(class) |>
     dplyr::filter(dplyr::n() > num) |>
     dplyr::ungroup() |>
     ggplot(aes(var, var, color = class)) +
     geom_point(alpha = num) +
     labs(title = text)"
)
```

Each pipeline stage gets its own enable/disable checkbox (G11). Toggling a stage off rewrites the generated code as if that step were never there; toggle it back on and the threshold widget resurfaces. Note also that `var` (and any custom consumer like `colvars`) introspects the upstream data automatically — pipe-walking through `|>` chains is part of the runtime, not your placeholder code.

## 6. Interactive output packages

`ggpaintr` does not lock you into `renderPlot`. The runtime's `state$runtime()` reactive returns a list whose `$plot` slot is the fitted `ggplot` object, suitable for any backend that accepts a `ggplot` — `plotly` and `ggiraph` are the two most common. `ptr_extract_plot(state)` is the same accessor in non-reactive contexts.

The recipe is the same in both subsections: lay out `ptr_module_ui("id", formula)` for the sidebar + default plot panel, add a second column with the alternative output, and in the server use `state <- ptr_module_server("id", formula)` to wire everything up, then bind your custom output to `state$runtime()`.

### 6.1 plotly tooltips

Original — a hover-tooltipped scatter:

```{r plotly-original}
library(plotly)

p <- ggplot(mpg, aes(displ, hwy, color = class,
                     text = paste(manufacturer, model, sep = " "))) +
       geom_point(size = 3, alpha = 0.7) +
       coord_cartesian(xlim = c(1, 7), ylim = c(10, 45))

plotly::ggplotly(p, tooltip = "text")
```

ggpaintr version — embed via the module helpers and own the plotly output:

```{r plotly-paintr}
library(plotly)

formula <- "ggplot(data = mpg,
                   aes(x = var, y = var, color = var,
                       text = paste(manufacturer, model, sep = ' '))) +
              geom_point(size = num, alpha = num) +
              coord_cartesian(xlim = range, ylim = range)"

ui <- fluidPage(
  fluidRow(
    column(5, ptr_module_ui("plotly_demo", formula)),
    column(7, plotly::plotlyOutput("interactive_plot", height = "500px"))
  )
)

server <- function(input, output, session) {
  state <- ptr_module_server("plotly_demo", formula)
  output$interactive_plot <- plotly::renderPlotly({
    res <- state$runtime()
    shiny::req(isTRUE(res$ok), res$plot)
    plotly::ggplotly(res$plot, tooltip = "text")
  })
}

shinyApp(ui, server)
```

Reading `state$runtime()` outside `isolate()` is what wires the plotly re-render to Update Plot clicks; `req(isTRUE(res$ok), res$plot)` keeps the panel quiet between draws and on transient error states.

### 6.2 ggiraph tooltips and click handling

Original — interactive points with a colour-picked highlight:

```{r ggiraph-original}
library(ggiraph)

ggplot(mpg, aes(displ, hwy)) +
  geom_point_interactive(aes(tooltip = model), size = 3, color = "#e63946")
```

ggpaintr version with a colour-picker placeholder so the user can recolour the layer:

```{r ggiraph-paintr}
library(ggiraph)
library(colourpicker)

ptr_define_placeholder_value(
  keyword       = "color",
  build_ui      = function(node, label = NULL, ...) {
    colourpicker::colourInput(node$id, label = label %||% "Highlight colour",
                              value = "#e63946")
  },
  resolve_expr  = function(value, node, ...) {
    if (is.null(value) || identical(value, "")) return(NULL)
    as.character(value)
  },
  copy_defaults = list(label = "Highlight colour for {param}")
)

formula <- "ggplot(data = mpg, aes(x = var, y = var)) +
              geom_point_interactive(aes(tooltip = var),
                                     size = num, color = color)"

ui <- fluidPage(
  fluidRow(
    column(5, ptr_module_ui("ggiraph_demo", formula)),
    column(7, ggiraph::girafeOutput("interactive_plot", height = "500px"))
  )
)

server <- function(input, output, session) {
  state <- ptr_module_server("ggiraph_demo", formula)
  output$interactive_plot <- ggiraph::renderGirafe({
    res <- state$runtime()
    shiny::req(isTRUE(res$ok), res$plot)
    ggiraph::girafe(code = print(res$plot))
  })
}

shinyApp(ui, server)
```

## 7. ggplot2 extension packages

`ggpaintr` does not know about extension geoms — it sees them as ordinary layer calls and parameterizes their arguments the same way it parameterizes core geoms. The five examples below cover a parallel-coordinates plot, ridge densities, repelled labels, alluvial flows, and `ggdist` distribution slabs.

### 7.1 ggpcp — parallel coordinates

Original (uses the `flea` dataset shipped by `GGally`):

```{r pcp-original}
library(ggpcp)
data(flea, package = "GGally")

flea |>
  pcp_select(species, tars1, tars2, head, aede1, aede2, aede3) |>
  pcp_scale(method = "uniminmax") |>
  pcp_arrange() |>
  ggplot(aes_pcp()) +
    geom_pcp_axes() +
    geom_pcp(aes(colour = species))
```

ggpaintr version — let the user pick which columns to put on the parallel axes via the `colvars` placeholder:

```{r pcp-paintr}
library(ggpcp)
data(flea, package = "GGally")

ptr_app(
  "ggplot(data = flea |>
                  pcp_select(colvars) |>
                  pcp_scale(method = 'uniminmax') |>
                  pcp_arrange(),
          mapping = aes_pcp()) +
     geom_pcp_axes() +
     geom_pcp(aes(colour = species)) +
     coord_cartesian(xlim = range, ylim = range)"
)
```

### 7.2 ggridges — ridge densities

Original (uses `lincoln_weather` shipped by `ggridges`):

```{r ridges-original}
library(ggridges)

ggplot(lincoln_weather,
       aes(x = `Mean Temperature [F]`, y = Month, fill = ..x..)) +
  geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01,
                               bandwidth = 3) +
  scale_fill_viridis_c(option = "C")
```

ggpaintr version — bandwidth, scale, and the colour aesthetic become widgets:

```{r ridges-paintr}
library(ggridges)

ptr_app(
  "ggplot(data = lincoln_weather, aes(x = var, y = var, fill = ..x..)) +
     geom_density_ridges_gradient(scale = num,
                                  rel_min_height = num,
                                  bandwidth = num) +
     scale_fill_viridis_c(option = 'C') +
     labs(title = text)"
)
```

### 7.3 ggrepel — labelled scatter

Original — label cars by row name, repelling overlaps:

```{r repel-original}
library(ggrepel)

mt <- data.frame(mtcars, model = rownames(mtcars))

ggplot(mt, aes(wt, mpg, label = model)) +
  geom_point(size = 3, color = "steelblue") +
  geom_text_repel(size = 3.5, max.overlaps = 12, box.padding = 0.3)
```

ggpaintr version — `max.overlaps`, `size`, and `box.padding` all turn into numeric inputs. The label aesthetic is bound by `var`:

```{r repel-paintr}
library(ggrepel)

mt <- data.frame(mtcars, model = rownames(mtcars))

ptr_app(
  "ggplot(data = mt, aes(x = var, y = var, label = var)) +
     geom_point(size = num, color = 'steelblue') +
     geom_text_repel(size = num,
                     max.overlaps = num,
                     box.padding = num) +
     labs(title = text)",
  envir = environment()
)
```

`envir = environment()` is the seam that lets the runtime see the local `mt` data frame. `ptr_app()` defaults to `parent.frame()`, which would pick up `mt` here automatically; passing `environment()` is explicit.

### 7.4 ggalluvial — flow diagrams

Original (uses the `vaccinations` dataset shipped by `ggalluvial`):

```{r alluvial-original}
library(ggalluvial)

ggplot(vaccinations,
       aes(x = survey, stratum = response, alluvium = subject,
           y = freq, fill = response)) +
  geom_flow(alpha = 0.7) +
  geom_stratum() +
  scale_x_discrete(expand = c(0.1, 0.1))
```

ggpaintr version — alpha and the fill aesthetic become widgets:

```{r alluvial-paintr}
library(ggalluvial)

ptr_app(
  "ggplot(data = vaccinations,
          aes(x = survey, stratum = response, alluvium = subject,
              y = freq, fill = var)) +
     geom_flow(alpha = num) +
     geom_stratum() +
     scale_x_discrete(expand = c(0.1, 0.1)) +
     labs(title = text)"
)
```

### 7.5 ggdist — distribution visualizations

Original — half-eye distributions per cylinder count:

```{r dist-original}
library(ggdist)

ggplot(mtcars, aes(x = factor(cyl), y = mpg)) +
  stat_halfeye(adjust = 0.5, justification = -0.2,
               .width = 0, point_colour = NA, slab_alpha = 0.6) +
  geom_boxplot(width = 0.15, outlier.shape = NA) +
  coord_flip()
```

ggpaintr version — every visual constant of `stat_halfeye()` becomes a slider or numeric input:

```{r dist-paintr}
library(ggdist)

ptr_app(
  "ggplot(data = mtcars, aes(x = factor(cyl), y = var)) +
     stat_halfeye(adjust = num,
                  justification = num,
                  .width = num,
                  point_colour = NA,
                  slab_alpha = num) +
     geom_boxplot(width = num, outlier.shape = NA) +
     coord_flip() +
     labs(title = text)"
)
```

## 8. Shared placeholders — one widget, two graphics

`ptr_app_grid()` is the turn-key entry point for multi-plot apps with shared widgets. Annotate any placeholder occurrence with `keyword(shared = "<id>")`; pass `shared_ui = list(<id> = function(id) widget_tag)` so the framework knows what tag to render at the top. The `<id>` namespace is shared across every formula in `plots`, so editing the top widget propagates to every plot.

Original — two static views of the iris dataset coupled by a shared column choice:

```{r shared-original}
library(dplyr)
library(tidyr)
library(ggpcp)

picked <- c("Sepal.Length", "Sepal.Width", "Petal.Length")

p1 <- iris |>
  pcp_select(Species, dplyr::all_of(picked)) |>
  pcp_scale(method = "uniminmax") |>
  pcp_arrange() |>
  ggplot(aes_pcp()) +
    geom_pcp_axes() +
    geom_pcp(aes(colour = Species))

p2 <- iris |>
  dplyr::select(Species, dplyr::all_of(picked)) |>
  tidyr::pivot_longer(-Species, names_to = "metric") |>
  ggplot(aes(metric, value, fill = Species)) +
    geom_boxplot()

# Display side-by-side with patchwork::wrap_plots(p1, p2) or similar.
```

ggpaintr version — one shared `selectInput` at the top of the grid, two formulas referencing `colvars(shared = "vars")`. Picking columns in the shared control updates both plots together when you click **Draw all**:

```{r shared-paintr}
library(ggpcp)

ptr_app_grid(
  plots = list(
    "ggplot(data = iris |>
                    pcp_select(Species, colvars(shared = 'vars')) |>
                    pcp_scale(method = 'uniminmax') |>
                    pcp_arrange(),
            mapping = aes_pcp()) +
       geom_pcp_axes() +
       geom_pcp(aes(colour = Species))",

    "ggplot(data = iris |>
                    dplyr::select(Species, colvars(shared = 'vars')) |>
                    tidyr::pivot_longer(-Species, names_to = 'metric'),
            aes(x = metric, y = value, fill = Species)) +
       geom_boxplot()"
  ),
  shared_ui = list(
    vars = function(id) {
      selectInput(id, "Columns to plot",
                  choices  = setdiff(names(iris), "Species"),
                  selected = c("Sepal.Length", "Sepal.Width", "Petal.Length"),
                  multiple = TRUE)
    }
  )
)
```

The shared-id mechanism works for any placeholder keyword — built-in `num(shared = "sz")` paired with `shared_ui = list(sz = function(id) sliderInput(...))` is the simplest two-line example; `colvars(shared = "vars")` above shows the same pattern for a custom consumer.

## 9. Programmatic extras and headless extraction

`ptr_gg_extra(state, ...)` attaches scales, themes, or any expression that combines with `+ <plot>` to the rendered output without changing the formula. Each call replaces the previously captured extras; `ptr_extract_code(state)` includes them in the printable code.

```{r gg-extra}
formula <- "ggplot(mtcars, aes(x = mpg, y = hp)) + geom_point()"

ui <- fluidPage(
  actionButton("add_log", "Toggle log-scale"),
  ptr_module_ui("p", formula)
)
server <- function(input, output, session) {
  state <- ptr_module_server("p", formula)
  observeEvent(input$add_log, {
    ptr_gg_extra(state, ggplot2::scale_x_log10())
  })
}

shinyApp(ui, server)
```

For headless use — composing custom UIs, snapshot tests, or pre-flight rendering — drive `ptr_server()` through `shiny::testServer()` and read the accessors directly:

```{r headless}
shiny::testServer(function(input, output, session) {
  state <- ptr_server(
    input, output, session,
    "ggplot(mtcars, aes(x = mpg, y = hp)) + geom_point()"
  )
  session$setInputs(.dummy = 1)
  session$userData$paintr <- state
}, {
  cat("CODE:\n", ptr_extract_code(session$userData$paintr), "\n")
  cat("PLOT class:", class(ptr_extract_plot(session$userData$paintr)), "\n")
})
```

## 10. Where to go next

If a section above introduced an unfamiliar piece of the API, the canonical references are:

- `?ptr_app`, `?ptr_module_ui`, `?ptr_module_server`, `?ptr_app_grid` — entry points.
- `?ptr_define_placeholder_value`, `?ptr_define_placeholder_consumer`, `?ptr_define_placeholder_source` — the three constructors that extend the registry.
- `?ptr_extract_plot`, `?ptr_extract_code`, `?ptr_extract_error`, `?ptr_gg_extra` — the embed-time accessors used by the plotly / ggiraph / extras sections.
- `dev/scripts/feature-sweep.R` — the developer feature sweep, which exercises every join-point above plus error paths and the denylist.


```{r}
library(tidyverse)

# Pipeline Example
iris %>%
  filter(Sepal.Length > 5) %>%
  mutate(Sepal.Area = Sepal.Length * Sepal.Width) %>%
  ggplot(aes(x = Species, y = Sepal.Area, fill = Species)) +
  geom_boxplot() +
  labs(title = "Piped Data Processing to Plot",
       y = "Sepal Area (Length * Width)")


ptr_app(
  'iris %>%
  filter(Sepal.Length > num) %>%
  mutate(new_var = var * var) %>%
  ggplot(aes(x = var, y = var, fill = var)) +
  geom_boxplot() +
  labs(title = "Piped Data Processing to Plot",
       y = "Sepal Area (Length * Width)")
  '
)
```



