---
title: "ggpaintr Placeholder Registry"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ggpaintr Placeholder Registry}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)

library(ggpaintr)
library(ggplot2)
library(rlang)
```

## Overview

`ggpaintr` now supports per-app placeholder registries. That means downstream
developers can define a new placeholder once, pass it into
`paintr_formula()`, `ggpaintr_app()`, `ggpaintr_server_state()`, or
`generate_shiny()`, and keep the rest of the package runtime unchanged.

The public entry points for this layer are:

- `ggpaintr_placeholder()`
- `ggpaintr_effective_placeholders()`
- `ggpaintr_missing_expr()`

Built-in placeholders such as `var`, `text`, `num`, `expr`, and `upload` now
use the same registry path as custom placeholders.

## A minimal custom placeholder

This `date` placeholder adds a Shiny `dateInput()` control and turns the chosen
date into `as.Date("YYYY-MM-DD")` in generated code.

```{r date-placeholder}
sales <- data.frame(
  day = as.Date("2024-01-01") + 0:4,
  value = c(10, 13, 12, 16, 18)
)

date_placeholder <- ggpaintr_placeholder(
  keyword = "date",
  build_ui = function(id, copy, meta, context) {
    shiny::dateInput(id, copy$label)
  },
  resolve_expr = function(value, meta, context) {
    if (is.null(value) || identical(as.character(value), "")) {
      return(ggpaintr_missing_expr())
    }

    rlang::expr(as.Date(!!as.character(value)))
  },
  copy_defaults = list(label = "Choose a date for {param}")
)

placeholders <- ggpaintr_effective_placeholders(
  list(date = date_placeholder)
)

names(placeholders)
```

## Using the registry in parsing and runtime

Pass the registry through `placeholders = ...` anywhere you want custom
placeholder support.

```{r date-runtime}
obj <- paintr_formula(
  "ggplot(data = sales, aes(x = day, y = value)) +
    geom_line() +
    geom_vline(xintercept = date)",
  placeholders = placeholders
)

runtime <- paintr_build_runtime(
  obj,
  list(
    "geom_line+checkbox" = TRUE,
    "geom_vline+checkbox" = TRUE,
    "geom_vline+2" = as.Date("2024-01-03")
  )
)

runtime$code_text
```

The same registry can also be used in the higher-level wrappers.

```{r date-app, eval = FALSE}
ggpaintr_app(
  "ggplot(data = sales, aes(x = day, y = value)) +
    geom_line() +
    geom_vline(xintercept = date)",
  placeholders = placeholders
)
```

## The placeholder metadata contract

Every placeholder occurrence becomes one metadata record in
`paintr_obj$placeholder_map`.

```{r meta-record}
meta <- obj$placeholder_map$geom_vline[["geom_vline+2"]]
meta
```

Each metadata record contains:

- `id`: the Shiny input id for that placeholder occurrence
- `keyword`: the placeholder keyword, such as `date`
- `layer_name`: the layer where the placeholder was found
- `param`: the argument name, such as `xintercept`
- `index_path`: the expression path used during replacement

These fields are passed into every placeholder hook so custom placeholder code
can behave differently by parameter or by layer.

## The hook contract

`ggpaintr_placeholder()` supports one required UI hook, one required resolver
hook, and three optional advanced hooks.

### `build_ui(id, copy, meta, context)`

`build_ui()` returns the UI control for one placeholder occurrence.

```{r build-ui-example}
date_placeholder$build_ui
```

The arguments are:

- `id`: the generated input id to bind
- `copy`: resolved copy text for this control
- `meta`: the metadata record for this occurrence
- `context`: a named list containing `paintr_obj`, `placeholders`,
  `copy_rules`, and `envir`

### `resolve_expr(value, meta, context)`

`resolve_expr()` turns the current input value into an expression suitable for
the completed plot expression.

```{r resolve-expr-example}
date_placeholder$resolve_expr
```

Return `ggpaintr_missing_expr()` when the argument should be removed from the
completed code instead of inserted.

### `resolve_input(input, id, meta, context)`

Use `resolve_input()` when the raw Shiny input needs preprocessing before it is
handed to `resolve_expr()`. If omitted, `ggpaintr` uses `input[[id]]`.

```{r resolve-input-example, eval = FALSE}
trimmed_text_placeholder <- ggpaintr_placeholder(
  keyword = "trimmed_text",
  build_ui = function(id, copy, meta, context) {
    shiny::textInput(id, copy$label)
  },
  resolve_input = function(input, id, meta, context) {
    trimws(input[[id]])
  },
  resolve_expr = function(value, meta, context) {
    if (identical(value, "")) {
      return(ggpaintr_missing_expr())
    }

    rlang::expr(!!value)
  }
)
```

### `bind_ui(input, output, metas, context)`

Use `bind_ui()` for deferred or dynamic UI registration. This is how the built-in
`var` placeholder registers delayed `renderUI()` outputs after data is known.

```{r bind-ui-example, eval = FALSE}
deferred_placeholder <- ggpaintr_placeholder(
  keyword = "dynamic_text",
  build_ui = function(id, copy, meta, context) {
    shiny::uiOutput(paste0("dynamic-", id))
  },
  bind_ui = function(input, output, metas, context) {
    for (meta in metas) {
      output[[paste0("dynamic-", meta$id)]] <- shiny::renderUI({
        shiny::textInput(meta$id, paste("Dynamic control for", meta$param))
      })
    }

    invisible(NULL)
  },
  resolve_expr = function(value, meta, context) {
    if (is.null(value) || identical(value, "")) {
      return(ggpaintr_missing_expr())
    }

    rlang::expr(!!value)
  }
)
```

### `prepare_eval_env(input, metas, eval_env, context)`

Use `prepare_eval_env()` when a placeholder needs to inject objects into the
evaluation environment before the plot is built. This is how the built-in
`upload` placeholder assigns uploaded data objects before code is evaluated.

```{r prepare-eval-env-example, eval = FALSE}
prepared_data_placeholder <- ggpaintr_placeholder(
  keyword = "prepared_data",
  build_ui = function(id, copy, meta, context) {
    shiny::textInput(id, copy$label, placeholder = "dataset_name")
  },
  resolve_expr = function(value, meta, context) {
    if (is.null(value) || identical(value, "")) {
      return(ggpaintr_missing_expr())
    }

    rlang::parse_expr(value)
  },
  prepare_eval_env = function(input, metas, eval_env, context) {
    eval_env$prepared_dataset <- data.frame(x = 1:3, y = c(2, 4, 3))
    eval_env
  }
)
```

## Copy defaults and custom copy rules

Each placeholder can ship its own default copy through `copy_defaults`. Those
defaults then participate in the normal `copy_rules` system.

```{r custom-copy}
date_copy_rules <- paintr_effective_copy_rules(
  list(
    defaults = list(date = list(label = "Pick any date")),
    params = list(
      xintercept = list(date = list(label = "Reference date"))
    ),
    layers = list(
      geom_vline = list(
        date = list(
          xintercept = list(help = "Choose a cutoff date for the vertical guide.")
        )
      )
    )
  ),
  placeholders = placeholders
)

paintr_resolve_copy(
  "control",
  keyword = "date",
  layer_name = "geom_vline",
  param = "xintercept",
  copy_rules = date_copy_rules,
  placeholders = placeholders
)
```

## Using custom placeholders with Shiny integration helpers

The phase-1 Shiny integration helpers work with custom placeholder registries
too.

```{r registry-integration, eval = FALSE}
ids <- ggpaintr_ids(
  control_panel = "builder_controls",
  draw_button = "render_plot",
  export_button = "export_app",
  plot_output = "main_plot",
  error_output = "main_error",
  code_output = "main_code"
)

ui <- shiny::fluidPage(
  shiny::sidebarLayout(
    shiny::sidebarPanel(
      ggpaintr_controls_ui(ids = ids)
    ),
    shiny::mainPanel(
      ggpaintr_outputs_ui(ids = ids)
    )
  )
)

server <- function(input, output, session) {
  paintr_state <- ggpaintr_server_state(
    "ggplot(data = sales, aes(x = day, y = value)) +
      geom_line() +
      geom_vline(xintercept = date)",
    ids = ids,
    placeholders = placeholders
  )

  ggpaintr_bind_control_panel(input, output, paintr_state, ids = ids)
  ggpaintr_bind_draw(input, paintr_state, ids = ids)
  ggpaintr_bind_export(output, paintr_state, ids = ids)
  ggpaintr_bind_plot(output, paintr_state, ids = ids)
  ggpaintr_bind_error(output, paintr_state, ids = ids)
  ggpaintr_bind_code(output, paintr_state, ids = ids)
}
```

## Exporting a standalone app with custom placeholders

Exported apps rebuild only the custom placeholder definitions they need. The
same placeholder registry can be passed into `generate_shiny()`.

```{r export-registry, eval = FALSE}
out_file <- tempfile(fileext = ".R")

generate_shiny(
  obj,
  out_file,
  style = FALSE,
  placeholders = placeholders
)
```

For an exported app to stay standalone, custom placeholders must be created with
`ggpaintr_placeholder()` and must define hook functions inline inside that call.
If a placeholder definition cannot be serialized safely, `generate_shiny()` will
error instead of writing a broken app.

## What exportable means

The current export path works by storing the original
`ggpaintr_placeholder(...)` call on each custom placeholder as
`definition_call`. When you export an app, `generate_shiny()` serializes that
stored call into the generated script and then rebuilds the registry there.

That is why inline hooks are exportable today:

- inline hook bodies live inside the stored `ggpaintr_placeholder(...)` call
- external helper references are only symbols, so the exported app would not
  contain the helper definitions themselves

You can inspect the stored call directly:

```{r definition-call}
date_placeholder$definition_call
```

### Exportable today

```{r exportable-inline, eval = FALSE}
date_placeholder <- ggpaintr_placeholder(
  keyword = "date",
  build_ui = function(id, copy, meta, context) {
    shiny::dateInput(id, copy$label)
  },
  resolve_expr = function(value, meta, context) {
    if (is.null(value) || identical(as.character(value), "")) {
      return(ggpaintr_missing_expr())
    }

    rlang::expr(as.Date(!!as.character(value)))
  }
)
```

### Not exportable today

```{r non-exportable-inline, eval = FALSE}
build_date_ui <- function(id, copy, meta, context) {
  shiny::dateInput(id, copy$label)
}

date_to_expr <- function(value, meta, context) {
  if (is.null(value) || identical(as.character(value), "")) {
    return(ggpaintr_missing_expr())
  }

  rlang::expr(as.Date(!!as.character(value)))
}

date_placeholder <- ggpaintr_placeholder(
  keyword = "date",
  build_ui = build_date_ui,
  resolve_expr = date_to_expr
)

# generate_shiny(..., placeholders = list(date = date_placeholder))
# errors because the exported app would not contain build_date_ui() or
# date_to_expr().
```

## Summary

The placeholder registry is the developer-facing extension point for adding new
control types without editing parser, UI, runtime, copy-rule, or export
internals. Start with `build_ui()` plus `resolve_expr()`, add the optional hooks
only when needed, and pass the same registry through parsing, app launch, Shiny
integration, and export.
