---
title: "Use cases"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Use cases}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment   = "#>",
  warning   = FALSE,
  message   = FALSE,
  eval      = interactive()
)
```

This vignette answers the question *how do I use ggpaintr for my situation?* Companion vignettes cover tailoring defaults (`vignette("ggpaintr-customization")`), sharing safely (`vignette("ggpaintr-safety")`), real ggplot recipes (`vignette("ggpaintr-gallery")`), and LLM integration (`vignette("ggpaintr-llm")`).

Every chunk is marked `eval = interactive()` and is runnable as-is at the R prompt after:

```{r libs}
library(ggpaintr)
```

## Formula tour

A ggpaintr formula is a single `ggplot()` call written as text. Drop one of five placeholder keywords anywhere a value would normally go, and the runtime substitutes the user's input back into the expression at render time. The five keywords are:

| Keyword  | Becomes                                       | Example formula fragment        | Renders as                                       |
|----------|-----------------------------------------------|---------------------------------|--------------------------------------------------|
| `var`    | column picker (data-aware `selectInput`)      | `aes(x = var, y = var)`         | two dropdowns scoped to the upstream data's columns |
| `text`   | free-text input                               | `labs(title = text)`            | one text field                                   |
| `num`    | numeric input                                 | `geom_point(size = num)`        | one numeric input                                |
| `expr`   | code editor (validated against a denylist)    | `facet_wrap(expr)`              | one expression box; see `vignette("ggpaintr-safety")` for the validation model |
| `upload` | file picker (returns a data frame)            | `ggplot(upload, …)`             | one file picker + an optional dataset-name field |

A formula that exercises four of the five:

```{r formula-tour}
ptr_app("
ggplot(iris, aes(x = var, y = var, color = var)) +
  geom_point(size = num) +
  labs(title = text) +
  facet_wrap(expr)
")
```

The full grammar reference — including positional-argument behavior, `shared = "<id>"` annotation, and the empty-call cleanup rule — lives at `?ptr_app`.

## Data pipelines in the formula

A layer's data argument is allowed to be a pipeline. Both `|>` and `%>%` are accepted, and the two can be mixed in the same chain. Each pipeline stage that contains a placeholder gets its own row of widgets under the layer's **Data** subtab, plus a per-stage enable/disable toggle that rewrites the generated code as if that step were never there. Column pickers downstream of a stage lazily re-resolve their column choices against the current Data-subtab inputs, so changing a filter threshold immediately refreshes the picker below it.

```{r pipeline-formula-tour}
ptr_app(
  "mpg |>
     dplyr::filter(displ > num) |>
     dplyr::group_by(class) |>
     dplyr::filter(dplyr::n() > num) |>
     dplyr::ungroup() |>
     ggplot(aes(var, var, color = class)) +
     geom_point(alpha = num)"
)
```

See `vignette(\"ggpaintr-gallery\")` § 4 for the same example with the original `dplyr` pipeline alongside the parameterized version.

### What we expect to work

- **Both pipe operators**, possibly mixed in the same chain: `df %>% filter(...) |> ggplot(...)`.
- **Bare-symbol stage heads**: `filter(...)`, `head(...)`, `transform(...)`.
- **Namespaced stage heads**: `dplyr::filter(...)`, `tidyr:::pivot_longer(...)`.
- **Parenthesised heads**: `(filter)(...)` — used occasionally to disambiguate function lookup.
- **`upload` as the pipeline head**: `upload |> dplyr::filter(...) |> ggplot(...)`. The upload widget feeds the pipeline; every downstream stage and placeholder resolves against the uploaded data.
- **Per-layer pipelines on multi-layer formulas**: each layer can have its own data-arg pipeline (`ggplot(data = df1 |> filter(...)) + geom_smooth(data = df2 |> filter(...))`) — the Data subtabs are scoped per layer.
- **Placeholders in any stage**: `var`, `text`, `num`, `expr` inside any stage's call resolve as if that stage were a regular layer call.

### What we know does **not** work yet

- **Anonymous functions as pipe stages.** `df |> (\(x) x[x$cyl > 4, ])() |> ggplot(...)` and `df |> (function(x) ...)() |> ggplot(...)` are not supported. The runtime resolves each stage's name from its call head; an inline lambda has no name, so the stage cannot be assigned a stable Shiny id or a layer label, and the Data subtab cannot render. The expression-level safety walker still inspects the lambda body, but the surrounding pipeline machinery does not. Workaround: lift the anonymous function to a named helper in the calling environment, then use it as a bare-symbol stage (`df |> my_helper() |> ggplot(...)`).

## L1 — All-in-one entry point

L1 is the dominant use case: hand ggpaintr a formula, get back a running Shiny app. Two entry points cover the common layouts.

| Entry point        | Use when                                                                 |
|--------------------|--------------------------------------------------------------------------|
| `ptr_app()`        | Default layout. The built-in chrome (title, layer picker, plot pane, code pane) is enough. |
| `ptr_app_grid()`   | Multiple formulas, one app — each plot gets its own tile, optionally with shared widgets at the top. |

For a different page shell or theme, write a thin wrapper on top of the public ggpaintr primitives — see `vignette("ggpaintr-customization")` § "Writing your own wrapper" for the recipe and a worked example.

### Single plot — `ptr_app()`

```{r app-basic}
ptr_app(
  "ggplot(iris, aes(var, var, color = var)) + geom_point() + labs(title = text)"
)
```

### Multiple plots and shared widgets — `ptr_app_grid()`

`ptr_app_grid()` is the turn-key entry point for multi-plot apps with shared widgets. It takes a list of formulas in `plots =` and renders one tile per formula. Annotate any placeholder occurrence with `keyword(shared = "<id>")`; pass `shared_ui = list(<id> = function(id) widget_tag)` so the framework knows what tag to render at the top of the page. The `<id>` namespace is shared across every formula in `plots`, so editing the top widget propagates to every plot.

```{r app-grid-shared}
ptr_app_grid(
  plots = list(
    "ggplot(iris, aes(x = var, y = Sepal.Length, fill = Species)) +
       geom_boxplot()",
    "ggplot(iris, aes(x = var, y = Sepal.Width, fill = Species)) +
       geom_violin()"
  ),
  shared_ui = list()  # no shared= annotations above; the first var stays per-tile
)
```

Adding `shared = 'metric'` to both `var` occurrences and supplying `shared_ui = list(metric = function(id) selectInput(id, "Metric", names(iris)))` lifts the column choice into a single widget that drives both plots when the user clicks **Draw all**. The shared mechanism works for any keyword — built-in `num(shared = "sz")` paired with `shared_ui = list(sz = function(id) sliderInput(id, "Size", 1, 10, 3))` is the simplest two-line example; custom consumers work the same way (see `vignette("ggpaintr-customization")` for registering custom widget types).

```{r app-grid-shared-added}
ptr_app_grid(
  plots = list(
    "ggplot(iris, aes(x = var(shared = 'metric'), y = Sepal.Length, fill = Species)) +
       geom_boxplot()",
    "ggplot(iris, aes(x = var(shared = 'metric'), y = Sepal.Width, fill = Species)) +
       geom_violin()"
  ),
  shared_ui = list(metric = function(id) selectInput(id, "Metric", names(iris)))
)
```

### Empty-call cleanup

A placeholder that resolves to "missing" (an empty `var` pick, a blank `text`, a cleared `num`) drops its argument from the generated code. When that leaves an empty call whose name is in ggpaintr's curated cleanup list, the whole call disappears too:

```r
ptr_app("
ggplot(iris, aes(x = var, y = var)) +
  geom_point() +
  labs(title = text) +              # title empty → labs() empty → labs() dropped
  facet_wrap(expr)                  # expr empty → facet_wrap() empty → facet_wrap() dropped
")
```

The cleanup list covers safe ggplot2 no-ops: `theme()`, `labs()`, `xlab` / `ylab` / `ggtitle`, `facet_wrap` / `facet_grid` / `facet_null`, `xlim` / `ylim` / `lims`, `expand_limits`, `guides`, `annotate`, and the empty mapping helpers (`aes()`, `vars()`). Third-party helpers (e.g. `pcp_theme()` from `ggpcp`) are *not* in the cleanup list — being absent is the "removal safety unknown" signal. Pass `safe_to_remove = c("pcp_theme")` to opt a specific name into the cleanup pass. An `expr` placeholder whose user supplies an expression always wins: whatever was typed is honoured verbatim, even if its top-level name is in `safe_to_remove`.

### Local data with non-syntactic columns

`var` placeholders need syntactic, unique column names — `Sepal.Length` is fine, `"Sepal Length"` is not, and `iris` deliberately ships clean. If your local data has spaces, reserved words, or duplicates in its names, pipe it through `ptr_normalize_column_names()` before referencing it from a formula:

```{r normalize-cols}
messy <- data.frame(
  check.names = FALSE,
  "first column"  = 1:3,
  "if"            = 4:6
)
clean <- ptr_normalize_column_names(messy)
names(clean)
#> "first_column" "if_"

ptr_app("ggplot(clean, aes(x = var, y = var)) + geom_point()")
```

Uploaded data goes through the same normalization automatically; the manual call is only needed for data already in your R session. `?ptr_normalize_column_names` documents the exact rules (`make.names()` plus uniqueness + reserved-word collision avoidance).

## L2 — Embed in your own Shiny app

L2 is "I already have a Shiny app and want to drop a ggpaintr-driven block somewhere inside it." Two paths cover the common cases: a Shiny module (one id, two helpers) or a free-form split (UI + outputs placed wherever you want).

### Module path — `ptr_module_ui()` / `ptr_module_server()`

The compact option. One `id`, controls + outputs live next to each other inside the module's namespace.

```{r module-app}
formula <- "ggplot(iris, aes(var, var, color = var)) + geom_point()"

ui <- shiny::fluidPage(
  shiny::titlePanel("My host app"),
  ptr_module_ui("p", formula)
)
server <- function(input, output, session) {
  ptr_module_server("p", formula)
}

shiny::shinyApp(ui, server)
```

`ptr_module_server()` returns the `ptr_state` from `ptr_init_state()` so embedders that want to react to it (e.g. call `ptr_gg_extra()`, read `ptr_extract_*()`) can capture it; otherwise the side effects are all that matter. `...` is forwarded to `ptr_init_state()` — so `ui_text =`, `checkbox_defaults =`, `expr_check =`, `safe_to_remove =`, and `shared =` are all available here too.

### Non-module path — `ptr_controls_ui()` + `ptr_outputs_ui()` + `ptr_server()`

When you want the controls in one part of the page and the plot in another, split the UI:

```{r split-app}
formula <- "ggplot(iris, aes(var, var, color = var)) + geom_point()"

ui <- shiny::fluidPage(
  shiny::titlePanel("Custom layout"),
  shiny::sidebarLayout(
    shiny::sidebarPanel(
      ptr_controls_ui(formula = formula)   # widgets only
    ),
    shiny::mainPanel(
      shiny::tags$h3("Plot + code + errors"),
      ptr_outputs_ui()                     # plot / error / code outputs
    )
  )
)
server <- function(input, output, session) {
  ptr_server(input, output, session, formula)
}

shiny::shinyApp(ui, server)
```

`ptr_controls_ui()` and `ptr_outputs_ui()` are the two halves that `ptr_module_ui()` renders together; each one self-wraps in a `<div class="ptr-app">` so the bundled stylesheet attaches without any host-side scaffolding. `ptr_server()` registers `ptr_register_plot()` / `_error()` / `_code()` for you — `ptr_outputs_ui()` reserves the three canonical output slots (`ptr_plot`, `ptr_error`, `ptr_code`) that those registrations target.

If you need to embed more than one `ggpaintr` formula on the same page — or share a sidebar with hand-written widgets that might collide — pass a matching `id` to all three: `ptr_controls_ui("p", formula)`, `ptr_outputs_ui("p")`, and `shiny::moduleServer("p", function(input, output, session) ptr_server(input, output, session, formula))`. The single-formula case (the example above) needs neither the `id` nor the `moduleServer()` wrapper.

### Splitting outputs further

`ptr_register_plot()`, `ptr_register_error()`, and `ptr_register_code()` are exposed for embedders who want fewer than three outputs (drop the code pane, for example), or who want to drive the outputs from their own `ptr_init_state()` without `ptr_server()`'s opinionated bundle. Each accepts `(output, state)` and writes to a canonical, namespaced output id; pair with `plotOutput()` / `uiOutput()` / `verbatimTextOutput()` at the namespaced output id.

If you need custom output names — your own `plotOutput("my_chart")` rather than ggpaintr's `ptr_plot` — read the latest plot from the returned `state` inside your own `renderPlot()` block; see *Custom rendering off the returned `state`* below (it is an L2 capability, since the server functions return `state`). If instead you want to own the *placement and markup* of each pane (plot / error / code / controls) while keeping ggpaintr's renderer, that is **L3** — one exported function per UI piece; see *L3 — Own every piece of the UI*.

### Shared widgets across embedded plots — `ptr_shared_ui()` / `ptr_shared_server()`

`ptr_app_grid()` (L1) builds a shared-widget panel for you. When you embed two or more ggpaintr blocks in your *own* Shiny app and want one widget to drive all of them, wire that panel by hand with the same two primitives `ptr_app_grid()` uses internally: `ptr_shared_ui()` for the panel tag and `ptr_shared_server()` for the linked reactives. Annotate the placeholder with `keyword(shared = "<id>")` in every formula, exactly as in the L1 grid example.

The contract:

- `ptr_shared_ui(formulas, shared_ui = list(), ...)` — pass the **same list of formula strings** you will hand to the modules. Returns one wrapped panel tag; place it wherever you want on the page. `shared_ui` maps each `<id>` to a `function(id) -> shiny.tag`, identical to the `ptr_app_grid()` argument; a `var(shared = "...")` consumer needs no entry (its picker is built for you).
- `ptr_shared_server(formulas, ...)` — call inside your server function. Returns a `ptr_shared_state` bundle carrying `shared`, `draw_trigger`, `shared_resolutions`, and `shared_stage_enabled`.
- `ptr_module_server(id, formula, shared_state = <bundle>)` — pass the bundle to every per-plot module. `shared_state =` is the convenience path; its slots are also accepted individually through `...`, and explicit `...` entries win.
- `ptr_server(input, output, session, formula, shared_state = <bundle>)` — the non-module equivalent. `ptr_server()` accepts the same `shared_state =` bundle, so a single-plot embed or a split/L3 layout can consume the shared panel without a `moduleServer()` wrapper.

```{r l2-shared}
plots <- list(
  "ggplot(iris, aes(x = var(shared = 'metric'), y = Sepal.Length, fill = Species)) + geom_boxplot()",
  "ggplot(iris, aes(x = var(shared = 'metric'), y = Sepal.Width,  fill = Species)) + geom_violin()"
)

ui <- shiny::fluidPage(
  shiny::titlePanel("My host app"),
  ptr_shared_ui(
    formulas  = plots,
    shared_ui = list(metric = function(id) shiny::selectInput(id, "Metric", names(iris)))
  ),
  shiny::fluidRow(
    shiny::column(6, ptr_module_ui("plot_1", plots[[1]])),
    shiny::column(6, ptr_module_ui("plot_2", plots[[2]]))
  )
)
server <- function(input, output, session) {
  shared <- ptr_shared_server(formulas = plots)
  ptr_module_server("plot_1", plots[[1]], shared_state = shared)
  ptr_module_server("plot_2", plots[[2]], shared_state = shared)
}

shiny::shinyApp(ui, server)
```

`ptr_shared_ui()` emits a **Draw all** button (top-level input id `ptr_shared_draw_all`) whenever `formulas` has two or more entries; the per-module **Update plot** buttons still work independently. Calling either primitive with a `formulas` list that declares no `shared = "..."` annotation is an error — drop the panel and use the plain L2 embed if nothing is shared.

For a **custom renderer** with shared widgets, keep this exact wiring and additionally capture each module's returned state — `state1 <- ptr_module_server("plot_1", plots[[1]], shared_state = shared)` — then read `state1$runtime()` inside your own `renderPlotly()` / `renderGirafe()` block, as in *Shared widgets driving custom renderers* below. The shared panel feeds the modules; the modules feed your custom renderers.

### `ptr_init_state()` and the advanced runtime arguments

`ptr_init_state(formula, ...)` is the state constructor that `ptr_server()` and `ptr_module_server()` call under the hood. It builds the `ptr_state` object but does **not** wire any inputs or start the runtime: `state$runtime()` stays `NULL` until `ptr_server()` installs the input/runtime observers (those `ptr_setup_*` steps are internal and not exported). So you do not drive a plot from `ptr_init_state()` alone — reach its advanced arguments by passing them through the `...` of `ptr_server()` / `ptr_module_server()`, which forward verbatim to it:

```{r init-state}
server <- function(input, output, session) {
  state <- ptr_server(
    input, output, session,
    formula = "ggplot(iris, aes(var, var)) + geom_point()",
    producer_debounce_ms = 300,                       # advanced ptr_init_state() arg, threaded through
    draw_trigger = shiny::reactive(input$my_button)
  )
  # state$runtime() / ptr_extract_*(state) are now usable; render your own panes off it.
}
```

`?ptr_init_state` documents every advanced argument: `shared` (host-level shared bindings), `shared_resolutions` / `shared_stage_enabled` (the remaining `ptr_shared_state` slots — see "Shared widgets across embedded plots" above), `draw_trigger` (custom redraw reactive), `producer_debounce_ms` (data-source debouncing), and the `ns` / `server_ns` namespace controls. All of them are accepted by `ptr_server()` and `ptr_module_server()` through `...`.

### How input ids are built — and how to avoid colliding with them

The framework names every widget it emits from the formula's syntax, so the same formula always produces the same ids. When you mix ggpaintr UI with your own widgets in the same Shiny namespace, the cleanest fix is to wrap the embed in `ptr_module_ui()` / `ptr_module_server()` with a unique `id` — everything ggpaintr emits then lives under `<id>-…` and your top-level inputs cannot collide. If you can't namespace, the surface to avoid is:

**Layer names.** Each top-level call in the formula contributes a layer name equal to the function being called (`ggplot`, `geom_point`, `facet_wrap`, …). Repeats get `-2`, `-3`, … suffixes:

| Formula fragment                          | Layer names                      |
|-------------------------------------------|----------------------------------|
| `geom_point() + geom_smooth()`            | `geom_point`, `geom_smooth`      |
| `geom_point() + geom_point()`             | `geom_point`, `geom_point-2`     |

**Per-placeholder input ids** follow `<layer>_<path>_<keyword>_<shared-or-NA>`, where `<path>` is the underscore-joined positional index path into the call. For example, `aes(x = var, y = var)` inside the `ggplot()` layer yields `ggplot_1_1_var_NA` and `ggplot_1_2_var_NA`; a shared annotation `var(shared = "x_col")` lands at `ggplot_1_1_var_x_col`.

**Layer-derived ids:**

| Pattern                                          | What it is                                                  |
|--------------------------------------------------|-------------------------------------------------------------|
| `<layer>_checkbox`                               | "include this layer" toggle (no checkbox on `ggplot`)       |
| `<layer>_<path>_stage_enabled`                   | per-stage toggle on `%>%` / `|>` data-arg pipelines         |
| `<placeholder-id>_ui`                            | `renderUI` container for a consumer placeholder             |
| `<layer>_subtab`, `ptr_layer_content_<layer>`    | internal layer-nav and panel-container ids                  |

**Top-level package-owned ids.** The framework writes to a fixed set of `ptr_`-prefixed top-level ids and they are **not** user-configurable: `ptr_plot` (plotOutput), `ptr_error` (uiOutput), `ptr_code` (verbatimTextOutput), `ptr_update_plot` (draw button), plus internal layer-nav inputs `ptr_layer_select` / `ptr_layer_tabset`. Treat the whole `ptr_` prefix as reserved.

> Rule of thumb: in a host app's top-level namespace, don't author Shiny inputs or outputs whose ids start with `ptr_`, or match `<layer-name-from-your-formula>_…`. If that feels brittle, wrap the embed in `ptr_module_ui()` / `ptr_module_server()` with a unique `id` and the collision surface disappears entirely.

### Custom rendering off the returned `state` (advanced L2)

"I want my own renderer." This was previously billed as a separate level, but it needs nothing beyond L2: both `ptr_module_server()` and `ptr_server()` **return** the `ptr_state`, so the moment you are embedding you can already swap ggpaintr's renderer for whatever Plotly / ggiraph / custom host you prefer. The runtime stays inside ggpaintr; you replace only the *output*. (Owning the *markup and placement* of each built-in pane is the separate **L3** story — see *L3 — Own every piece of the UI* below.)

There are two reading paths off a `ptr_state`, and the choice depends on whether you are inside a reactive context:

| Inside a `renderX({...})` / `reactive({...})` / `observeEvent({...})` | Outside reactive contexts (tests, download handlers, snapshots) |
|----|----|
| Read `state$runtime()` and unpack its slots (`$ok`, `$plot`, `$code_text`, `$error`). This takes the reactive dependency that wires re-renders to *Update plot* clicks. | Call `ptr_extract_plot(state)` / `ptr_extract_code(state)` / `ptr_extract_error(state)`. These accessors wrap `shiny::isolate()` — they return the *current* value without registering a reactive dependency. |

Mixing the two in a render block — calling `ptr_extract_plot(state)` inside `renderPlotly({...})` — silently breaks reactivity (the block fires once on mount and never again). Stick to `state$runtime()` inside reactive blocks.

### Custom plot renderer — `state$runtime()$plot` inside `renderPlotly()`

The pattern: wire the runtime with the **non-module split** so you can leave the bundled plot pane out entirely, then unpack `state$runtime()` inside a `plotly::renderPlotly()` block (or `ggiraph::renderGirafe()`, or any custom renderer). The split path is what makes "omit the bundled pane" concrete: render `ptr_controls_ui()` for the widgets, render **no** `ptr_outputs_ui()`, and place your own output container instead. `ptr_server()` still registers the canonical `ptr_plot` / `ptr_error` / `ptr_code` outputs, but with no UI slot bound to them they stay inert — and it returns the same `ptr_state` you read with `state$runtime()`. (With the L3 pieces below you can be even more explicit: render `ptr_ui_controls()` and simply never place `ptr_ui_plot()`, so there is no bundled pane at all.)

```{r l3-plotly}
formula <- "ggplot(iris, aes(x = var, y = var, color = var)) + geom_point()"

ui <- shiny::fluidPage(
  shiny::fluidRow(
    shiny::column(5, ptr_controls_ui(formula = formula)),            # widgets only — no ptr_outputs_ui()
    shiny::column(7, plotly::plotlyOutput("custom_plot", height = "500px"))
  )
)
server <- function(input, output, session) {
  state <- ptr_server(input, output, session, formula)               # runtime wired; bundled outputs registered but inert
  output$custom_plot <- plotly::renderPlotly({
    res <- state$runtime()
    shiny::req(isTRUE(res$ok), res$plot)
    plotly::ggplotly(res$plot)
  })
}

shiny::shinyApp(ui, server)
```

`req(isTRUE(res$ok), res$plot)` keeps the panel quiet between draws and on transient error states.

If you would rather keep ggpaintr's chrome *and* add a second custom view, the module path works too — `state <- ptr_module_server("p", formula)` next to `ptr_module_ui("p", formula)` — but `ptr_module_ui()` always composes `ptr_controls_ui()` **and** `ptr_outputs_ui()`, so the bundled plot pane renders alongside your custom one. There is no argument to drop the bundled pane from `ptr_module_ui()`; use the split above, or compose the L3 pieces (`ptr_ui_controls("p", …)` + your own output, skipping `ptr_ui_plot()`), when you want exactly one plot.

`vignette("ggpaintr-gallery")` §5 *Interactive output packages* is the canonical worked-example reference for this custom-renderer pattern — it shows the full plotly and ggiraph setups end-to-end, including the ggiraph variant with click handling and a custom-placeholder colour picker. The chunk above is the minimal scaffold; gallery §5 is the runnable recipe.

### Shared widgets driving custom renderers

Combine the L2 shared panel with L3 custom outputs: one `ptr_shared_ui()` panel feeds the modules, each module's returned state feeds its own `plotly::renderPlotly()`. The wiring is the L2 shared recipe plus a captured `state` per module — the modules need their own namespaces (each formula's per-placeholder ids would otherwise collide), so the module path, not the split, is the right tool here.

```{r l3-shared}
plots <- list(
  "ggplot(iris, aes(x = var(shared = 'metric'), y = Sepal.Length, fill = Species)) + geom_boxplot()",
  "ggplot(iris, aes(x = var(shared = 'metric'), y = Sepal.Width,  fill = Species)) + geom_violin()"
)

ui <- shiny::fluidPage(
  ptr_shared_ui(
    formulas  = plots,
    shared_ui = list(metric = function(id) shiny::selectInput(id, "Metric", names(iris)))
  ),
  shiny::fluidRow(
    shiny::column(3, ptr_controls_ui("plot_1", plots[[1]])),
    shiny::column(9, plotly::plotlyOutput("custom_1")),
  ),
  shiny::fluidRow(
    shiny::column(3, ptr_controls_ui("plot_2", plots[[2]])),
    shiny::column(9, plotly::plotlyOutput("custom_2")),
  )
)
server <- function(input, output, session) {
  shared <- ptr_shared_server(formulas = plots)
  state1 <- ptr_module_server("plot_1", plots[[1]], shared_state = shared)
  state2 <- ptr_module_server("plot_2", plots[[2]], shared_state = shared)
  output$custom_1 <- plotly::renderPlotly({
    res <- state1$runtime(); shiny::req(isTRUE(res$ok), res$plot)
    plotly::ggplotly(res$plot)
  })
  output$custom_2 <- plotly::renderPlotly({
    res <- state2$runtime(); shiny::req(isTRUE(res$ok), res$plot)
    plotly::ggplotly(res$plot)
  })
}

shiny::shinyApp(ui, server)
```

`ptr_module_ui()` is replaced by `ptr_controls_ui("plot_N", ...)` + your own `plotlyOutput()` so each tile shows exactly one (custom) plot, not the bundled pane alongside it.

**Passing the shared bundle to `ptr_server()` instead of `ptr_module_server()`.** `ptr_server()` accepts the bundle directly through `shared_state =`, exactly like `ptr_module_server()`. For a single-plot custom renderer with no module wrapper:

```{r l3-shared-server}
shared <- ptr_shared_server(formulas = list(formula))   # formula carries a shared = "..." annotation
state  <- ptr_server(input, output, session, formula, shared_state = shared)
```

The four slots (`shared`, `draw_trigger`, `shared_resolutions`, `shared_stage_enabled`) are still accepted individually through `...` if you need to override just one — an explicit `...` value wins over the bundle.

### Custom code panel

Bind to `state$runtime()` inside `renderText({...})` and pull `code_text` off it. Inside the same server as the `state <-` assignment above:

```{r l3-code-pane}
output$my_code <- shiny::renderText({
  state$runtime()$code_text %||% ""
})
```

For non-reactive contexts (a download handler, a snapshot test), use the public accessor `ptr_extract_code(state)`. It returns a single string and includes any `ptr_gg_extra()` expressions captured on the state.

### Custom error UI

`state$runtime()$error` is the latest runtime error string (or `NULL` when the last cycle succeeded). Inside the same server:

```{r l3-error-ui}
output$my_status <- shiny::renderUI({
  msg <- state$runtime()$error
  if (is.null(msg)) shiny::tags$div(class = "status-ok", "Plot is up to date")
  else              shiny::tags$div(class = "status-error", msg)
})
```

`ptr_extract_error(state)` is the non-reactive counterpart.

### Programmatic layer injection — `ptr_gg_extra()`

`ptr_gg_extra(state, ...)` evaluates one or more `ggplot2` expressions and attaches the results as "extras" on the state. The runtime folds them into the rendered plot during the next cycle, and `state$runtime()$code_text` (and `ptr_extract_code(state)`) include them in the printable code. Each call replaces the previously captured extras — pass everything you want layered on in a single call.

```{r l3-gg-extra}
formula <- "ggplot(mtcars, aes(x = mpg, y = hp)) + geom_point()"

ui <- shiny::fluidPage(
  shiny::actionButton("add_log", "Toggle log-scale"),
  ptr_module_ui("p", formula)
)
server <- function(input, output, session) {
  state <- ptr_module_server("p", formula)
  shiny::observeEvent(input$add_log, {
    ptr_gg_extra(state, ggplot2::scale_x_log10())
  })
}

shiny::shinyApp(ui, server)
```

Eval errors from the captured expressions leave the existing extras untouched (atomic update). Extras are also suppressed automatically when the underlying runtime reports a failure, so stale extras from a prior successful draw never surface during an error state.

## L3 — Own every piece of the UI

L2 lets you place ggpaintr's two *halves* (`ptr_controls_ui()` / `ptr_outputs_ui()`) wherever you like, and read `state` for a custom renderer. L3 goes one level finer: **every piece of ggpaintr's public UI has its own exported function**, so you can pick exactly the pieces you want, place each one anywhere in your own layout, and wire them with the public server API. The server contract is unchanged — the pieces write to / read from the same canonical ids `ptr_server()` already targets — so L3 is purely a UI-composition story.

The pieces:

| Function | What it emits | Key arguments |
|---|---|---|
| `ptr_ui_header(title)` | The branded header bar (logo + title) | `title` (default `"ggpaintr"`) |
| `ptr_ui_controls(id, formula, …)` | The generated control widgets (layer picker, per-layer panels, *Update plot*) | `id`, `formula`, `ui_text`, `checkbox_defaults`, `expr_check`, `render_shared_section` |
| `ptr_ui_plot(id, error, code_toggle)` | The plot card (`ptr_plot`) | `error` nests the inline error slot (default `TRUE`); `code_toggle` adds the show-code button |
| `ptr_ui_error(id)` | The inline error slot (`ptr_error`) on its own | `id` |
| `ptr_ui_code(id, style)` | The generated-code pane (`ptr_code`) | `style = "panel"` (plain, always visible — default) or `"window"` (slide-out chrome) |
| `ptr_ui_code_toggle(id, plot_error)` | Plot + slide-out code, wired to the show-code toggle — exactly the block the bundled apps render | `plot_error` nests the inline error (default `TRUE`) |
| `ptr_ui_assets(css)` | The CSS/JS bundle (theme + structural CSS + code-window JS + your `css`) | `css` |

Three facts make the difference between L3 working and looking broken:

1. **Keep a Bootstrap page at the root.** ggpaintr's controls use a `shinyWidgets::pickerInput()` (the layer selector) and the Bootstrap grid (`fluidRow()` / `column()` / `sidebarLayout()`). Those need Bootstrap's own CSS/JS, which Shiny loads **only** when the outermost UI object is a Bootstrap page builder — `shiny::fluidPage()`, `bootstrapPage()`, `fillPage()`, `navbarPage()`, …. A bare `shiny::tags$div()` root is *not* enough: the picker renders unstyled/broken even with `ptr_ui_assets()` present, because that bundle is ggpaintr's *own* CSS/JS (theme + code-window behaviour), not Bootstrap. The bundled `ptr_app()` / `ptr_module_ui()` always wrap in `fluidPage()` for exactly this reason — so should you.
2. **The pieces emit no assets and no wrapper.** Unlike `ptr_controls_ui()` / `ptr_outputs_ui()` (which self-wrap in `<div class="ptr-app">` and inject assets), the L3 pieces are bare. The bundled `ggpaintr.css` theme is scoped under `.ptr-app`, so inside the Bootstrap page you must (a) include `ptr_ui_assets()` once — anywhere; it is idempotent and self-deduping — and (b) wrap the region you want themed in `shiny::tags$div(class = "ptr-app", …)`. Skip either and the widgets still *work* but lose the ggpaintr styling. The reliable shape, mirroring the bundled apps, is `fluidPage(ptr_ui_assets(), div(class = "ptr-app", …pieces…))`.
3. **The slide-out code toggle is opt-in.** A standalone `ptr_ui_code()` is a plain, always-visible panel that needs no wiring. The draggable slide-out window (with the show-code button, Copy, Close, drag) only works when the plot and a `style = "window"` code pane sit inside one `.ptr-output` scope wired together — use `ptr_ui_code_toggle()` for that exact bundle. Without it, place `ptr_ui_plot()` and `ptr_ui_code()` independently and the code pane simply stands on its own.

Ids must line up: pass the same `id` to every piece and to the server (a `shiny::moduleServer(id, …)` wrapping `ptr_server()`, or `ptr_module_server(id, …)`). The single-embedding case needs neither the `id` nor the wrapper.

A fully hand-laid-out page — header, controls, plot, error, and code each in their own region:

```{r l3-pieces}
formula <- "ggplot(iris, aes(x = var, y = var, color = var)) + geom_point()"

ui <- shiny::fluidPage(                     # Bootstrap page root — loads pickerInput / grid deps
  ptr_ui_assets(),                          # include the CSS/JS bundle once
  shiny::tags$div(
    class = "ptr-app",                      # scopes the bundled theme
    ptr_ui_header("Iris explorer"),
    shiny::fluidRow(
      shiny::column(4, ptr_ui_controls(formula = formula)),
      shiny::column(8, ptr_ui_plot(error = FALSE))   # plot only; error placed below
    ),
    ptr_ui_error(),                         # error banner in its own row
    shiny::tags$details(
      shiny::tags$summary("Generated code"),
      ptr_ui_code()                         # plain, always-visible code panel
    )
  )
)
server <- function(input, output, session) {
  ptr_server(input, output, session, formula)   # binds ptr_plot / ptr_error / ptr_code
}

shiny::shinyApp(ui, server)
```

Want the familiar slide-out code window while still owning the surrounding layout? Swap the plot + code pieces for the one wiring helper:

```{r l3-pieces-toggle}
ui <- shiny::fluidPage(
  ptr_ui_assets(),
  shiny::tags$div(
    class = "ptr-app",
    ptr_ui_header("Iris explorer"),
    shiny::sidebarLayout(
      shiny::sidebarPanel(ptr_ui_controls(formula = formula)),
      shiny::mainPanel(ptr_ui_code_toggle())   # plot + inline error + slide-out code, wired
    )
  )
)
# server unchanged
```

`ptr_ui_code_toggle()` is byte-for-byte the output block `ptr_app()` and `ptr_module_ui()` render internally, so behaviour and performance match the all-in-one path exactly — it is just exposed so you can place it yourself.

### L3 with shared widgets

The shared panel composes with the pieces. Build the panel with `ptr_shared_ui()` / `ptr_shared_server()` as in L2, place the per-plot pieces yourself, and hand each `ptr_server()` the bundle via `shared_state =` (no `moduleServer()` wrapper needed when ids are distinct per plot):

```{r l3-pieces-shared}
plots <- list(
  "ggplot(iris, aes(x = var(shared = 'metric'), y = Sepal.Length, fill = Species)) + geom_boxplot()",
  "ggplot(iris, aes(x = var(shared = 'metric'), y = Sepal.Width,  fill = Species)) + geom_violin()"
)

ui <- shiny::fluidPage(
  ptr_ui_assets(),
  shiny::tags$div(
    class = "ptr-app",
    ptr_shared_ui(
      formulas  = plots,
      shared_ui = list(metric = function(id) shiny::selectInput(id, "Metric", names(iris)))
    ),
    shiny::fluidRow(
      shiny::column(6, ptr_ui_controls("p1", plots[[1]]), ptr_ui_plot("p1")),
      shiny::column(6, ptr_ui_controls("p2", plots[[2]]), ptr_ui_plot("p2"))
    )
  )
)
server <- function(input, output, session) {
  shared <- ptr_shared_server(formulas = plots)
  shiny::moduleServer("p1", function(input, output, session) {
    ptr_server(input, output, session, plots[[1]], shared_state = shared)
  })
  shiny::moduleServer("p2", function(input, output, session) {
    ptr_server(input, output, session, plots[[2]], shared_state = shared)
  })
}

shiny::shinyApp(ui, server)
```

### Placing individual placeholder widgets — `build_ui_for()`

`ptr_ui_controls()` is the whole controls panel. If you need to scatter the *individual* placeholder widgets across the page rather than the assembled panel, drop to the exported `build_ui_for()` S3 generic — the same one ggpaintr uses internally to turn one parsed node into its widget. Walk the translated tree (`ptr_translate(formula)$layers`) and call `build_ui_for()` per node, placing each returned tag yourself. This is the lowest-level UI escape hatch; see `?build_ui_for` for the per-node contract and `vignette("ggpaintr-customization")` for registering custom widget types.
