---
title: "ggpaintr Extensibility"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ggpaintr Extensibility}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)

library(ggpaintr)
```

## Overview

`ggpaintr` still ships the default `ggpaintr_app()` and `ggpaintr_server()`
wrappers, but it now also exposes a supported integration layer for users who
already have a Shiny app and want to embed the generated `ggpaintr` controls,
runtime, and export behavior into their own UI.

The supported integration pieces are:

- `ggpaintr_ids()`
- `ggpaintr_server_state()`
- `ggpaintr_bind_control_panel()`
- `ggpaintr_bind_draw()`
- `ggpaintr_bind_export()`
- `ggpaintr_bind_plot()`
- `ggpaintr_bind_error()`
- `ggpaintr_bind_code()`
- `ggpaintr_plot_value()`
- `ggpaintr_error_value()`
- `ggpaintr_code_value()`
- `ggpaintr_controls_ui()`
- `ggpaintr_outputs_ui()`

## Recipe 1: Embed `ggpaintr` with the default binders

```{r, eval = FALSE}
library(ggpaintr)
library(shiny)

ui <- fluidPage(
  titlePanel("Embedded ggpaintr"),
  sidebarLayout(
    sidebarPanel(
      ggpaintr_controls_ui()
    ),
    mainPanel(
      ggpaintr_outputs_ui()
    )
  )
)

server <- function(input, output, session) {
  paintr_state <- ggpaintr_server_state(
    "ggplot(data = iris, aes(x = var, y = var)) +
      geom_point() +
      labs(title = text)"
  )

  ggpaintr_bind_control_panel(input, output, paintr_state)
  ggpaintr_bind_draw(input, paintr_state)
  ggpaintr_bind_export(output, paintr_state)
  ggpaintr_bind_plot(output, paintr_state)
  ggpaintr_bind_error(output, paintr_state)
  ggpaintr_bind_code(output, paintr_state)
}

shinyApp(ui, server)
```

This is the closest supported equivalent to `ggpaintr_server()`, but it lets
you place the controls and outputs inside your own page layout.

## Recipe 2: Customize top-level ids

Use `ggpaintr_ids()` when your app already has its own naming scheme or you want
to mount `ggpaintr` outputs in specific UI containers.

```{r, eval = FALSE}
ids <- ggpaintr_ids(
  control_panel = "builder_controls",
  draw_button = "render_plot",
  export_button = "export_app",
  plot_output = "main_plot",
  error_output = "main_error",
  code_output = "main_code"
)

ids
```

```{r, eval = FALSE}
ui <- fluidPage(
  sidebarLayout(
    sidebarPanel(
      ggpaintr_controls_ui(ids = ids)
    ),
    mainPanel(
      ggpaintr_outputs_ui(ids = ids)
    )
  )
)

server <- function(input, output, session) {
  paintr_state <- ggpaintr_server_state(
    "ggplot(data = iris, aes(x = var, y = var)) + geom_point()",
    ids = ids
  )

  ggpaintr_bind_control_panel(input, output, paintr_state, ids = ids)
  ggpaintr_bind_draw(input, paintr_state, ids = ids)
  ggpaintr_bind_export(output, paintr_state, ids = ids)
  ggpaintr_bind_plot(output, paintr_state, ids = ids)
  ggpaintr_bind_error(output, paintr_state, ids = ids)
  ggpaintr_bind_code(output, paintr_state, ids = ids)
}
```

Only the six top-level ids are configurable in this phase. Internal placeholder
ids and dynamic `var-*` output ids remain package-owned.

## Recipe 3: Customize the returned plot in `renderPlot()`

The default `ggpaintr_bind_plot()` binder preserves the current wrapper
behavior. If you want to modify the built plot before rendering, write your own
`renderPlot()` and use `ggpaintr_plot_value()`.

```{r, eval = FALSE}
server <- function(input, output, session) {
  paintr_state <- ggpaintr_server_state(
    "ggplot(data = iris, aes(x = var, y = var)) + geom_point()"
  )

  ggpaintr_bind_control_panel(input, output, paintr_state)
  ggpaintr_bind_draw(input, paintr_state)
  ggpaintr_bind_export(output, paintr_state)
  ggpaintr_bind_error(output, paintr_state)
  ggpaintr_bind_code(output, paintr_state)

  output$outputPlot <- renderPlot({
    plot_obj <- ggpaintr_plot_value(paintr_state$runtime())

    if (is.null(plot_obj)) {
      plot.new()
      return(invisible(NULL))
    }

    plot_obj + ggplot2::theme_minimal()
  })
}
```

`ggpaintr_plot_value()` returns the raw `ggplot` object on success and `NULL`
otherwise. That keeps the advanced customization seam side-effect free and lets
you decide how to render failure states in your own app.

## Pure helpers versus bind helpers

Use the bind helpers when you want the standard `ggpaintr` behavior with custom
layout or custom top-level ids.

Use the pure value helpers when you want to own the rendering details:

- `ggpaintr_plot_value()` for plot customization
- `ggpaintr_error_value()` for custom `renderUI()` workflows
- `ggpaintr_code_value()` for custom `renderText()` or downstream processing
