---
title: "ggpaintr in the Wild"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ggpaintr in the Wild}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment   = "#>",
  warning   = FALSE,
  message   = FALSE,
  eval      = interactive()
)
```

```{r clipboard, echo = FALSE, eval = TRUE}
xaringanExtra::use_clipboard()
```

## 1. Setup

Every chunk in this vignette is marked `eval = interactive()` — they will not run during vignette build, but each block is runnable as-is at the R prompt after:

```{r libs, eval = TRUE}
library(ggpaintr)   # also attaches ggplot2
library(shiny)
library(rlang)
```

Some sections also load `dplyr`, `plotly`, `ggiraph`, or one of the ggplot2 extension packages (`ggpcp`, `ggridges`, `ggrepel`, `ggalluvial`, `ggdist`). Each section names what it needs.

## 2. What this vignette is

The other vignettes teach the workflow piece by piece. This one shows what the pieces add up to. Every section pairs a complete `ggplot2` (or extension-package) graphic with the `ggpaintr` formula that turns it into an interactive Shiny app. Read the original to know what is being built; read the formula to see which arguments become widgets.

Copy a chunk into a session, install the extension package named at the top of the section if you do not have it, and run it. The examples assume the setup chunks above have already executed.

Several gallery sections below reference custom placeholders — `range` (a slider returning a length-2 numeric), `colvars` (a multi-column picker), and `date` (a date-range picker) — alongside the five built-ins (`var`, `text`, `num`, `expr`, `upload`). The three constructors that register such widgets (`ptr_define_placeholder_value()`, `ptr_define_placeholder_consumer()`, `ptr_define_placeholder_source()`), the full hook contract, and a worked example for each are covered in `vignette("ggpaintr-customization")`. Run the placeholder-registration chunk from that vignette once per R session before launching any gallery example that uses these keywords.

## 3. A realistic ggplot2 graphic, parameterized

A typical exploratory plot stacks several layers and reads several hard-coded constants. Here is a multi-layer fuel-economy chart from the `mpg` dataset:

```{r mpg-original}
ggplot(mpg, aes(displ, hwy, color = class)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_smooth(method = "loess", se = FALSE, span = 0.75) +
  facet_wrap(~ drv) +
  scale_color_brewer(palette = "Set2") +
  labs(title = "Highway MPG vs engine displacement",
       x = "Engine displacement (L)", y = "Highway MPG") +
  coord_cartesian(xlim = c(1, 7), ylim = c(10, 45))
```

Replace each tuning knob with a placeholder token. `var` becomes a column picker populated from the data; `num` becomes `numericInput`s; `text` becomes `textInput`s; the custom `range` becomes a slider:

```{r mpg-paintr}
ptr_app(
  "ggplot(data = mpg, aes(x = var, y = var, color = var)) +
     geom_point(alpha = num, size = num) +
     geom_smooth(method = 'loess', se = FALSE, span = num) +
     facet_wrap(~ var) +
     scale_color_brewer(palette = 'Set2') +
     labs(title = text, x = text, y = text) +
     coord_cartesian(xlim = range, ylim = range)"
)
```

Every aesthetic, alpha, size, span, and the two coordinate ranges is now a widget. Rebuilding the original graphic from the running app is a matter of picking those values; the code panel under the plot lifts the selections back into a static script.

## 4. Data-cleaning pipelines into `ggplot()`

`ggpaintr` parses any expression that evaluates to a `ggplot` object, including the common "clean, then plot" pipeline where the data is piped into `ggplot()`. Pipeline stages with placeholders get a per-layer **Data** sub-tab in the sidebar; column pickers further down the pipeline lazily resolve their upstream against the current Data-tab inputs, so picking a column in stage one immediately refreshes the choices for stage two without requiring an explicit "apply" click.

```{r pipe-original}
library(dplyr)

mpg |>
  dplyr::filter(displ > 1.5) |>
  dplyr::group_by(class) |>
  dplyr::filter(dplyr::n() > 5) |>
  dplyr::ungroup() |>
  ggplot(aes(displ, hwy, color = class)) +
  geom_point(alpha = 0.5)
```

The same pipeline parameterized — the two `filter` thresholds and the point alpha become widgets:

```{r pipe-paintr}
library(dplyr)

ptr_app(
  "mpg |>
     dplyr::filter(displ > num) |>
     dplyr::group_by(class) |>
     dplyr::filter(dplyr::n() > num) |>
     dplyr::ungroup() |>
     ggplot(aes(var, var, color = class)) +
     geom_point(alpha = num) +
     labs(title = text)"
)
```

Each pipeline stage gets its own enable/disable checkbox (G11). Toggling a stage off rewrites the generated code as if that step were never there; toggle it back on and the threshold widget resurfaces. Note also that `var` (and any custom consumer like `colvars`) introspects the upstream data automatically — pipe-walking through `|>` chains is part of the runtime, not your placeholder code.

## 5. Interactive output packages

`ggpaintr` does not lock you into `renderPlot`. The runtime's `state$runtime()` reactive returns a list whose `$plot` slot is the fitted `ggplot` object, suitable for any backend that accepts a `ggplot` — `plotly` and `ggiraph` are the two most common. `ptr_extract_plot(state)` is the same accessor in non-reactive contexts.

The recipe is the same in both subsections: lay out `ptr_module_ui("id", formula)` for the sidebar + default plot panel, add a second column with the alternative output, and in the server use `state <- ptr_module_server("id", formula)` to wire everything up, then bind your custom output to `state$runtime()`.

### 5.1 plotly tooltips

Original — a hover-tooltipped scatter:

```{r plotly-original}
library(plotly)

p <- ggplot(mpg, aes(displ, hwy, color = class,
                     text = paste(manufacturer, model, sep = " "))) +
       geom_point(size = 3, alpha = 0.7) +
       coord_cartesian(xlim = c(1, 7), ylim = c(10, 45))

plotly::ggplotly(p, tooltip = "text")
```

ggpaintr version — embed via the module helpers and own the plotly output:

```{r plotly-paintr}
library(plotly)

formula <- "ggplot(data = mpg,
                   aes(x = var, y = var, color = var,
                       text = paste(manufacturer, model, sep = ' '))) +
              geom_point(size = num, alpha = num) +
              coord_cartesian(xlim = range, ylim = range)"

ui <- fluidPage(
  fluidRow(
    column(5, ptr_module_ui("plotly_demo", formula)),
    column(7, plotly::plotlyOutput("interactive_plot", height = "500px"))
  )
)

server <- function(input, output, session) {
  state <- ptr_module_server("plotly_demo", formula)
  output$interactive_plot <- plotly::renderPlotly({
    res <- state$runtime()
    shiny::req(isTRUE(res$ok), res$plot)
    plotly::ggplotly(res$plot, tooltip = "text")
  })
}

shinyApp(ui, server)
```

Reading `state$runtime()` outside `isolate()` is what wires the plotly re-render to Update Plot clicks; `req(isTRUE(res$ok), res$plot)` keeps the panel quiet between draws and on transient error states.

### 5.2 ggiraph tooltips and click handling

Original — interactive points with a colour-picked highlight:

```{r ggiraph-original}
library(ggiraph)

ggplot(mpg, aes(displ, hwy)) +
  geom_point_interactive(aes(tooltip = model), size = 3, color = "#e63946")
```

ggpaintr version with a colour-picker placeholder so the user can recolour the layer:

```{r ggiraph-paintr}
library(ggiraph)
library(colourpicker)

ptr_define_placeholder_value(
  keyword       = "color",
  build_ui      = function(node, label = NULL, ...) {
    colourpicker::colourInput(node$id, label = label %||% "Highlight colour",
                              value = "#e63946")
  },
  resolve_expr  = function(value, node, ...) {
    if (is.null(value) || identical(value, "")) return(NULL)
    as.character(value)
  },
  copy_defaults = list(label = "Highlight colour for {param}")
)

formula <- "ggplot(data = mpg, aes(x = var, y = var)) +
              geom_point_interactive(aes(tooltip = var),
                                     size = num, color = color)"

ui <- fluidPage(
  fluidRow(
    column(5, ptr_module_ui("ggiraph_demo", formula)),
    column(7, ggiraph::girafeOutput("interactive_plot", height = "500px"))
  )
)

server <- function(input, output, session) {
  state <- ptr_module_server("ggiraph_demo", formula)
  output$interactive_plot <- ggiraph::renderGirafe({
    res <- state$runtime()
    shiny::req(isTRUE(res$ok), res$plot)
    ggiraph::girafe(code = print(res$plot))
  })
}

shinyApp(ui, server)
```

## 6. ggplot2 extension packages

`ggpaintr` does not know about extension geoms — it sees them as ordinary layer calls and parameterizes their arguments the same way it parameterizes core geoms. The five examples below cover a parallel-coordinates plot, ridge densities, repelled labels, alluvial flows, and `ggdist` distribution slabs.

### 6.1 ggpcp — parallel coordinates

Original (uses the `flea` dataset shipped by `GGally`):

```{r pcp-original}
library(ggpcp)
data(flea, package = "GGally")

flea |>
  pcp_select(species, tars1, tars2, head, aede1, aede2, aede3) |>
  pcp_scale(method = "uniminmax") |>
  pcp_arrange() |>
  ggplot(aes_pcp()) +
    geom_pcp_axes() +
    geom_pcp(aes(colour = species))
```

ggpaintr version — let the user pick which columns to put on the parallel axes via the `colvars` placeholder:

```{r pcp-paintr}
library(ggpcp)
data(flea, package = "GGally")

ptr_app(
  "ggplot(data = flea |>
                  pcp_select(colvars) |>
                  pcp_scale(method = 'uniminmax') |>
                  pcp_arrange(),
          mapping = aes_pcp()) +
     geom_pcp_axes() +
     geom_pcp(aes(colour = species)) +
     coord_cartesian(xlim = range, ylim = range)"
)
```

### 6.2 ggridges — ridge densities

Original (uses `lincoln_weather` shipped by `ggridges`):

```{r ridges-original}
library(ggridges)

ggplot(lincoln_weather,
       aes(x = `Mean Temperature [F]`, y = Month, fill = ..x..)) +
  geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01,
                               bandwidth = 3) +
  scale_fill_viridis_c(option = "C")
```

ggpaintr version — bandwidth, scale, and the colour aesthetic become widgets:

```{r ridges-paintr}
library(ggridges)

ptr_app(
  "ggplot(data = lincoln_weather, aes(x = var, y = var, fill = ..x..)) +
     geom_density_ridges_gradient(scale = num,
                                  rel_min_height = num,
                                  bandwidth = num) +
     scale_fill_viridis_c(option = 'C') +
     labs(title = text)"
)
```

### 6.3 ggrepel — labelled scatter

Original — label cars by row name, repelling overlaps:

```{r repel-original}
library(ggrepel)

mt <- data.frame(mtcars, model = rownames(mtcars))

ggplot(mt, aes(wt, mpg, label = model)) +
  geom_point(size = 3, color = "steelblue") +
  geom_text_repel(size = 3.5, max.overlaps = 12, box.padding = 0.3)
```

ggpaintr version — `max.overlaps`, `size`, and `box.padding` all turn into numeric inputs. The label aesthetic is bound by `var`:

```{r repel-paintr}
library(ggrepel)

mt <- data.frame(mtcars, model = rownames(mtcars))

ptr_app(
  "ggplot(data = mt, aes(x = var, y = var, label = var)) +
     geom_point(size = num, color = 'steelblue') +
     geom_text_repel(size = num,
                     max.overlaps = num,
                     box.padding = num) +
     labs(title = text)",
  envir = environment()
)
```

`envir = environment()` is the seam that lets the runtime see the local `mt` data frame. `ptr_app()` defaults to `parent.frame()`, which would pick up `mt` here automatically; passing `environment()` is explicit.

### 6.4 ggalluvial — flow diagrams

Original (uses the `vaccinations` dataset shipped by `ggalluvial`):

```{r alluvial-original}
library(ggalluvial)

ggplot(vaccinations,
       aes(x = survey, stratum = response, alluvium = subject,
           y = freq, fill = response)) +
  geom_flow(alpha = 0.7) +
  geom_stratum() +
  scale_x_discrete(expand = c(0.1, 0.1))
```

ggpaintr version — alpha and the fill aesthetic become widgets:

```{r alluvial-paintr}
library(ggalluvial)

ptr_app(
  "ggplot(data = vaccinations,
          aes(x = survey, stratum = response, alluvium = subject,
              y = freq, fill = var)) +
     geom_flow(alpha = num) +
     geom_stratum() +
     scale_x_discrete(expand = c(0.1, 0.1)) +
     labs(title = text)"
)
```

### 6.5 ggdist — distribution visualizations

Original — half-eye distributions per cylinder count:

```{r dist-original}
library(ggdist)

ggplot(mtcars, aes(x = factor(cyl), y = mpg)) +
  stat_halfeye(adjust = 0.5, justification = -0.2,
               .width = 0, point_colour = NA, slab_alpha = 0.6) +
  geom_boxplot(width = 0.15, outlier.shape = NA) +
  coord_flip()
```

ggpaintr version — every visual constant of `stat_halfeye()` becomes a slider or numeric input:

```{r dist-paintr}
library(ggdist)

ptr_app(
  "ggplot(data = mtcars, aes(x = factor(cyl), y = var)) +
     stat_halfeye(adjust = num,
                  justification = num,
                  .width = num,
                  point_colour = NA,
                  slab_alpha = num) +
     geom_boxplot(width = num, outlier.shape = NA) +
     coord_flip() +
     labs(title = text)"
)
```

## 7. Where to go next

This gallery is the recipe book. The other vignettes answer different user questions:

- **`vignette("ggpaintr-use-cases")`** — the three-level ladder. L1 all-in-one (`ptr_app()` / `ptr_app_bslib()` / `ptr_app_grid()`); L2 embed in your own Shiny app (`ptr_controls_ui` / `ptr_outputs_ui` / `ptr_server` and the module pair `ptr_module_ui` / `ptr_module_server`); L3 custom rendering via the extract accessors (`ptr_extract_plot()` / `_code()` / `_error()` and `ptr_gg_extra()`). The plotly / ggiraph examples in §5 above are the canonical L3 worked examples — use-cases L3 cross-links back here.
- **`vignette("ggpaintr-customization")`** — `ui_text` overrides, the three custom-placeholder constructors (full hook contract and worked examples — the `range` / `colvars` / `date` constructors used throughout this gallery live there), and the shared-placeholder pattern for one widget driving multiple plots.
- **`vignette("ggpaintr-safety")`** — `expr_check`, the denylist + AST-walker safety model, upload trust boundary, public-deployment checklist.
- **`vignette("ggpaintr-llm")`** — wiring ggpaintr up as an LLM-callable tool with ellmer.

For canonical API reference, follow the `?fn` links:

- `?ptr_app`, `?ptr_module_ui`, `?ptr_module_server`, `?ptr_app_grid` — entry points.
- `?ptr_define_placeholder_value`, `?ptr_define_placeholder_consumer`, `?ptr_define_placeholder_source` — the three constructors that extend the registry.
- `?ptr_extract_plot`, `?ptr_extract_code`, `?ptr_extract_error`, `?ptr_gg_extra` — the embed-time accessors used by the plotly / ggiraph sections.
