---
title: "Using ggpaintr from an LLM with ellmer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using ggpaintr from an LLM with ellmer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  warning  = FALSE,
  message  = FALSE
)
```

```{r libs}
library(ggpaintr)
```

## 1. Motivation

Large language models are fluent at writing raw `shiny::fluidPage()` +
`server <- function(input, output, session) {...}` scaffolding from
memory. They are *not* fluent at writing ggpaintr, because every model
today has zero prior exposure to this package. Without help, a user
asking "build me a Shiny app where I can pick columns from iris and
see a scatter plot" gets hundreds of lines of hand-rolled Shiny code
that the ggpaintr `ptr_app()` one-liner would have covered.

This vignette shows how to make any [ellmer](https://ellmer.tidyverse.org/)
chat session reach for ggpaintr first, using two exported helpers:

- `ptr_llm_primer()` — the system-prompt preamble that teaches the
  model the three-level integration model and the "call the docs tool
  before writing code" rule.
- `ptr_llm_register()` — one-line registration of the `ggpaintr_docs`
  tool that serves focused runnable examples on demand.

The tool is backed by `ptr_llm_topic()`, which returns the content of
one file under `inst/llm/topics/` (10 topics, ~1500–5000 characters
each). The LLM pulls only the topic it needs, instead of the whole
manual loading the context window on every turn.

## 2. Quick start — three lines of setup

```{r quickstart, eval = FALSE}
library(ellmer)
library(ggpaintr)

chat <- chat_anthropic(system_prompt = ptr_llm_primer())
ptr_llm_register(chat)

chat$chat("Build a Shiny app where the user picks X and Y columns from mtcars.")
```

Under the hood:

- `chat_anthropic()` loads the primer (~290 tokens) into the system
  prompt, which stays in context every turn.
- `ptr_llm_register()` attaches a single tool named `ggpaintr_docs` to
  the chat session. The tool's one argument — `name` — is constrained
  by `type_enum()` to the exact set of topic names returned by
  `ptr_llm_topics()`, so the LLM cannot request a topic that does not
  exist.

On the first turn the model sees the user's prompt, recognises it as a
plot-explorer task, calls `ggpaintr_docs(name = "level1_ptr_app")`, and
writes a `ptr_app(...)` snippet rather than raw Shiny.

## 3. What the primer + tool split actually buys you

A common alternative is to stuff the entire ggpaintr manual into the
system prompt. That works but wastes tokens: every chat turn pays for
the full manual whether or not it is relevant.

The primer/tool split is a lightweight form of retrieval:

| Layer        | When it loads            | What it teaches                                     |
|--------------|--------------------------|-----------------------------------------------------|
| Primer       | Every turn (~290 tokens) | ggpaintr exists; 3-level model; decision rule.      |
| `ggpaintr_docs(topic)` | On demand        | One focused runnable example (~300–1300 tokens).    |

The primer has an explicit rule: *"Before writing R code for an
interactive plot task, call `ggpaintr_docs(topic)` to fetch the runnable
example."* That rule plus the enum-constrained argument is what makes
the LLM reliably reach for the tool instead of hallucinating an API.

## 4. The available topics

```{r topics}
ptr_llm_topics()
```

A short guide to each:

| Topic                    | When the LLM should call it                                              |
|--------------------------|--------------------------------------------------------------------------|
| `overview`               | First orientation — the 3-level model and decision rule.                 |
| `formula_syntax`         | Which placeholder keyword goes where; `var` vs formula-level transforms. |
| `level1_ptr_app`         | Quick turn-key app; no custom Shiny.                                     |
| `level2_embed`           | Embed ggpaintr inside the user's own Shiny app.                          |
| `level2_custom_ids`      | Avoid id collisions with host-app widgets via `ptr_build_ids()`.         |
| `level2_ui_text`         | Override labels, help, placeholders; six-section cascade.                |
| `level3_headless`        | Render plots without Shiny — batch reports, fixtures, static sites.      |
| `level3_custom_render`   | Own `renderPlot()` with host-app post-processing (theme picker, etc.).   |
| `level3_gg_extra`        | Round-trip host layers into the generated code pane.                     |
| `custom_placeholder`     | Date pickers, sliders, any widget not in the five built-ins.             |

Every topic's first runnable code block includes a `library()` preamble,
so the LLM's paste-in output works without guessing package loads.

## 5. Inspecting what the LLM will see

You can read any topic directly, without booting a chat session, which
is useful when writing integration tests or checking content before
release:

```{r inspect}
# Head of the primer
cat(substr(ptr_llm_primer(), 1, 400), "\n...")
```

```{r inspect-topic}
# A full topic payload
cat(ptr_llm_topic("level1_ptr_app"))
```

## 6. Provider swap and model choice

The tool definition is provider-neutral — the same registration line
works with every chat constructor ellmer exports.

```{r providers, eval = FALSE}
# Anthropic
chat <- chat_anthropic(system_prompt = ptr_llm_primer(),
                       model = "claude-sonnet-4-6")
ptr_llm_register(chat)

# OpenAI
chat <- chat_openai(system_prompt = ptr_llm_primer(),
                    model = "gpt-4o-mini")
ptr_llm_register(chat)

# Google
chat <- chat_google_gemini(system_prompt = ptr_llm_primer())
ptr_llm_register(chat)

# Local (ollama)
chat <- chat_ollama(system_prompt = ptr_llm_primer(),
                    model = "llama3.1")
ptr_llm_register(chat)
```

The primer is small; any model that supports tool calling will work.
For fast iteration on the registration itself, a small and cheap model
(Haiku, gpt-4o-mini) is a reasonable default.

## 7. Manual registration — what the helper does

`ptr_llm_register()` is a thin convenience wrapper. If you need to
customise the tool (different name, tighter description, extra
arguments), do it by hand:

```{r manual, eval = FALSE}
library(ellmer)

ggpaintr_docs <- tool(
  ptr_llm_topic,
  name        = "ggpaintr_docs",
  description = paste(
    "Fetch a runnable ggpaintr example for one integration topic.",
    "Call before writing R code for interactive ggplot tasks."
  ),
  arguments = list(
    name = type_enum(
      ptr_llm_topics(),
      "Which topic to fetch."
    )
  )
)

chat <- chat_anthropic(system_prompt = ptr_llm_primer())
chat$register_tool(ggpaintr_docs)
```

The pieces:

- `tool(ptr_llm_topic, name = ..., description = ..., arguments = ...)` —
  ellmer's constructor, wrapping the exported R function.
- `type_enum(ptr_llm_topics(), ...)` — constrains the LLM's choice of
  topic to the exact names the package ships, preventing hallucinated
  topic names.
- `chat$register_tool(...)` — attaches the tool to the chat session.

If you use this form, refresh the enum by re-running
`ptr_llm_topics()` whenever the package updates. The helper snapshots
the list once at registration time; any new topics will not be visible
to the LLM until you re-register.

## 8. Testing without spending tokens

The ellmer `tool()` object wraps the underlying R function — you can
still call the function directly for tests or sanity checks:

```{r test-direct}
# Pull a topic without a chat session.
payload <- ptr_llm_topic("level2_embed")
nchar(payload)
```

`ptr_llm_topics()` returns the canonical list; if you are adding a
topic in a downstream repo, assert the name is present before sending
the registration out to users.

## 9. When to re-register

- **After upgrading ggpaintr** — new topics will not be reachable until
  `ptr_llm_register()` runs again (the enum list is snapshotted).
- **Switching chat providers** — registration is per-`Chat` object, so
  every new session needs its own registration.
- **Changing the tool name** — rebuild the chat from scratch; `Chat`
  objects do not expose a public unregister API in current ellmer.

## 10. Current behavior boundary

- The primer and topic files are bundled under `inst/llm/`. They are
  hand-curated from `README.Rmd`, `vignette("ggpaintr-extensibility")`,
  and `vignette("ggpaintr-placeholder-registry")` — but regenerating
  them from the source vignettes is not yet automated. If you edit a
  vignette chunk, update the corresponding topic file manually.
- `ptr_llm_register()` requires
  [ellmer](https://ellmer.tidyverse.org/) in `Suggests:` — it is not
  loaded until you call the helper. If ellmer is not installed,
  `ptr_llm_register()` errors with a clear message.
- The helper sets only the tool. The system prompt is your
  responsibility — pass `ptr_llm_primer()` to the `chat_*()`
  constructor explicitly.
