---
title: "ggpaintr Showcase Demo"
author: "Will Ju"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## About this demo

`ggpaintr` turns a ggplot-like formula string into a Shiny app.
Placeholder tokens inside the formula (`var`, `text`, `num`, `expr`,
`upload`) become Shiny widgets automatically.

Run chunks one at a time. Each `ptr_app*()` call launches an interactive
app; close the window before launching the next one.

```{r packages}
library(ggpaintr)
library(ggplot2)
library(shiny)
library(shinyWidgets)
library(rlang)
# Install once if missing: install.packages("colourpicker")
library(colourpicker)
library(palmerpenguins)
```

------------------------------------------------------------------------

# Part 0 - Shiny App with ggplot2

## A simple Shiny App for Iris

A `selectInput` for `x`, `y`, and `color`, respectively.

```{r simple-iris-1}
library(shiny)
library(ggplot2)

ui <- fluidPage(
  titlePanel("Iris scatter"),
  sidebarLayout(
    sidebarPanel(
      selectInput("x", "X variable",
                  choices = names(iris), selected = "Sepal.Length"),
      selectInput("y", "Y variable",
                  choices = names(iris), selected = "Sepal.Width"),
      selectInput("color", "Color variable",
                  choices = names(iris), selected = "Species")
    ),
    mainPanel(plotOutput("plot"))
  )
)

server <- function(input, output, session) {
  output$plot <- renderPlot({
    ggplot(iris, aes(x = .data[[input$x]],
                     y = .data[[input$y]],
                     color = .data[[input$color]])) +
      geom_point()
  })
}

shinyApp(ui, server)
```

Add a "draw" button? A `textInput` for the plot title?

```{r simple-iris-2}
library(shiny)
library(ggplot2)

ui <- fluidPage(
  titlePanel("Iris scatter"),
  sidebarLayout(
    sidebarPanel(
      selectInput("x", "X variable",
                  choices = names(iris)[1:4], selected = "Sepal.Length"),
      selectInput("y", "Y variable",
                  choices = names(iris)[1:4], selected = "Sepal.Width"),
      selectInput("color", "Color variable",
                  choices = names(iris), selected = "Species"),
      # for title
      textInput("title", "Plot title", value = "Iris scatter"),
      
      # draw button
      actionButton("draw", "Draw", class = "btn-primary")
    ),
    mainPanel(plotOutput("plot"))
  )
)

server <- function(input, output, session) {
  # define reactive on button
  plot_data <- eventReactive(input$draw, {
    list(x = input$x, y = input$y, color = input$color, title = input$title)
  }, ignoreNULL = FALSE)
  
  output$plot <- renderPlot({
    p <- plot_data()
    ggplot(iris, aes(x = .data[[p$x]],
                     y = .data[[p$y]],
                     color = .data[[p$color]])) +
      geom_point() +
      labs(title = p$title) # add labs
  })
}

shinyApp(ui, server)
```

Allow `facet_wrap` in the plot? Add `geom_smooth`?

```{r simple-iris-3}
library(shiny)
library(ggplot2)

ui <- fluidPage(
  titlePanel("Iris scatter"),
  sidebarLayout(
    sidebarPanel(
      selectInput("x", "X variable",
                  choices = names(iris)[1:4], selected = "Sepal.Length"),
      selectInput("y", "Y variable",
                  choices = names(iris)[1:4], selected = "Sepal.Width"),
      selectInput("color", "Color variable",
                  choices = names(iris), selected = "Species"),
      # control for facet
      selectInput("facet", "Facet by",
                  choices = c("None" = "", names(iris)), selected = ""),
      # control for geom_smooth
      selectInput("smooth_method", "Smoothing method",
                  choices = c("None" = "", "lm", "loess", "gam", "glm"),
                  selected = ""),
      textInput("title", "Plot title", value = "Iris scatter"),
      actionButton("draw", "Draw", class = "btn-primary")
    ),
    mainPanel(plotOutput("plot"))
  )
)

server <- function(input, output, session) {
  plot_data <- eventReactive(input$draw, {
    list(
      x = input$x, y = input$y, color = input$color,
      facet = input$facet, smooth_method = input$smooth_method,
      title = input$title
    )
  }, ignoreNULL = FALSE)
  
  output$plot <- renderPlot({
    p <- plot_data()
    g <- ggplot(iris, aes(x = .data[[p$x]],
                          y = .data[[p$y]],
                          color = .data[[p$color]])) +
      geom_point() +
      labs(title = p$title)
    
    if (nzchar(p$smooth_method)) { # meaningful method input
      g <- g + geom_smooth(method = p$smooth_method)
    }
    if (nzchar(p$facet)) { # meaningful facet input
      g <- g + facet_wrap(vars(.data[[p$facet]]))
    }
    g
  })
}

shinyApp(ui, server)
```

Add more controls? Handle error message? See the code behind? At some point, we will have to maintain `ui.R` and `server.R` as two separate files.

## With `ggpaintr`

```{r simple-iris-ggpaintr}
ptr_app("
 ggplot(data = iris, aes(x = var, y = var, color = var)) +
   geom_point() + 
   geom_smooth(method = text) + 
   facet_wrap(expr) + 
   labs(title = text)
")
```

Change the design

```{r simple-iris-ggpaintr-2}
library(palmerpenguins)

ptr_app("
 ggplot(data = penguins, aes(x = var, y = var, color = var, fill = var)) +
   geom_point() + 
   geom_boxplot() + 
   geom_smooth(method = text) + 
   facet_wrap(expr) + 
   labs(title = text)
")
```

------------------------------------------------------------------------

# Part 1 — Core Features

## Example 1.1: The simplest possible app

One formula, two `var` pickers, one `text` box. Everything else is free.

```{r basic}
ptr_app("
ggplot(data = iris, aes(x = var, y = var)) +
  geom_point() +
  labs(title = text)
")
```

Try: `x = Sepal.Length`, `y = Sepal.Width`, `title = Iris scatter`, then
click **Update plot**. Watch the generated code pane mirror your
choices.

## Example 1.2: All five placeholders in one app

`var` (column picker), `text` (string), `num` (numeric), `expr` (parsed
R expression), `upload` (file input). Each layer is toggleable.

```{r all-placeholders}
ptr_app("
ggplot(data = upload, aes(x = var, y = var)) +
  geom_point(aes(color = var), size = num, alpha = num) +
  geom_smooth(method = text, se = expr) +
  labs(title = text, x = text, y = text) +
  facet_wrap(expr) +
  theme(legend.position = text)
")
```

Suggested flow:

-   upload `tests/testthat/fixtures/simple_numeric.csv` (blank the
dataset name)
-   `x = x`, `y = y`, `color = group`
-   `size = 2.5`, `alpha = 0.7`
-   `method = lm`, `se = TRUE` (expr, parsed as R)
-   `title = Uploaded scatter`, `legend.position = bottom`
-   `facet_wrap expr = ~ group`

Uncheck individual layers to see them vanish from both plot and code.

## Example 1.3: `expr` as a compatibility escape hatch

`expr` is parsed as an R expression at draw time, so any ggplot2 feature
(scales, coords, stats, annotations, vector-valued args) can be dropped
in without needing a dedicated placeholder.

```{r expr-compat}
ptr_app("
ggplot(data = mtcars, aes(x = var, y = var, color = factor(cyl))) +
  geom_point(size = num) +
  geom_smooth(method = text, formula = expr, se = expr) +
  scale_color_manual(values = expr) +
  coord_cartesian(xlim = expr, ylim = expr) +
  labs(title = text) +
  expr
")
```

Suggested inputs showing the full reach of `expr`:

-   `x = wt`, `y = mpg`, `size = 3`
-   `method = lm`, `formula = y ~ x`, `se = FALSE`
-   `scale_color_manual values = c('4' = 'steelblue', '6' = 'firebrick', '8' = 'goldenrod')`
-   `coord_cartesian xlim = c(1, 6)`, `ylim = c(10, 35)`
-   `title = mtcars with expr features`
-   Trailing
`expr = theme_minimal(base_size = 14) + theme(legend.position = 'top')`

Any ggplot component you can write in R code, you can pipe through
`expr`.

------------------------------------------------------------------------

# Part 2 — Three Levels of Extensibility

## Level 1: One-liner wrapper with UI text overrides

`ptr_app()` / `ptr_app_bslib()` — the shipped wrappers. Customize
visible copy with `ui_text` without writing any Shiny code.

```{r extensibility-level-1}
ptr_app_bslib(
  "
ggplot(data = iris, aes(x = var, y = var)) +
  geom_point(aes(color = var), size = num) +
  labs(title = text)
",
ui_text = list(
  shell = list(
    title       = list(label = "Iris Explorer"),
    draw_button = list(label = "Render")
  ),
  params = list(
    x     = list(var  = list(label = "X-axis column")),
    y     = list(var  = list(label = "Y-axis column")),
    color = list(var  = list(label = "Colour by")),
    title = list(text = list(label = "Chart title"))
  ),
  layers = list(
    geom_point = list(num = list(size = list(
      label = "Point size",
      help  = "Typical values: 1 – 5"
    )))
  )
)
)
```

## Level 2: Embed into your own Shiny layout

`ptr_server_state()` plus the `ptr_register_*()` bind helpers let you
drop `ggpaintr` into a larger Shiny UI with custom ids. You own the
layout; the helpers wire up the runtime.

```{r extensibility-level-2}
custom_ids <- ptr_build_ids(
  control_panel = "builder_controls",
  draw_button   = "render_plot",
  plot_output   = "main_plot",
  error_output  = "main_error",
  code_output   = "main_code"
)

ui <- fluidPage(
  titlePanel("Embedded ggpaintr"),
  sidebarLayout(
    sidebarPanel(
      ptr_input_ui(ids = custom_ids),
      tags$hr(),
      h4("Session status"),
      textOutput("runtimeStatus")
    ),
    mainPanel(
      ptr_output_ui(ids = custom_ids)
    )
  )
)

server <- function(input, output, session) {
  ptr_state <- ptr_server_state(
    "ggplot(data = iris, aes(x = var, y = var)) +
      geom_point(aes(color = var), size = num) +
      labs(title = text)",
    ids = custom_ids
  )
  
  ptr_setup_controls(input, output, ptr_state, ids = custom_ids)
  ptr_register_draw(input, ptr_state, ids = custom_ids)
  ptr_register_plot(output, ptr_state, ids = custom_ids)
  ptr_register_error(output, ptr_state, ids = custom_ids)
  ptr_register_code(output, ptr_state, ids = custom_ids)
  
  # Your own reactive reading ptr_state$runtime()
  output$runtimeStatus <- renderText({
    r <- ptr_state$runtime()
    if (is.null(r))          return("Waiting for first draw...")
    if (isTRUE(r$ok))        return(paste("OK —", nchar(r$code_text), "chars of code"))
    r$message
  })
}

shinyApp(ui, server)
```

## Level 3: Own the render path with pure extractors + `ptr_gg_extra()`

Swap in your own `renderPlot()`. Pull the built plot with
`ptr_extract_plot()`, append any ggplot components, and wrap those
components with `ptr_gg_extra()` so they round-trip into the
generated-code pane too.

```{r extensibility-level-3}
ui <- fluidPage(
  titlePanel("Custom render path"),
  sidebarLayout(
    sidebarPanel(ptr_input_ui()),
    mainPanel(
      plotOutput("outputPlot", height = "500px"),
      uiOutput("outputError"),
      verbatimTextOutput("outputCode")
    )
  )
)

server <- function(input, output, session) {
  ptr_state <- ptr_server_state(
    "ggplot(data = iris, aes(x = var, y = var, color = Species)) +
      geom_point(size = num) +
      labs(title = text)"
  )
  
  ptr_setup_controls(input, output, ptr_state)
  ptr_register_draw(input, ptr_state)
  ptr_register_error(output, ptr_state)
  ptr_register_code(output, ptr_state)
  
  output$outputPlot <- renderPlot({
    plot_obj <- ptr_extract_plot(ptr_state$runtime())
    if (is.null(plot_obj)) {
      plot.new(); return(invisible(NULL))
    }
    
    # ptr_gg_extra() captures these additions so ptr_register_code()
    # appends them to outputCode — keeps the rendered plot and the
    # code pane in sync.
    plot_obj + ptr_gg_extra(
      ptr_state,
      theme_minimal(base_size = 16),
      scale_color_brewer(palette = "Dark2"),
      theme(legend.position = "bottom")
    )
  })
}

shinyApp(ui, server)
```

## One Embedding Demonstration

[`ggiraph`](https://davidgohel.github.io/ggiraph/) 

```{r}
library(ggplot2)
library(ggiraph)
data <- mtcars
data$carname <- row.names(data)

gg_point = ggplot(data = data) +
  geom_point_interactive(aes(x = wt, y = qsec, color = disp,
                             tooltip = carname, data_id = carname)) + 
  theme_minimal()

girafe(ggobj = gg_point)
```


```{r}
library(ggiraph)

ui <- fluidPage(
  titlePanel("Custom render path"),
  sidebarLayout(
    sidebarPanel(ptr_input_ui()),
    mainPanel(
      ggiraph::girafeOutput("outputPlot"), # ggiraph
      uiOutput("outputError"),
      verbatimTextOutput("outputCode")
    )
  )
)

server <- function(input, output, session) {
  ptr_state <- ptr_server_state(
    "ggplot(data = penguins, aes(x = var, y = var, color = var)) +
      geom_point_interactive(
        aes( data_id = var, tooltip = var), size = 3,
        hover_nearest = TRUE) +
      labs(title = text)"
  )
  
  ptr_setup_controls(input, output, ptr_state)
  ptr_register_draw(input, ptr_state)
  ptr_register_error(output, ptr_state)
  ptr_register_code(output, ptr_state)
  
  output$outputPlot <- renderGirafe({
    plot_obj <- ptr_extract_plot(ptr_state$runtime())
    if (is.null(plot_obj)) {
      plot.new(); return(invisible(NULL))
    }
    
    x <- girafe(
      code = print(plot_obj),
      width_svg = 6, height_svg = 5,
      options = list(
        opts_hover(css = "fill:#FF3333;stroke:black;cursor:pointer;", 
                   reactive = TRUE),
        opts_selection(
          type = "multiple", css = "fill:#FF3333;stroke:black;")
      ))
    x
    
  })
}

shinyApp(ui, server)

```

`plotly`

```{r}
library(plotly)

ui <- fluidPage(
  titlePanel("Custom render path"),
  sidebarLayout(
    sidebarPanel(ptr_input_ui()),
    mainPanel(
      plotlyOutput("outputPlot"), # for plotly
      uiOutput("outputError"),
      verbatimTextOutput("outputCode")
    )
  )
)

server <- function(input, output, session) {
  # ptr_state <- ptr_server_state(
  #   "ggplot(data = penguins, aes(x = var, y = var, color = var)) +
  #     geom_point_interactive(
  #       aes( data_id = var, tooltip = var), size = 3,
  #       hover_nearest = TRUE) +
  #     labs(title = text)"
  # )
  ptr_state <- ptr_server_state(
    "ggplot(data = penguins, aes(x = var, y = var, color = var)) +
      geom_point() + 
      labs(title = text)"
  )
  
  ptr_setup_controls(input, output, ptr_state)
  ptr_register_draw(input, ptr_state)
  ptr_register_error(output, ptr_state)
  ptr_register_code(output, ptr_state)
  
  output$outputPlot <- renderPlotly({ # render
    plot_obj <- ptr_extract_plot(ptr_state$runtime())
    if (is.null(plot_obj)) {
      plot.new(); return(invisible(NULL))
    }
    
    ggplotly(plot_obj)

    # x <- girafe(
    #   code = print(plot_obj),
    #   width_svg = 6, height_svg = 5,
    #   options = list(
    #     opts_hover(css = "fill:#FF3333;stroke:black;cursor:pointer;",
    #                reactive = TRUE),
    #     opts_selection(
    #       type = "multiple", css = "fill:#FF3333;stroke:black;")
    #   ))
    # x

  })
}

shinyApp(ui, server)

```


------------------------------------------------------------------------

# Part 3 — Custom Placeholders

## Example 3.1: A `color` placeholder with `colourpicker::colourInput`

Replace the default `text`-box approach for colours with a real colour
picker. `copy_defaults` ships labels; `resolve_expr()` quotes the chosen
colour as a string literal in generated code.

```{r color-placeholder}
# install.packages("colourpicker")  # if needed

color_placeholder <- ptr_define_placeholder(
  keyword = "color",
  
  # build_ui: one widget per placeholder occurrence in the formula.
  build_ui = function(id, copy, meta, context) {
    default <- if (identical(meta$param, "fill")) "#4E79A7" else "#E15759"
    colourpicker::colourInput(
      id,
      label = copy$label,
      value = default,
      # showColour = "both",
      # palette = "limited"
    )
  },
  
  # resolve_expr: chosen colour → a quoted string literal in the code.
  resolve_expr = function(value, meta, context) {
    if (is.null(value) || identical(value, "")) {
      return(ptr_missing_expr())
    }
    rlang::expr(!!as.character(value))
  },
  
  copy_defaults = list(label = "Pick {param} colour")
)

color_placeholders <- ptr_merge_placeholders(
  list(color = color_placeholder)
)

ptr_app(
  "
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(color = color, fill = color, size = num, shape = 21) +
  geom_smooth(method = 'lm', color = color, se = FALSE) +
  labs(title = text)
",
placeholders = color_placeholders
)
```

Try picking three different colours and watch the generated code update
with the hex strings.

## Example 3.2: A custom `date` placeholder with param-aware copy

A second custom placeholder: a `dateInput` whose chosen date is spliced
as `as.Date("YYYY-MM-DD")` into the generated code. Shows how
`ptr_merge_ui_text()` layers copy rules for a custom keyword.

```{r date-placeholder}
sales <- data.frame(
  day     = as.Date("2024-01-01") + 0:9,
  value   = c(10, 13, 12, 16, 18, 17, 19, 22, 20, 24),
  segment = rep(c("A", "B"), each = 5)
)

date_placeholder <- ptr_define_placeholder(
  keyword = "date",
  build_ui = function(id, copy, meta, context) {
    shiny::dateInput(id, label = copy$label, value = "2024-01-01")
  },
  resolve_expr = function(value, meta, context) {
    if (is.null(value) || identical(as.character(value), "")) {
      return(ptr_missing_expr())
    }
    rlang::expr(as.Date(!!as.character(value)))
  },
  copy_defaults = list(label = "Choose a date for {param}")
)

date_placeholders <- ptr_merge_placeholders(list(date = date_placeholder))

date_ui_text <- ptr_merge_ui_text(
  list(
    defaults = list(date = list(label = "Pick any date")),
    params   = list(xintercept = list(date = list(label = "Reference date"))),
    layers   = list(
      geom_vline = list(date = list(xintercept = list(
        help = "Cut-off date for the vertical guide."
      )))
    )
  ),
  placeholders = date_placeholders
)

ptr_app(
  "
ggplot(data = sales, aes(x = day, y = value, color = segment)) +
  geom_line() +
  geom_point(size = num) +
  geom_vline(xintercept = date, linetype = 'dashed') +
  labs(title = text)
",
placeholders = date_placeholders,
ui_text      = date_ui_text
)
```

Change the reference date; the vertical line and the `as.Date(...)`
string in the code pane both update on draw.

------------------------------------------------------------------------

